---
title: "Analysis of bulk RNA seq in chronic and acute murine colitis"
author: "Jennifer Fransson"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: yes
  bookdown::html_document2:
    code_folding: hide
    toc: yes
    toc_float:
      collapsed: no
  html_document:
    df_print: paged
    toc: yes
params:
  dataset: 1
editor_options:
  chunk_output_type: inline
---

```{r Setup, echo = FALSE}
knitr::opts_chunk$set(results    = "hold",
                      message    = FALSE,
                      warning    = FALSE)

if(!knitr::is_html_output()){
  my_plot_hook <- function(x, options)
   paste("\n", knitr::hook_plot_tex(x, options), "\n")
  knitr::knit_hooks$set(plot = my_plot_hook)
}
```



# Introduction

This script describes the analyses performed to compare disease process of T cell transfer colitis and Il10^-/-^ colitis for the study "The Spatiotemporal Transcriptomic Landscape in Murine Chronic Colitis". The datasets used here have been generated in the scripts Il10KO_bulk.html, TcellTransfer_bulk_Rag2.html, and TcellTransfer_bulk_SCID.html. In addition, a dataset previously published by Czarnewski et al (Nat Comm, 2019) was downloaded from [here](https://github.com/czarnewski/uc_classification).

# Load libraries and other scripts

In addition to the packages loaded, an extra script (bulk_functions.R) is sourced prior to analysis. This script contains functions and parameters for visualization that are used for this analysis as well as the analysis of T cells transfer bulk RNAseq.

```{r}

library(ggplot2)
library(cowplot)
library(grid)
library(mgcv)
library(rafalib)
library(edgeR)
library(reshape2)
library(enrichR)
library(ggvenn)
library(data.table)
library(ggpubr)
library(progeny)
library(pheatmap)
library(igraph)
library(ggraph)


source("../code/Bulk/bulk_functions.R")
datadir = "../data/Bulk/"


gtheme = theme(legend.key.size = unit(10,"pt"), legend.title = element_text(size = 8),
        legend.text = element_text(size = 8), legend.margin = margin(0,0,0,0),
        legend.spacing = unit(2,"pt"))


```

# Read datasets

```{r}

# Identify genes originally present in all datasets

presentgenes= read.csv2("../data/BulkCommonGenes.csv")$gene


```


```{r}


IL10_DE = read.csv2(paste0(datadir,"Il10KO/DE_modules.csv"), row.names = 1)
Rag2_DE = read.csv2(paste0(datadir,"TCellTransfer_Rag2/DE_modules.csv"), row.names = 1)
SCID_DE = read.csv2(paste0(datadir,"TCellTransfer_SCID/DE_modules.csv"), row.names = 1)
IL10_DE$GeneName = rownames(IL10_DE)
Rag2_DE$GeneName = rownames(Rag2_DE)
SCID_DE$GeneName = rownames(SCID_DE)

dss_modules = read.csv("https://github.com/czarnewski/uc_classification/blob/master/processed_files/gene_modules.csv?raw=TRUE")
colnames(dss_modules) = c("GeneName","module")

dss_de = read.csv("https://github.com/czarnewski/uc_classification/blob/master/processed_files/DEgenes.csv?raw=TRUE")
colnames(dss_de)[1] = "GeneName"

DSS_DE = merge(dss_modules, dss_de, by = "GeneName")


de_res_modules = lapply(list(IL10_DE, SCID_DE, Rag2_DE, DSS_DE), function(x){
  x = x[x$GeneName %in% presentgenes,]
  x = x[!is.na(x$module),]
  x
})
names(de_res_modules) = c("il10","scid","rag2","dss")


```


```{r, fig.height = 3, fig.width = 3}

full_v = ggvenn(lapply(de_res_modules, function(x){
  x$GeneName}),
  c("il10","scid","rag2","dss"), show_percentage = FALSE)

full_v


```

```{r}

IL10_DE_DF = read.csv2(paste0(datadir,"Il10KO/Human_DE.csv"), row.names = 1)
Rag2_DE_DF = read.csv2(paste0(datadir,"TCellTransfer_Rag2/Human_DE.csv"), row.names = 1)
SCID_DE_DF = read.csv2(paste0(datadir,"TCellTransfer_SCID/Human_DE.csv"), row.names = 1)

mousemodules = DSS_DE

homologs <- fread('http://www.informatics.jax.org/downloads/reports/HMD_HumanPhenotype.rpt')
homologs <- data.frame("mouse"=homologs$V3, "human"=homologs$V1)

mouse_human = homologs[match(rownames(mousemodules), homologs$mouse),]
mouse_human = mouse_human[!is.na(mouse_human$mouse),]

mouse_human$cluster = mousemodules$module[match(mouse_human$mouse, rownames(mousemodules))]

DSS_DE_DF = getHumanDEs(mouse_human, homologs)

DSS_DE_DF$mousemodule = mouse_human[match(DSS_DE_DF$mouse, mouse_human$mouse),"module"]

```


```{r, fig.height = 3, fig.width = 3}

combined_DE_DF = unique(do.call("rbind",
        lapply(list(IL10_DE_DF, Rag2_DE_DF, SCID_DE_DF, DSS_DE_DF), function(x){
  x[,c("mouse", "human", "Taman_UC_logFC", "Lloyd_UC_logFC",
       "Taman_UC_reg", "Lloyd_UC_reg")]
})))

combined_DE_DF$UC = (!is.na(combined_DE_DF$Lloyd_UC_reg) &
                       !is.na(combined_DE_DF$Taman_UC_reg))

uc_v = ggvenn(lapply(de_res_modules, function(x){
  x[x$GeneName %in% combined_DE_DF$mouse[combined_DE_DF$UC],"GeneName"]}),
  c("il10","scid","rag2","dss"), show_percentage = FALSE)

uc_v

```


```{r, fig.width = 6, fig.height = 2}

n_v = full_v$layers[[4]]$data$n
names(n_v) = full_v$layers[[4]]$data$name

full_matrix = matrix(
  c(0, n_v[c("B","AB","A",
             "C","BC","ABC","AC",
             "CD","BCD","ABCD","ACD",
             "D","BD","ABD","AD")]),
  ncol = 4, nrow = 4)

colnames(full_matrix) = gsub("C","Rag2",gsub("D","DSS",c("0","C","C&D","D")))
rownames(full_matrix) = gsub("A","Il10",gsub("B","SCID",c("0","B","A&B","A")))

full_melt = merge(reshape2::melt(full_matrix, value.name = "full_n"),
                  reshape2::melt(full_matrix/sum(full_matrix), 
                                 value.name = "full_prop"))

n_v = uc_v$layers[[4]]$data$n
names(n_v) = uc_v$layers[[4]]$data$name

uc_matrix = matrix(
  c(0, n_v[c("B","AB","A",
             "C","BC","ABC","AC",
             "CD","BCD","ABCD","ACD",
             "D","BD","ABD","AD")]),
  ncol = 4, nrow = 4)

colnames(uc_matrix) = gsub("C","Rag2",gsub("D","DSS",c("0","C","C&D","D")))
rownames(uc_matrix) = gsub("A","Il10",gsub("B","SCID",c("0","B","A&B","A")))

uc_melt = merge(reshape2::melt(uc_matrix, value.name = "uc_n"),
reshape2::melt(uc_matrix/sum(uc_matrix), value.name = "uc_prop"))

combined_melt = merge(full_melt, uc_melt)

combined_melt$Var1 = factor(combined_melt$Var1,
                            levels = rev(rownames(uc_matrix)))

ggplot(combined_melt, aes(x = Var2, y = Var1)) +
  geom_point(aes(size = full_n, fill = uc_n/full_n*100), shape = 21) + 
  scale_size_continuous(range = c(0,10)) + 
  theme(axis.text = element_text(size = 8),
        axis.title = element_blank(),
        legend.direction = "horizontal",
        plot.title = element_text(size = 10)) +
  scale_x_discrete(position = "top") + gtheme + 
  scale_fill_gradient(low = "white",high = "#8888FF", limits = c(0,NA),
                      breaks = c(0,5,10,15)) + 
  labs(title = "Proportion UC DE genes out of TCT genes",
       size = "# DE mouse genes",
       fill = "% genes with\nDE homolog in UC")


```

```{r}

allsig = do.call("rbind",lapply(names(de_res_modules), function(x){
  y = de_res_modules[[x]][,Reduce(intersect,lapply(de_res_modules, colnames))]
  y$model = x
  y
}))

allsig$inUC = factor(allsig$GeneName %in% combined_DE_DF$mouse[combined_DE_DF$UC],
                     levels = c("TRUE","FALSE"))

percentages = data.frame(true = table(allsig$model, allsig$inUC)[,1],
           false = table(allsig$model, allsig$inUC)[,2])
percentages$sum = percentages$true + percentages$false
percentages$perc = percentages$true/percentages$sum*100
percentages$model = factor(rownames(percentages), levels = rev(rownames(percentages)))

ggplot(allsig, aes(x = factor(model, levels = levels(percentages$model)))) + 
  geom_bar(aes(fill = inUC)) + gtheme + 
  scale_fill_manual(values = c("#CCCCCC", "#AAAAAA")) +
  theme(axis.text = element_text(size = 8),
        axis.ticks.x = element_line(color = "black"),
        panel.background = element_blank(),
        panel.grid = element_blank(),
        panel.grid.major.y = element_line(color = "black")) + 
  labs(x = "Model") +
  geom_text(data = percentages, 
            aes(y = sum-true/2,
                label = paste0("DE in UC: ",true," (",signif(perc,2),"%)")),
            hjust = 0.5, vjust = 0.5) + 
  scale_y_continuous(breaks = c(0,2000,4000,6000), limits = c(0,6500), name = "count")


```

```{r, fig.height = 2, fig.width = 6}

plot_grid(plotlist = lapply(unique(allsig$model), function(model){
  ggplot(allsig[allsig$model == model,],
         aes(x = inUC, y = -log10(FDR))) + 
    geom_jitter(shape = 16, alpha = 0.5, stroke = NA, size = 1) + 
    geom_boxplot(outlier.size = -1,
                 color = "red",
                 alpha = 0.2) + 
    labs(title = model, x = "DE in UC", y = "-log10(FDR)") + 
    gtheme + 
    stat_compare_means(method = "wilcox", aes(y = FDR)) +
    theme(axis.text = element_text(size = 8),
          axis.ticks.y = element_line(color = "black"),
          legend.title = element_blank(),
          plot.title = element_text(size = 8),
          plot.margin = margin(2,2,2,2,"pt"))}),
  ncol = 4, align = "h")

```


```{r}


uc_conserved = lapply(de_res_modules, function(x){
  x[x$GeneName %in% combined_DE_DF$mouse[combined_DE_DF$UC],"GeneName"]})

conserved = intersect(uc_conserved[[1]], 
          intersect(uc_conserved[[2]], 
                    intersect(uc_conserved[[3]],
                              uc_conserved[[4]])))

write.csv2(combined_DE_DF[combined_DE_DF$mouse %in% conserved,], 
           paste0(datadir,"uc_conserved_table.csv"), 
           row.names = FALSE)

conserved_go = enrichr(conserved, databases = "GO_Biological_Process_2021")


top10_uc = conserved_go$GO_Biological_Process_2021[1:10,]

top10_uc$log10p = -log10(top10_uc$Adjusted.P.value)

top10_uc$Name = gsub(" \\(.*","",top10_uc$Term)
top10_uc$GOID = gsub(".*\\(","",gsub("\\)","",top10_uc$Term))
top10_uc$GeneList = sapply(strsplit(top10_uc$Genes,";"), 
                           paste, collapse = ", ")

uc_go_edges = as.data.frame(do.call("rbind",lapply(1:10, function(i){
  do.call("rbind",
          lapply(strsplit(top10_uc$Genes[i], ";")[[1]], function(g){
            c(from = top10_uc$Name[i], to = g)
          }))
})))

genes = lapply(1:10, function(i){
          strsplit(top10_uc$Genes[i], ";")[[1]]
})

uc_gene_nodes = data.frame(Name = unique(uc_go_edges$to),
           GOID = unique(uc_go_edges$to), 
           Overlap = NA,
           log10p = NA)

uc_gene_nodes$Type = "Gene"
top10_uc$Type = "GO_Term"

uc_go_nodes = rbind(top10_uc[,c("Name", "GOID", "Overlap", "log10p","Type")],
                    uc_gene_nodes)


write.csv(uc_go_edges, 
          paste0(datadir, "uc_conserved_go_edges.csv"),
          quote = FALSE, row.names = FALSE)
write.csv(uc_go_nodes, 
          paste0(datadir, "uc_conserved_go_nodes.csv"),
          quote = FALSE, row.names = FALSE)

uc_go_edges$from = gsub("ted sign", "ted\nsign",
                     gsub("to ","to\n",
                          gsub("of ","of\n",
                               gsub(" \\(.*","",uc_go_edges$from))))

uc_go_nodes$Name = gsub("ted sign", "ted\nsign",
                     gsub("to ","to\n",
                          gsub("of ","of\n",
                               gsub(" \\(.*","",uc_go_nodes$Name ))))



```

Top 10 enriched GO Biological Process terms in `r length(conserved)` conserved genes:

`r knitr::kable(top10_uc[,c("Term","Adjusted.P.value","GeneList")])`


```{r, fig.height = 6}

ig <- graph_from_data_frame(d=uc_go_edges, vertices=uc_go_nodes, directed = FALSE)

tg <- tidygraph::as_tbl_graph(ig) %>% 
  tidygraph::activate(nodes) %>% 
  dplyr::mutate(label=name)


```

```{r, fig.height = 6}


tg %>%
   ggraph(layout = "fr") +
   geom_edge_arc(colour= "gray50",
                 lineend = "round",
                 strength = 0,
                 alpha = .3) +
   geom_node_point(aes(size = Type),
                   color = "grey") +
   geom_node_text(aes(label = name), 
                  repel = FALSE,  
                  colour="gray10") +
  theme_graph(background = "white") +
  scale_edge_width(range = c(0.2, 2.5)) + 
    theme_graph(base_family = 'Helvetica', base_size = 8) +
  gtheme

```

```{r}

sessionInfo()

```

