---
title: "Il10KO single cell analysis (Epithelial)"
editor_options:
  chunk_output_type: inline
date: "27 February, 2024"
output:
  bookdown::html_document2:
    code_folding: hide
    toc: yes
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
params:
  dataset: 1
---

```{r Setup, echo = FALSE}
knitr::opts_chunk$set(fig.width  = 7,
                      results    = "hold",
                      message    = FALSE,
                      cache.lazy = FALSE,
                      warning    = FALSE)


```




```{r setup}
 
library(Seurat)
library(ggplot2)
library(cowplot)
library(clustree)
library(dplyr)
library(reshape2)

source("../code/SingleCell/color_list.R")

```


```{r}
 
datadir = "../data/SingleCell/Il10KO/R_objects/"

owncache = "../data/SingleCell/Il10KO/cache_epithelial/"

violintheme = theme(axis.text = element_text(size = 8),
        plot.title = element_text(size = 8),
        legend.key.size = unit(8, "pt"),
        axis.title = element_blank())
dimtheme = theme(legend.text = element_text(size = 8),
        plot.title = element_text(size = 8),
        legend.key.size = unit(8, "pt"))

qc_feats <- c("nFeature_RNA", "nCount_RNA", "percent_mito", "percent_ribo")

```

```{r, include = FALSE}

if(file.exists(owncache)){
  message(paste0("Intermediate data files will be loaded from and saved to ",owncache))
}else{
  message("Intermediate data files will not be saved or loaded")
}

```

# Initial data 

This report describes the analysis performed on scRNA-seq data from epithelial cells isolated from Il10-/- colitis mouse colons. The data has first been processed together with immune and stroma cells, see "Il10KO_SingleCell_DataPreparation.html". The data object produced from that analysis is saved as the object data/SingleCell/Il10KO/R_objects/epithelial_data.rds.

Due to a large heterogeneity of cell quality in multiple samples, a filtering step is added to only include cells with > 1500 unique genes.

## Pre-filtering QC



```{r}

alldata = readRDS(paste0(datadir,"epithelial_data.rds"))

```

```{r, fig.height=4, fig.width=6}

VlnPlot(alldata, group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) &
  violintheme

```

## Filtering

```{r}

if(!file.exists(paste0(owncache,"first_dr.RData"))){
  
  alldata <- subset(alldata, cells = WhichCells(alldata, expression = nFeature_RNA>1500))
  
  suppressWarnings(suppressMessages(alldata <- FindVariableFeatures(alldata, selection.method = "vst", 
    nfeatures = 2000, verbose = FALSE, assay = "RNA")))

  
  alldata <- ScaleData(alldata, vars.to.regress = c("percent_mito", "nFeature_RNA"), 
    assay = "RNA")

  alldata <- RunPCA(alldata, npcs = 50, verbose = F)
  
  alldata <- RunUMAP(alldata, reduction = "pca", dims = 1:30, n.components = 2, n.neighbors = 30, 
    n.epochs = 200, min.dist = 0.3, learning.rate = 1, spread = 1)
  if(file.exists(owncache)){
    save(alldata, file = paste0(owncache,"first_dr.RData"))
  }
}else{
  load(paste0(owncache,"first_dr.RData"))
}

```

# Quality control

Number of features and counts and percentage of mitochondrial and ribosomal RNA are plotted below.

```{r qc, fig.height=4, fig.width=6}

VlnPlot(alldata, group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) &
  violintheme

```

# Dimensional reduction - pre-integration

The data is first processed in the same way as the full dataset, with PCA on the 2000 most highly variable genes and subsequent UMAP for visualization.

## Highly variable features

```{r topgenes, fig.height = 3, fig.width = 5}

top20 <- head(VariableFeatures(alldata), 20)
lp = LabelPoints(plot = VariableFeaturePlot(alldata), points = top20, repel = TRUE) +
  theme(axis.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        axis.text = element_text(size = 8))

lp$layers[[2]]$aes_params$size = 3

lp

```

## PCA

```{r, fig.height = 3, fig.width = 6}


plot_grid(ncol = 3, 
          DimPlot(alldata, reduction = "pca", group.by = "orig.ident", 
                  dims = 1:2, shuffle = TRUE) +
            NoLegend() + dimtheme + 
            theme(axis.text = element_text(size = 8),
                  axis.title = element_text(size = 8)),
          DimPlot(alldata, reduction = "pca", group.by = "orig.ident", 
                  dims = 3:4, shuffle = TRUE) + 
            NoLegend() + dimtheme + 
            theme(axis.text = element_text(size = 8),
                  axis.title = element_text(size = 8)),
          DimPlot(alldata, reduction = "pca", group.by = "orig.ident",
                  dims = 5:6, shuffle = TRUE) + 
            NoLegend() + dimtheme + 
            theme(axis.text = element_text(size = 8),
                  axis.title = element_text(size = 8)))

```

```{r, fig.height = 3, fig.width = 6}

VizDimLoadings(alldata, dims = 1:4, reduction = "pca", ncol = 4, balanced = T, nfeatures = 20) &
  theme(axis.text = element_text(size = 8), axis.title = element_text(size = 8))

```

```{r, fig.height = 2, fig.width = 4}

ElbowPlot(alldata, reduction = "pca", ndims = 50) + 
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 8))

```

## UMAP

```{r firstumap, fig.height = 3, fig.width = 6}

plot_grid(ncol = 2, DimPlot(alldata, reduction = "pca", raster = TRUE,
                            group.by = "orig.ident", shuffle = TRUE) + 
            NoLegend() + NoAxes() + dimtheme, 
    DimPlot(alldata, reduction = "umap", group.by = "orig.ident", raster = TRUE,
            shuffle = TRUE) +
      NoLegend() + NoAxes() + dimtheme)

plot_grid(ncol = 2, DimPlot(alldata, reduction = "umap", raster = TRUE,
                            group.by = "orig.ident", shuffle = TRUE) +
            theme(legend.position = "bottom") + NoAxes() + dimtheme, 
          DimPlot(alldata, reduction = "umap", raster = TRUE,
                  group.by = "Genotype", shuffle = TRUE) +
            theme(legend.position = "bottom") + NoAxes() + dimtheme)

plot_grid(ncol = 2, DimPlot(alldata, reduction = "umap", raster = TRUE,
                  group.by = "RPCA_cluster", label = TRUE, shuffle = TRUE) + 
            NoLegend() + NoAxes() + dimtheme)

FeaturePlot(alldata, reduction = "umap", features = c("Ptprc","Epcam"), raster = TRUE,
    order = T, slot = "data", combine = T) & dimtheme & NoAxes() & NoLegend()

```

# Integration

To enable equivalent clustering across samples, CCA integration is run with standard Seurat functions. 

```{r}

if(!file.exists(paste0(owncache,"integration.RData"))){
  if(!file.exists(paste0(owncache,"anchors.RData"))){
    alldata <- SplitObject(alldata, split.by = "orig.ident")
  
    gc()
    
    for(i in 1:length(alldata)){
      alldata[[i]] = NormalizeData(alldata[[i]])
      alldata[[i]] <- FindVariableFeatures(alldata[[i]], selection.method = "vst",
            nfeatures = 2000)
    }
    
    data.anchors <- FindIntegrationAnchors(
      object.list = alldata, dims = 1:30, 
        reduction = "cca")
    rm(alldata)
    gc()
    
    if(file.exists(owncache)){
      save(data.anchors, file = paste0(owncache,"anchors.RData"))
    }
  }else{
    rm(alldata)
    gc()
    load(paste0(owncache,"anchors.RData"))
  }
  
  alldata <- IntegrateData(anchorset = data.anchors, dims = 1:30,
                                    new.assay.name = "CCA")
  
  rm(data.anchors)
  gc()
  
  alldata <- ScaleData(alldata)
  alldata <- RunPCA(alldata, npcs = 30)
  alldata <- RunUMAP(alldata, dims = 1:30, n.neighbors = 50)
  
  if(file.exists(owncache)){
    save(alldata, file = paste0(owncache,"integration.RData"))
  }
}else{
  load(paste0(owncache,"integration.RData"))
}

```

## Dimensional reduction

```{r, fig.height = 2, fig.width=5}

plot_grid(ncol = 2,
  
  DimPlot(alldata, reduction = "pca", group.by = "orig.ident", pt.size = 0.01, shuffle = TRUE) +
    NoAxes() + ggtitle("PCA integrated") + NoLegend() + dimtheme,
  DimPlot(alldata, reduction = "umap", group.by = "orig.ident", pt.size = 0.01, shuffle = TRUE) +
    NoAxes() + ggtitle("UMAP integrated") + NoLegend() + dimtheme
)

alldata@active.assay = "RNA"
FeaturePlot(alldata, reduction = "umap", features = c("Ptprc","Epcam"), 
    order = T, slot = "data", combine = T) & dimtheme & NoAxes()
alldata@active.assay = "CCA"

```


```{r, fig.height = 6, fig.width=4}

FeaturePlot(alldata, reduction = "umap", 
            features = c("percent_mito","percent_ribo","S.Score",
                         "G2M.Score", "nFeature_RNA"), 
    order = T, slot = "data", combine = T, ncol = 2) &
  NoAxes() & NoLegend() & dimtheme

```

# Clustering

The dataset is clustered using SNN and the Louvain algorithm at 4 resolutions. Resolution 1.0 is chosen for further analysis.

```{r}

if(!file.exists(paste0(owncache, "clustered.RData"))){
  alldata@active.assay <- "CCA"
  
  alldata@active.assay

  alldata <- FindNeighbors(alldata, dims = 1:30, k.param = 350, prune.SNN = 1/15)
  names(alldata@graphs)
  
  # Clustering with louvain (algorithm 1)
  for (res in c(0.25, 0.5, 1, 1.5)) {
      alldata <- FindClusters(alldata, graph.name = "CCA_snn",
                              resolution = res, algorithm = 1)
  }
  if(file.exists(owncache)){
    save(alldata, file = paste0(owncache, "clustered.RData"))
  }
}else{
  load(paste0(owncache, "clustered.RData"))
}

```


```{r dimplots, fig.height=6, fig.width = 6}

plot_grid(ncol = 2, 
          DimPlot(alldata, reduction = "umap", group.by = "CCA_snn_res.0.25",
                  label = TRUE) + dimtheme + NoLegend() + NoAxes() + 
            ggtitle("louvain_0.25"),
          DimPlot(alldata, reduction = "umap", group.by = "CCA_snn_res.0.5", 
                  label = TRUE) + dimtheme +  NoLegend() + NoAxes() + 
            ggtitle("louvain_0.5"),
          DimPlot(alldata, reduction = "umap", group.by = "CCA_snn_res.1", 
                  label = TRUE) + dimtheme + NoLegend() + NoAxes() + 
            ggtitle("louvain_1"), 
          DimPlot(alldata, reduction = "umap", group.by = "CCA_snn_res.1.5", 
                  label = TRUE) + dimtheme + NoLegend() + NoAxes() + 
            ggtitle("louvain_1.5"))


```

```{r clustreeplot, fig.height = 3, fig.width = 6}

ct = clustree(alldata@meta.data, prefix = "CCA_snn_res.")

ct$layers[[1]]$aes_params$edge_width = 1

plot_grid(ggplot(unique(ct$data[,c("y","CCA_snn_res.")]),
       aes(y = y)) +
  geom_text(x = 1, aes(label = CCA_snn_res.), size = 3) + 
  scale_x_continuous(limits = c(0,2), position = "top") + 
  theme(panel.background = element_blank(),
        panel.grid = element_blank(),
        axis.text = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_text(size = 8),
        axis.ticks = element_blank()) + 
    labs(x= "Resolution"),
  ct + 
  scale_size_continuous(range = c(1,5)) + 
  scale_colour_discrete(guide = "none") + 
  theme(legend.key.size = unit(8,"pt"),
        legend.title = element_text(size = 8),
        legend.margin = margin(1,1,1,1,unit = "pt")),
  align = "h", rel_widths = c(1,8))

```

# Differential expression and annotation

The different clusters are annotated using differential expression and known marker genes. For legibility, the annotations are here given prior to differential expression and gene marker information.

```{r}

clusternames = 
  data.frame(Cluster = c(0:8),
             Clustername = c("Colonocytes","Goblet","TA G2-M",
                             "TA S-G2","Goblet Reg4","TA G1-S",
                             "ISC","EEC","Prolif + Infl"))

clusterorder = c("ISC","TA G1-S","TA S-G2","TA G2-M",
                 "Colonocytes","Prolif + Infl","Goblet",
                 "Goblet Reg4","EEC")


alldata = AddMetaData(alldata, 
                      factor(clusternames$Clustername[
                        match(alldata$CCA_snn_res.1,clusternames$Cluster)], 
                        levels = clusterorder),
                      "NamedCluster")

alldata = SetIdent(alldata, value = "NamedCluster")

color_list = get_color_list(9)
names(color_list) = clusterorder

```

```{r, fig.height = 4, fig.width = 4}

dp = DimPlot(alldata, label = TRUE, shuffle = TRUE) + 
  NoAxes() + NoLegend() + dimtheme + scale_color_manual(values = color_list)

dp$layers[[2]]$aes_params$size = 3

dp

```

```{r, fig.width = 6, fig.height = 8}

VlnPlot(alldata, features = c("nFeature_RNA", "nCount_RNA", "percent_mito", "percent_ribo", 
                              "S.Score", "G2M.Score"), 
    ncol = 2, pt.size = 0) & scale_fill_manual(values = color_list) & violintheme

```


## Differential expression

The top 25 DEGs (based on p-value) for each cluster are shown below, ordered by log2(FC).

```{r,  fig.height = 3, fig.width=2}

if(!file.exists(paste0(owncache, "clustermarkers.rds"))){

  markers_genes <- FindAllMarkers(alldata, logfc.threshold = 0.2, test.use = "wilcox", 
  min.pct = 0.1, min.diff.pct = 0.2, only.pos = TRUE, max.cells.per.ident = 1000, 
  assay = "RNA", random.seed = 42)
  
  if(file.exists(owncache)){
    saveRDS(markers_genes, paste0(owncache, "clustermarkers.rds"))
  }
}else{
  
  markers_genes = readRDS(paste0(owncache, "clustermarkers.rds"))
  
}

top25 <- markers_genes %>% group_by(cluster) %>% top_n(-25, p_val)

out = lapply(unique(top25$cluster), function(i){
  markersdata = top25[top25$cluster==i,]
  markersdata$gene = factor(markersdata$gene, 
                            levels = markersdata$gene[order(markersdata$avg_log2FC)])
  print(ggplot(markersdata, aes(x = avg_log2FC, y = gene)) + 
          labs(title = i) + 
    geom_bar(stat = "identity") + 
    theme(axis.title.y = element_blank(),
          axis.title.x = element_text(size = 8),
          plot.title = element_text(size = 8),
          axis.text = element_text(size = 8)))
  
})

```

```{r}

markergenes = 
  c("Stmn1","Lgr5","Ascl2","Smoc2","Hells","Mcm6","Pcna","Top2a","Mki67",
    "Hist1h2ap","Ccnb2","Cenpa","Cdc20","Muc3","Areg","Sycn","Wfdc2",
    "Lgals3","Car1","Ccl20","Ubd","Clu","Tnf","Muc2","Ang4","Agr2",
    "Ccl6","Mptx1","Reg4","Chga","Chgb","Pyy","Cck","Dclk1")

```




```{r, fig.height = 6, fig.width = 4}

dot = DotPlot(alldata, features=rev(markergenes), assay = "RNA")
dot = dot + aes(x = id, y = features.plot) + scale_x_discrete(position = "top") + 
  labs(y = "Genes",x = "Clusters") + 
  scale_size(range = c(0.5,2.5), breaks = c(0,50,100), limits = c(0,100)) +
  theme(legend.position = "none", legend.title = element_blank(),
        axis.text.x = element_text(size = 8, angle = -45, hjust = 1),
        axis.text.y = element_text(size = 8, face = "italic"),
        legend.key.width = unit(0.3, "cm"), 
        legend.key.height = unit(0.3,"cm"), 
        legend.text = element_text(size = 8),
        axis.title = element_text(size = 8))
dot$scales$scales[[1]]$limits = c(-1.8,2.7)

plot_grid(dot,get_legend(dot + theme(legend.position = "right")),
          rel_widths = c(5,1))

```


# Population changes over time

The proportion of each cluster among the total stromal population is shown across time points below. The order of clusters is based on the average correlation between proportion and time point for each background (untreated samples only), with positive correlation at the top and negative correlation on the bottom.

```{r}


freqs = table(alldata$orig.ident, alldata$NamedCluster)

df = as.data.frame(apply(freqs,1,function(x){x/sum(x)}))
df$Cluster = rownames(df)

fulldf = melt(df)
fulldf$variable = as.numeric(fulldf$variable)

fulldf = fulldf[!is.nan(fulldf$value),]


```



```{r}


frequencies = merge(melt(table(data.frame(cluster=alldata$NamedCluster,
                 sample = alldata$orig.ident)), value.name = "Freq"),
                 melt(apply(table(data.frame(cluster=alldata$NamedCluster,
                                             sample = alldata$orig.ident)),2,
                            function(x){x/sum(x)}), value.name = "Proportion"))

frequencies$time = as.numeric(gsub("W","",alldata$Week[match(frequencies$sample, alldata$orig.ident)]))
frequencies$genotype = factor(alldata$Genotype[match(frequencies$sample, alldata$orig.ident)],
                              levels = c("WT","Il10KO"))

frequencies = frequencies[order(frequencies$cluster),]

cumulFreq = frequencies
cumulFreq$cumulProp = NA
for(i in unique(cumulFreq$sample)){
  cumulFreq$cumulProp[cumulFreq$sample %in% i] = rev(cumsum(rev(cumulFreq$Proportion[cumulFreq$sample %in% i])))
}
cumulFreq = rbind(cumulFreq,cumulFreq)
cumulFreq$xadj = rep(c(-1,1), c(nrow(frequencies),nrow(frequencies)))


```


```{r, fig.height = 7}

corlist = lapply(unique(frequencies$cluster), function(i){
  clufreq = frequencies[frequencies$cluster==i  & 
                        frequencies$genotype == "Il10KO",]

  clucor =  cor(clufreq$Proportion,as.numeric((clufreq$time)))
  clucor
  })
names(corlist) = unique(frequencies$cluster)

averagecor = unlist(corlist)

frequencies$cluster = 
  factor(frequencies$cluster, 
         levels = names(averagecor)[order(unlist(averagecor), decreasing = TRUE)])
cumulFreq$cluster = 
  factor(cumulFreq$cluster,
         levels = names(averagecor)[order(unlist(averagecor), decreasing = TRUE)])

frequencies$sample = factor(frequencies$sample, 
                          levels = 
                            unique(frequencies$sample[order(frequencies$genotype,
                                                   frequencies$time)]))
frequencies$samplenumber = as.numeric(frequencies$sample)


cumulFreq$sample = factor(cumulFreq$sample, 
                          levels = 
                            levels(frequencies$sample))
cumulFreq$samplenumber = as.numeric(cumulFreq$sample)


```


```{r, fig.height = 3.5, fig.width = 3.5}


datasets = lapply(unique(frequencies$genotype), function(bg){
  bgfreq = frequencies[frequencies$genotype == bg,]
  
  cumbgfr = cumulFreq[paste(cumulFreq$sample,
                                   cumulFreq$oldtime) %in% 
                               paste(bgfreq$sample,
                                     bgfreq$oldtime),]
  list(bgfreq, cumbgfr)
})


barplot = ggplot(data = frequencies, 
       aes(x = samplenumber, y = Proportion, fill = cluster)) + 
  theme(panel.background = element_blank(), axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
            scale_fill_manual(values =
                                color_list[as.character(unique(sort(frequencies$cluster)))]) + 
            scale_x_continuous(breaks = sort(unique(frequencies$samplenumber)),
                               labels = 
                                 gsub("BALBc","WT",
                                      gsub("_IMM","",
                                           gsub("G_","", 
                                                gsub("IL10_","",
                                                     levels(frequencies$sample))))))

for(ds in datasets){
  barplot = barplot + 
    geom_area(data = ds[[2]],
              alpha = 0.7, linewidth = 0.3, stat = 'identity',
              aes(fill = cluster, group = cluster,
                  y = Proportion, x = samplenumber + xadj*0.3))
}

barplot + geom_bar(stat = "identity", width = 0.6, color = "black") + 
  theme(legend.key.size = unit(8,"pt"))


```

# Subset analyses

To evaluate the changes of gene expression specfically in colonocytes and progenitors as well as goblet cells, these clusters are isolated. The UMAP shows the included cells for each analysis. Subsets are saved to `r datadir`.

```{r}

alldata = AddMetaData(alldata, alldata$NamedCluster, "stemcolono")
alldata = AddMetaData(alldata, alldata$NamedCluster, "goblet")

alldata$stemcolono[!(alldata$stemcolono %in% 
                  c("ISC","TA G1-S","TA S-G2","TA G2-M",
                 "Colonocytes"))] = NA

alldata$goblet[!(alldata$goblet %in% 
                  c("Goblet", "Goblet Reg4"))] = NA


```

```{r, fig.height = 3, fig.width = 9}

plot_grid(plotlist = lapply(c("stemcolono","goblet"),
       function(x){
         dp = DimPlot(alldata, group.by = x, label = TRUE, shuffle = TRUE) +
            scale_color_manual(values = color_list, na.value = "#CACACA") +
            dimtheme & NoLegend() & NoAxes()
         dp$layers[[2]]$aes_params$size = 3
         dp
       }), ncol = 3)

```

```{r}

alldata@active.assay = "RNA"
alldata[["CCA"]] = NULL

stemcolono = 
  subset(alldata, 
         cells = WhichCells(alldata, expression = 
                              stemcolono %in% alldata$NamedCluster))
goblet = subset(alldata, cells = WhichCells(alldata, expression = goblet %in% alldata$NamedCluster))

rm(alldata)
trash = gc()

saveRDS(stemcolono, file = paste0(datadir,"colonocytesandprogenitors.rds"))
saveRDS(goblet, file = paste0(datadir,"gobletcells.rds"))

```



# Session info

```{r}

sessionInfo()

```
