---
title: "Preparation of data for Il10KO single cell analysis"
editor_options:
  chunk_output_type: inline
date: "27 February, 2024"
output:
  bookdown::html_document2:
    code_folding: hide
    toc: yes
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
params:
  dataset: 1
---

```{r Setup, echo = FALSE}
knitr::opts_chunk$set(fig.width  = 7,
                      results    = "hold",
                      message    = FALSE,
                      cache.lazy = FALSE,
                      warning    = FALSE)


```




```{r setup}
 
library(Seurat)
library(Matrix)
library(ggplot2)

library(cowplot)
library(DoubletFinder)
# library(clustree)
# 
library(future)
library(future.apply)
 
datadir = "../data/SingleCell/Il10KO/expression_matrices"

owncache = "../data/SingleCell/Il10KO/cache/"

violintheme = theme(axis.text = element_text(size = 8),
        plot.title = element_text(size = 8),
        legend.key.size = unit(8, "pt"),
        axis.title = element_blank())
dimtheme = theme(legend.text = element_text(size = 8),
        plot.title = element_text(size = 8),
        legend.key.size = unit(8, "pt"))

dotviolin = function(v){
  v[[1]]$layers[[2]]$aes_params$shape = 16
  v[[2]]$layers[[2]]$aes_params$shape = 16
  v[[3]]$layers[[2]]$aes_params$shape = 16
  v[[4]]$layers[[2]]$aes_params$shape = 16
  v[[1]]$layers[[2]]$aes_params$alpha = 0.3
  v[[2]]$layers[[2]]$aes_params$alpha = 0.3
  v[[3]]$layers[[2]]$aes_params$alpha = 0.3
  v[[4]]$layers[[2]]$aes_params$alpha = 0.3
  v & violintheme
}

metadata = read.csv2(paste0(datadir,"/../metadata.csv"))
sampleorder = c('BALBc_6W_EPI', 'BALBc_6W_IMM', 'BALBc_6W_STR',
                'G_10W_EPI', 'G_10W_IMM', 'G_10W_STR',
                'G_12W_EPI', 'G_12W_IMM', 'G_12W_STR',
                'G_14W_EPI', 'G_14W_IMM', 'G_14W_STR',
                'G_8W_EPI', 'G_8W_IMM', 'G_8W_STR',
                'IL10_6W_EPI', 'IL10_6W_IMM', 'IL10_6W_STR')


qc_feats <- c("nFeature_RNA", "nCount_RNA", "percent_mito", "percent_ribo")

```

```{r, include = FALSE}

if(file.exists(owncache)){
  message(paste0("Intermediate data files will be loaded from and saved to ",owncache))
}else{
  message("Intermediate data files will not be saved or loaded")
}

```

# Gene expression matrices

The gene expression matrices were produced using CellRanger on the raw .fastq-files. These files can be found at 

```{r loading_matrices}

if(!file.exists(paste0(owncache,"/loading_matrices.RData"))){

plan("multisession", workers = 4, gc = TRUE)

loadObject = function(d){
  sample = sub("\\.h5","",sub(".*/","",d))
  expression_mat = Read10X_h5(filename = d)
  if(class(expression_mat)=="dgCMatrix"){
    data_obj = CreateSeuratObject(expression_mat, sample)
  }else{
      data_obj = CreateSeuratObject(expression_mat$`Gene Expression`, sample)
      data_obj[["FB"]] = CreateAssayObject(expression_mat$`Antibody Capture`)
      data_obj@active.assay = "RNA"
  }
  data_obj
}

system.time(
  {
    datafiles = list.files(datadir, full.names = TRUE)
    all_matrices = future_lapply(datafiles, function(d){
      message(d)
      sample = sub("\\.h5","",sub(".*/","",d))
      message(sample)
  loadObject(d)
})
names(all_matrices) = sub("\\.h5","",sub(".*/","",datafiles))
gc()})

alldata = merge(x = all_matrices[[sampleorder[1]]],
                y = all_matrices[sampleorder[-1]],
                add.cell.ids = sampleorder)

  immunesamples = sampleorder[grep("_IMM",sampleorder)]
  immune_matrices = lapply(all_matrices[immunesamples], 
                         function(x){
                           x@active.assay = "FB"
                           x[["RNA"]] = NULL
                           x
                         })
  rm(all_matrices)

  alldata_FB = merge(immune_matrices[[1]],
                   immune_matrices[-1],
                   add.cell.ids = immunesamples)
  
  if(file.exists(owncache)){
    save(alldata,alldata_FB, file = paste0(owncache,"/loading_matrices.RData"))
  }

rm(loadObject)
gc()

plan("sequential")


}else{
  load(paste0(owncache,"/loading_matrices.RData"))
}


```

```{r addmetadata}


colnames(metadata) = gsub("X","",gsub("\\.","",colnames(metadata)))

alldata = AddMetaData(alldata,
                      factor(metadata[match(alldata$orig.ident, metadata$libraryname),"genotype"],
                             levels = c("WT","Il10KO")),
                      "Genotype")
alldata = AddMetaData(alldata, metadata[match(alldata$orig.ident, metadata$libraryname),"time"],
                      "Week")
alldata = AddMetaData(alldata,
                      metadata[match(alldata$orig.ident, metadata$libraryname),"celltype"],
                      "Type")



alldata <- PercentageFeatureSet(alldata, "^mt-", col.name = "percent_mito")

alldata <- PercentageFeatureSet(alldata, "^Rp[sl]", col.name = "percent_ribo")

```

# Plot QC vars

Number of features and counts and percentage of mitochondrial and ribosomal RNA are plotted below.

```{r qc, fig.height=4, fig.width=6}


VlnPlot(alldata, group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) &
  violintheme


```


# Filter cells and genes

The same QC as above are plotted below, only including cells and genes that pass filtering limits (gene limit: > 3 copies expressed, cell filter: > 200 unique genes). The top expressed genes are also shown. Mitochondrial and ribosomal genes are filtered out of the data, as are the highly expressed genes Malat1 and Gm42418.

```{r filtering}

if(!file.exists(paste0(owncache,"/filtering.RData"))){


alldata <- subset(alldata, features = rownames(alldata)[Matrix::rowSums(alldata) > 3],
                  cells = intersect(WhichCells(alldata, expression = nFeature_RNA > 200),
                                    intersect(WhichCells(alldata, expression = percent_mito < 10),
                                    WhichCells(alldata, expression = percent_ribo > 5))))

gc()
save(alldata, file = paste0(owncache,"/filtering.RData"))

}else{
  rm(alldata)
  gc()
  load(paste0(owncache,"/filtering.RData"))
}

```

```{r, fig.height=4, fig.width=6}

VlnPlot(alldata, group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) &
    NoLegend() & violintheme


VlnPlot(alldata,
        idents = unique(alldata@meta.data$orig.ident[alldata@meta.data$Type == "Immune"]),
        group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) +
    NoLegend() & violintheme
VlnPlot(alldata,
        idents = unique(alldata@meta.data$orig.ident[alldata@meta.data$Type == "Stroma"]),
        group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) +
    NoLegend() & violintheme
VlnPlot(alldata,
        idents = unique(alldata@meta.data$orig.ident[alldata@meta.data$Type == "Epithelial"]),
        group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) +
    NoLegend() & violintheme

```

```{r topexpressed, fig.height = 6, fig.width = 6, cache = TRUE}

par(mar = c(4, 8, 2, 1))
C <- subset(alldata, downsample = 1000, seed = 42)@assays$RNA@counts
C <- Matrix::t(Matrix::t(C)/Matrix::colSums(C)) * 100
#nC = ncol(C)

most_expressed <- order(apply(C, 1, median), decreasing = T)[20:1]

boxplot(as.matrix(t(C[most_expressed, ])), cex = 0.1, las = 1, xlab = "% total count per cell",
     col = (scales::hue_pal())(20)[20:1], horizontal = TRUE)


rm(C)
rm(most_expressed)



# Filter MALAT1, Gm42418, mitochondrial genes and ribosomal genes
alldata <- alldata[!grepl("Malat1", rownames(alldata)) &
                     !grepl("Gm42418", rownames(alldata)) &
                     !grepl("^mt-", rownames(alldata)) & !
                     grepl('^Rp[sl]', rownames(alldata)), ]


```


`r knitr::kable(table(alldata$orig.ident))`




# Cell cycle scores

The S-phase and G2M-phase scores are calculated based on the cell cycle gene lists defined in the Seurat package (translated to mouse genes originally using biomaRt).

```{r cellcyclescores}

if(!file.exists(paste0(owncache,"/cellcyclescores.RData"))){

  g2mFeat = c("Smc4","Dlgap5","Ctcf","Cdc25c","Gtse1","Kif20b",
              "Ttk","Tacc3","Top2a","Tpx2","Cdca3","Ncapd2",
              "Cks2","Anp32e","G2e3","Lbr","Cdca2","Cks1brt",
              "Ckap2","Hmgb2","Kif11","Nek2","Cenpe","Hjurp",
              "Ect2","Aurkb","Cks1b","Kif23","Nuf2","Hmmr",
              "Cdca8","Psrc1","Anln","Cdc20","Birc5","Ndc80",
              "Rangap1","Ckap5","Kif2c","Cenpf","Nusap1","Cenpa",
              "Aurka","Ube2c","Ckap2l","Mki67","Tubb4b","Bub1",
              "Ccnb2")
  sFeat = c("Msh2","Exo1","Mcm4","Rrm2","Mcm2","Chaf1b","Gmnn",
            "Cdc45","Slbp","Ubr7","Cdc6",
            "Rad51ap1","Rpa2","Hells","Fen1","Gins2","Uhrf1",
            "Mcm6","Ung","Dscc1","Usp1","Clspn",
            "Cdca7","Pola1","Nasp","Dtl","Mcm5","Wdr76","Prim1",
            "Casp8ap2","Tipin","Blm","Rrm1",
            "Brip1","Rad51","Tyms","E2f8","Rfc2","Ccne2","Pcna")




# Before running CellCycleScoring the data need to be normalized and
# logtransformed.

alldata = NormalizeData(alldata)

alldata <- CellCycleScoring(object = alldata, g2m.features = g2mFeat,
    s.features = sFeat)

rm(g2mFeat)
rm(sFeat)
gc()


if(file.exists(owncache)){
  save(alldata, file = paste0(owncache,"/cellcyclescores.RData"))
}

}else{
  rm(alldata)
  gc()
  load(paste0(owncache,"/cellcyclescores.RData"))
}


```


```{r, fig.width = 6, fig.height = 3}

VlnPlot(alldata, features = c("S.Score", "G2M.Score"), group.by = "orig.ident",
        pt.size = 0) & violintheme

```


# Find Doublets

Doublets are identified and excluded using the DoubletFinder package.



```{r doublets}

if(!file.exists(paste0(owncache,"/doublets.RData"))){

  
  if(!file.exists(paste0(owncache,"/doublet_prep.RData"))){
    
      alldata = FindVariableFeatures(alldata)
      alldata = ScaleData(alldata, vars.to.regress = c("nFeature_RNA", "percent_mito"))
      alldata = RunPCA(alldata, npcs = 20)
      alldata = RunUMAP(alldata, dims = 1:10)
      if(file.exists(owncache)){
      save(alldata, file = paste0(owncache,"/doublet_prep.RData"))
    }
  
  }else{
    rm(alldata)
    gc()
    load(paste0(owncache,"/doublet_prep.RData"))
  }
  

  if(!file.exists(paste0(owncache,"sweeplist.rds"))){
    sweeplist = list()

    for(i in unique(alldata$orig.ident)){
      message(i)
      data.filt=subset(alldata, idents = i)
      sweep.res <- paramSweep_v3(data.filt)#, num.cores = 3)
      sweep.stats <- summarizeSweep(sweep.res, GT = FALSE)
      bcmvn <- find.pK(sweep.stats)
  
      sweeplist[[i]] = bcmvn }
    saveRDS(sweeplist, paste0(owncache,"sweeplist.rds"))
  }else{
    sweeplist = readRDS(paste0(owncache,"sweeplist.rds"))
  }
  
  if(!file.exists(paste0(owncache,"singletList.rds"))){
    singletList = list()
    for(i in unique(alldata$orig.ident)){
      message(i)
      
      data.filt=subset(alldata, idents = i)
    nExp <- round(ncol(data.filt) *
                    ifelse(ncol(data.filt)<12000,0.08,
                           ifelse(ncol(data.filt)<22000,0.16,0.24)))
    # expect 8%-24% doublets
    mypK = as.numeric(as.character(sweeplist[[i]]$pK[
      which.max(sweeplist[[i]]$BCmetric)]))
    data.filt <- doubletFinder_v3(data.filt, pN = 0.25,
                                  pK = mypK, nExp = nExp, PCs = 1:10)
    DF.name = colnames(data.filt@meta.data)[grepl("DF.classification",
                                                  colnames(data.filt@meta.data))]
    singletList[[i]] = data.filt@meta.data[, DF.name]
    names(singletList[[i]]) = colnames(data.filt)

    rm(data.filt)
    gc()
    }
    saveRDS(singletList, paste0(owncache,"singletList.rds"))
  }else{
    singletList = readRDS(paste0(owncache,"singletList.rds"))
  }

alldata = AddMetaData(alldata, unlist(singletList)[match(colnames(alldata), gsub(".*\\.","",names(unlist(singletList))))], "DF")


if(file.exists(owncache)){
  save(alldata, file = paste0(owncache,"/doublets.RData"))
  saveRDS(alldata@meta.data,
           file = paste0(owncache,"/doublets_metadata.rds"))
}

}else{
  rm(alldata)
  gc()
  load(paste0(owncache,"/doublets.RData"))
}


```

```{r, fig.height = 3, fig.width = 6}

plot_grid(ncol = 2, DimPlot(alldata, group.by = "orig.ident") + NoAxes() + NoLegend() + dimtheme,
    DimPlot(alldata, group.by = "DF") + NoAxes()) + dimtheme


```



```{r df_dimplot, fig.height = 4, fig.width=4}

DimPlot(alldata, group.by = "DF", split.by = "orig.ident", ncol = 6) + NoAxes() & 
  theme(plot.title = element_text(size = 8),
        strip.text = element_text(size = 8),
        legend.position = "bottom") + dimtheme 

```


```{r df_violins, fig.height = 2, fig.width = 4}

VlnPlot(alldata, features = "nFeature_RNA", group.by = "DF", pt.size = 0) + violintheme
VlnPlot(alldata, features = "nFeature_RNA", split.by = "DF", group.by = "orig.ident", pt.size = 0)  &
  theme(plot.title = element_text(size = 8),
        axis.text = element_text(size = 8)) + violintheme


```

`r knitr::kable(table( alldata$orig.ident, alldata$DF))`

```{r}

alldata = alldata[, alldata@meta.data[, "DF"] == "Singlet"]

```



```{r cellcounts, fig.height = 3, fig.width = 4}

cellfreq = as.data.frame(table(alldata$orig.ident), stringsAsFactors = FALSE)
cellfreq$Type = alldata$Type[match(cellfreq$Var1, alldata$orig.ident)]
ggplot(cellfreq, aes(x = Var1, y = Freq, color = Type)) + geom_point() +
  theme_classic() +
  theme(panel.background = element_blank(),
        legend.key = element_blank(),
        panel.grid.major.y = element_line(color = "#DDDDDD"),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Sample", y = "Number of cells")

```


```{r, fig.height = 3, fig.width = 6}

VlnPlot(alldata, group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) +
    NoLegend() &
  violintheme

```


```{r typeviolins, fig.height = 3, fig.width = 6}

VlnPlot(alldata,
        idents = unique(alldata@meta.data$orig.ident[alldata@meta.data$Type == "Immune"]),
        group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) +
    NoLegend() &
  violintheme
VlnPlot(alldata,
        idents = unique(alldata@meta.data$orig.ident[alldata@meta.data$Type == "Stroma"]),
        group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) +
    NoLegend() &
  violintheme
VlnPlot(alldata,
        idents = unique(alldata@meta.data$orig.ident[alldata@meta.data$Type == "Epithelial"]),
        group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) +
    NoLegend() &
  violintheme



```




```{r firstnormalize}

if(!file.exists(paste0(owncache,"/firstnormalize.RData"))){

alldata = NormalizeData(alldata)

suppressWarnings(suppressMessages(alldata <- FindVariableFeatures(alldata, selection.method = "vst",
    nfeatures = 2000, verbose = FALSE, assay = "RNA")))
alldata <- ScaleData(alldata, vars.to.regress = c("percent_mito", "nFeature_RNA"),     assay = "RNA")

alldata <- RunPCA(alldata, npcs = 50, reduction.name = "pca_raw")
alldata <- RunUMAP(alldata, reduction = "pca_raw", dims = 1:30, n.components = 2, n.neighbors = 30,
    n.epochs = 200, min.dist = 0.3, learning.rate = 1, spread = 1, reduction.name = "umap_raw")


if(file.exists(owncache)){
  save(alldata, file = paste0(owncache,"/firstnormalize.RData"))
}

}else{
  rm(alldata)
  gc()
  load(paste0(owncache,"/firstnormalize.RData"))
}

```


```{r topgenes, fig.height = 3, fig.width = 5}

top20 <- head(VariableFeatures(alldata), 20)

LabelPoints(plot = VariableFeaturePlot(alldata), points = top20, repel = TRUE) +
  theme(axis.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        axis.text = element_text(size = 8))

```


# PCA

```{r firstpca, fig.height = 2, fig.width = 4}

plot_grid(plot_grid(ncol = 3,
          DimPlot(alldata, reduction = "pca_raw", group.by = "orig.ident",
                  dims = 1:2, shuffle = TRUE) + NoLegend() + NoAxes() + dimtheme,
          DimPlot(alldata, reduction = "pca_raw", group.by = "orig.ident",
                  dims = 3:4, shuffle = TRUE) + NoLegend() + NoAxes() + dimtheme,
          DimPlot(alldata, reduction = "pca_raw", group.by = "orig.ident",
                  dims = 5:6, shuffle = TRUE) + NoLegend() + NoAxes() + dimtheme),
          get_legend(DimPlot(alldata, reduction = "pca_raw", group.by = "orig.ident",
                  dims = 1:2, shuffle = TRUE) + theme(legend.position = "bottom") + dimtheme),
          ncol = 1, rel_heights = c(3,1)
)
plot_grid(plot_grid(ncol = 3,
          DimPlot(alldata, reduction = "pca_raw", group.by = "Type",
                  dims = 1:2, shuffle = TRUE) + NoLegend() + NoAxes() + dimtheme,
          DimPlot(alldata, reduction = "pca_raw", group.by = "Type",
                  dims = 3:4, shuffle = TRUE) + NoLegend() + NoAxes() + dimtheme,
          DimPlot(alldata, reduction = "pca_raw", group.by = "Type",
                  dims = 5:6, shuffle = TRUE) + NoLegend() + NoAxes() + dimtheme),
          get_legend(DimPlot(alldata, reduction = "pca_raw", group.by = "Type",
                  dims = 1:2, shuffle = TRUE) + theme(legend.position = "bottom") + dimtheme),
          ncol = 1, rel_heights = c(3,1))

```

```{r, fig.height = 3, fig.width = 6}

VizDimLoadings(alldata, dims = 1:5, reduction = "pca_raw", ncol = 5, balanced = T) &
  theme(axis.text = element_text(size = 8))

```

```{r, fig.height = 2, fig.width = 4}

ElbowPlot(alldata, reduction = "pca_raw", ndims = 50) + 
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 8))


```


# UMAP

```{r firstumap, fig.height = 3, fig.width = 6}

plot_grid(ncol = 3, DimPlot(alldata, reduction = "umap_raw", group.by = "orig.ident",
                            shuffle = TRUE) + NoAxes() + NoLegend() + dimtheme,
    DimPlot(alldata, reduction = "umap_raw", group.by = "Week",
                            shuffle = TRUE) + NoAxes() + NoLegend() + dimtheme,
    DimPlot(alldata, reduction = "umap_raw", group.by = "Type", label = TRUE,
                            shuffle = TRUE) + NoAxes() + NoLegend() + dimtheme)

```


# Integrate data




```{r normalizeintegrated}



if(!file.exists(paste0(owncache,"/normalizeintegrated.RData"))){

  if(!file.exists(paste0(owncache,"/fullintegration.RData"))){
    
    if(!file.exists(paste0(owncache,"/getanchors.RData"))){

      alldata <- AddMetaData(alldata, gsub("_IMM","",gsub("_STR","",
                                  gsub("_EPI","",alldata$orig.ident))), 
                           "source")
      data.list <- SplitObject(alldata, split.by = "source")
      rm(alldata)
      gc()
      
      message("object split")
      
      data.list = future_lapply(data.list, function(x){
        message("sample")
          x <- NormalizeData(x, verbose = FALSE)
          x <- FindVariableFeatures(x, selection.method = "vst",
              nfeatures = 2000, verbose = FALSE)
          x
      })
      
      message("data normalized")
      
      features <- SelectIntegrationFeatures(object.list = data.list)
      
      message("features selected")
      
      
      data.list <- future_lapply(X = data.list, FUN = function(x) {
        x <- ScaleData(x, features = features, verbose = FALSE)
        x <- RunPCA(x, features = features, verbose = FALSE)
      })
      
      message("pcas run")
      
      data.anchors <- FindIntegrationAnchors(object.list = data.list, 
                                             dims = 1:30,
          reduction = "rpca", anchor.features = features)
      
      message("anchors found")
      
      rm(data.list)
      gc()
      
      
      if(file.exists(owncache)){
        save(data.anchors, file = paste0(owncache,"/getanchors.RData"))
      }
      
      }else{
        rm(alldata)
        gc()
        load(paste0(owncache,"/getanchors.RData"))
      }
    
    
    
    alldata <- IntegrateData(anchorset = data.anchors, 
                             dims = 1:30, new.assay.name = "RPCA")
    rm(data.anchors)
    gc()
    if(file.exists(owncache)){
      save(alldata, file = paste0(owncache,"/fullintegration.RData"))
    }
  }else{
    rm(alldata)
    gc()
    load(paste0(owncache,"/fullintegration.RData"))
  }
  
  
  alldata <- ScaleData(alldata)
  alldata <- RunPCA(alldata, npcs = 30)
  alldata <- RunUMAP(alldata, dims = 1:30)
  if(file.exists(owncache)){
    save(alldata, file = paste0(owncache,"/normalizeintegrated.RData"))
  }

}else{
  rm(alldata)
  gc()
  load(paste0(owncache,"/normalizeintegrated.RData"))
}

```


```{r, fig.height = 3, fig.width = 6}

plot_grid(ncol = 3,


  DimPlot(alldata, reduction = "pca", group.by = "orig.ident", shuffle = TRUE) +
    NoAxes() + ggtitle("PCA integrated") + NoLegend() + dimtheme,
  DimPlot(alldata, reduction = "umap", group.by = "orig.ident", shuffle = TRUE) +
    NoAxes() + ggtitle("UMAP integrated") + NoLegend() + dimtheme,
  get_legend(DimPlot(alldata, reduction = "pca", group.by = "orig.ident", shuffle = TRUE) +
    NoAxes() + ggtitle("PCA integrated") + dimtheme)
)

```


```{r, fig.height = 6, fig.width = 6}

DimPlot(alldata, reduction = "umap", split.by = "orig.ident", 
        ncol = 6)+NoAxes() + NoLegend() & theme(strip.text= element_text(size = 8)) + 
  dimtheme

```


```{r, fig.height = 3, fig.width = 3}

DimPlot(alldata, reduction = "umap", group.by = "Genotype", 
        shuffle = TRUE) + NoAxes() + dimtheme


```

# Plot QC in UMAP

```{r, fig.height = 4, fig.width = 4}

FeaturePlot(alldata, reduction = "umap",
            features = qc_feats,
    order = T, slot = "data", combine = T) & NoAxes() &
  theme(plot.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        legend.key.size = unit(8,"pt"))

```



# Plot genes of interest





```{r, fig.height = 4, fig.width = 4}

DefaultAssay(alldata) = "RNA"

FeaturePlot(alldata, reduction = "umap",
            features = c("Mki67","Pcna", "Ptprc","Epcam"),
    order = T, slot = "data", combine = T) & NoAxes() &
  theme(plot.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        legend.key.size = unit(8,"pt"))

DefaultAssay(alldata) = "RPCA"

```


# Clustering to identify sorting contaminations


```{r}

if(!file.exists(paste0(owncache,"/fullclustering.RData"))){


  alldata@active.assay = "RPCA"
  
  alldata <- FindNeighbors(alldata, dims = 1:30,
                           k.param = 350, prune.SNN = 1/15)
  
  alldata <- FindClusters(alldata, graph.name = "RPCA_snn",
                          resolution = 0.25, algorithm = 1)
  
  alldata = AddMetaData(alldata, alldata$RPCA_snn_res.0.25,"RPCA_cluster")
  alldata$seurat_clusters = NULL
  alldata$RPCA_snn_res.0.25 = NULL
  
  save(alldata, file = paste0(owncache,"/fullclustering.RData"))
}else{
  rm(alldata)
  gc()
  load(paste0(owncache,"/fullclustering.RData"))
}


```


```{r, fig.height = 2, fig.width = 2}

DimPlot(alldata, label = TRUE) + NoAxes() + NoLegend() + dimtheme

```

```{r, fig.height = 3, fig.width = 6}

DefaultAssay(alldata) = "RNA"
DotPlot(alldata, features = c("Epcam","Ptprc")) + coord_flip() + 
  theme(legend.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        axis.text = element_text(size = 8),
        axis.title = element_text(size = 8))
DefaultAssay(alldata) = "RPCA"

```


```{r, fig.height = 3, fig.width = 5}

ggplot(alldata@meta.data, aes(fill = Type, x = RPCA_cluster)) + geom_bar(position = "fill")

ggplot(alldata@meta.data, aes(fill = orig.ident, x = RPCA_cluster)) +
  geom_bar(position = "fill") 


```


```{r, fig.height = 3, fig.width = 6}

VlnPlot(alldata, group.by = "RPCA_cluster", features = qc_feats, pt.size = 0, ncol = 2) &
  theme(plot.title = element_text(size = 8),
        axis.text = element_text(size = 8),
        axis.title = element_blank()) + violintheme

```



# Data cleaning

```{r cleandata}

if(!file.exists(paste0(owncache,"/lastscale.RData")) |
   !file.exists(paste0(datadir,"/../R_objects/immune_data.rds")) |
   !file.exists(paste0(datadir,"/../R_objects/stroma_data.rds")) |
   !file.exists(paste0(datadir,"/../R_objects/epithelial_data.rds"))){

  
  
  immuneclusters = c(0,1,3,6,7,13,14,15,18)
stromaclusters = c(2,4,8,9,10,12,16,17,20,21)
epiclusters = c(5,11,19)

sel.clust = "RPCA_cluster"

alldata <- SetIdent(alldata, value = "Type")

immunetype = WhichCells(alldata, ident = "Immune")
stromatype = WhichCells(alldata, ident = "Stroma")
epitype = WhichCells(alldata, ident = "Epithelial")

alldata <- SetIdent(alldata, value = sel.clust)
alldata$ident_clust = paste0(alldata@active.ident,"_",alldata$orig.ident)

immunecells = WhichCells(alldata, ident = immuneclusters, cells = immunetype)
stromacells = WhichCells(alldata, ident = stromaclusters, cells = stromatype)
epicells = WhichCells(alldata, ident = epiclusters, cells = epitype)

alldata = subset(alldata, cells = c(immunecells,stromacells, epicells))

cleandata = alldata

rm(alldata)

gc()

immunedata = subset(cleandata, cells = immunecells)
immunedata@active.assay = "RNA"
immunedata[["RPCA"]] = NULL
immunedata[["FB"]] = subset(alldata_FB, cells = Cells(immunedata))[["FB"]]
saveRDS(immunedata, paste0(datadir,"/../R_objects/immune_data.rds"))
rm(immunedata)

stromadata = subset(cleandata, cells = stromacells)
stromadata@active.assay = "RNA"
stromadata[["RPCA"]] = NULL
saveRDS(stromadata, paste0(datadir,"/../R_objects/stroma_data.rds"))
rm(stromadata)

epidata = subset(cleandata, cells = epicells)
epidata@active.assay = "RNA"
epidata[["RPCA"]] = NULL
saveRDS(epidata, paste0(datadir,"/../R_objects/epithelial_data.rds"))
rm(epidata)

gc()






cleandata <- ScaleData(cleandata)
cleandata <- RunPCA(cleandata, npcs = 30)
cleandata <- RunUMAP(cleandata, dims = 1:30)
save(cleandata, file = paste0(owncache,"/lastscale.RData"))
}else{
  rm(alldata)
  gc()
  load(paste0(owncache,"/lastscale.RData"))
}
```


## UMAPs

```{r dimplots,fig.height = 2, fig.width = 2}


DimPlot(cleandata, reduction = "umap", label = TRUE,
                        group.by = "RPCA_cluster", shuffle = TRUE) +
            NoAxes() + labs(title = "Clustering") + NoLegend() + 
  theme(plot.title = element_text(size = 10)) + dimtheme


```




```{r, fig.height = 3, fig.width=7}

DimPlot(cleandata, reduction = "umap", split.by = "Type") +
  NoAxes() + NoLegend() + dimtheme

```




```{r umaps, fig.height = 3, fig.width=3}

typecolors = c(Stroma = "#F4DF9E",Immune = "#D6A7A0", Epithelial = "#A6D8DA")
d_timecolors = scale_color_viridis_d(option = "viridis", direction = 1)
modelcolors = c(WT = "#9E68B1", Il10KO = "#FCA549")



DimPlot(cleandata, reduction = "umap", group.by = "Week", shuffle = TRUE) +
            NoAxes() + d_timecolors + dimtheme
DimPlot(cleandata, reduction = "umap", group.by = "Type", shuffle = TRUE) +
            NoAxes() + labs(title = "Sort fraction") +
             scale_color_manual(values = typecolors) + dimtheme


DimPlot(cleandata, reduction = "umap", group.by = "Genotype", 
                  shuffle = TRUE) +
            NoAxes() + 
             scale_color_manual(values = modelcolors) + dimtheme



```




# Session info

```{r}

sessionInfo()

```
