---
title: "Il10KO single cell analysis (Stroma)"
editor_options:
  chunk_output_type: inline
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    code_folding: hide
    toc: yes
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
params:
  dataset: 1
---

```{r Setup, echo = FALSE}
knitr::opts_chunk$set(fig.width  = 7,
                      results    = "hold",
                      message    = FALSE,
                      cache.lazy = FALSE,
                      warning    = FALSE)


```




```{r setup}
 
library(Seurat)
library(ggplot2)
library(cowplot)
library(clustree)


source("../code/SingleCell/color_list.R")
source("../code/SingleCell/NMFsetup.R", chdir = TRUE)

```


```{r}
 
datadir = "../data/SingleCell/Il10KO/R_objects/"

owncache = "../data/SingleCell/Il10KO/cache_stroma/"

violintheme = theme(axis.text = element_text(size = 8),
        plot.title = element_text(size = 8),
        legend.key.size = unit(8, "pt"),
        axis.title = element_blank())
dimtheme = theme(legend.text = element_text(size = 8),
        plot.title = element_text(size = 8),
        legend.key.size = unit(8, "pt"))

qc_feats <- c("nFeature_RNA", "nCount_RNA", "percent_mito", "percent_ribo")

```

```{r, include = FALSE}

if(file.exists(owncache)){
  message(paste0("Intermediate data files will be loaded from and saved to ",owncache))
}else{
  message("Intermediate data files will not be saved or loaded")
}

```

# Initial data 

Due to a large heterogeneity of cell quality in the week 10 sample, a filtering step is added to remove cells with < 2000 unique genes expressed in this sample. 

```{r}

if(!file.exists(paste0(owncache,"first_dr.RData"))){
  alldata = readRDS(paste0(datadir,"stroma_data.rds"))

  alldata = subset(alldata,
                   cells = WhichCells(alldata, 
                                      expression = nFeature_RNA>2000 | 
                                        orig.ident != "G_10W_STR"))
  
  
  suppressWarnings(suppressMessages(alldata <- FindVariableFeatures(alldata, selection.method = "vst", 
    nfeatures = 2000, verbose = FALSE, assay = "RNA")))

  
  alldata <- ScaleData(alldata, vars.to.regress = c("percent_mito", "nFeature_RNA"), 
    assay = "RNA")

  alldata <- RunPCA(alldata, npcs = 50, verbose = F)
  
  alldata <- RunUMAP(alldata, reduction = "pca", dims = 1:30, n.components = 2, n.neighbors = 30, 
    n.epochs = 200, min.dist = 0.3, learning.rate = 1, spread = 1)
  if(file.exists(owncache)){
    save(alldata, file = paste0(owncache,"first_dr.RData"))
  }
}else{
  load(paste0(owncache,"first_dr.RData"))
}

```

# Plot QC vars

Number of features and counts and percentage of mitochondrial and ribosomal RNA are plotted below.

```{r qc, fig.height=4, fig.width=6}

VlnPlot(alldata, group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) &
  violintheme

```

# Find variable genes

```{r topgenes, fig.height = 3, fig.width = 5}

top20 <- head(VariableFeatures(alldata), 20)
lp = LabelPoints(plot = VariableFeaturePlot(alldata), points = top20, repel = TRUE) +
  theme(axis.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        axis.text = element_text(size = 8))

lp$layers[[2]]$aes_params$size = 3

lp

```

# PCA

```{r, fig.height = 3, fig.width = 6}


plot_grid(ncol = 3, 
          DimPlot(alldata, reduction = "pca", group.by = "orig.ident", 
                  dims = 1:2, shuffle = TRUE) +
            NoLegend() + NoAxes() + dimtheme,
          DimPlot(alldata, reduction = "pca", group.by = "orig.ident", 
                  dims = 3:4, shuffle = TRUE) + 
            NoLegend() + NoAxes() + dimtheme,
          DimPlot(alldata, reduction = "pca", group.by = "orig.ident",
                  dims = 5:6, shuffle = TRUE) + 
            NoLegend() + NoAxes() + dimtheme)

```

```{r, fig.height = 3, fig.width = 6}

VizDimLoadings(alldata, dims = 1:5, reduction = "pca", ncol = 5, balanced = T) &
  theme(axis.text = element_text(size = 8), axis.title = element_text(size = 8))

```

```{r, fig.height = 2, fig.width = 4}

ElbowPlot(alldata, reduction = "pca", ndims = 50) + 
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 8))

```

# UMAP

```{r firstumap, fig.height = 3, fig.width = 6}

plot_grid(ncol = 2, DimPlot(alldata, reduction = "pca", raster = TRUE,
                            group.by = "orig.ident", shuffle = TRUE) + 
            NoLegend() + NoAxes() + dimtheme, 
    DimPlot(alldata, reduction = "umap", group.by = "orig.ident", raster = TRUE,
            shuffle = TRUE) +
      NoLegend() + NoAxes() + dimtheme)

plot_grid(ncol = 2, DimPlot(alldata, reduction = "umap", raster = TRUE,
                            group.by = "orig.ident", shuffle = TRUE) +
            theme(legend.position = "bottom") + NoAxes() + dimtheme, 
          DimPlot(alldata, reduction = "umap", raster = TRUE,
                  group.by = "Genotype", shuffle = TRUE) +
            theme(legend.position = "bottom") + NoAxes() + dimtheme)

plot_grid(ncol = 2, DimPlot(alldata, reduction = "umap", raster = TRUE,
                  group.by = "RPCA_cluster", label = TRUE, shuffle = TRUE) + 
            NoLegend() + NoAxes() + dimtheme)

FeaturePlot(alldata, reduction = "umap", features = c("Ptprc","Epcam"), raster = TRUE,
    order = T, slot = "data", combine = T) & dimtheme & NoAxes() & NoLegend()

```

# Integration

```{r}

if(!file.exists(paste0(owncache,"integration.RData"))){
  if(!file.exists(paste0(owncache,"anchors.RData"))){
    alldata <- SplitObject(alldata, split.by = "orig.ident")
  
    gc()
    
    for(i in 1:length(alldata)){
      alldata[[i]] = NormalizeData(alldata[[i]])
      alldata[[i]] <- FindVariableFeatures(alldata[[i]], selection.method = "vst",
            nfeatures = 2000)
    }
    
    data.anchors <- FindIntegrationAnchors(
      object.list = alldata, dims = 1:30, 
        reduction = "cca")
    rm(alldata)
    gc()
    
    if(file.exists(owncache)){
      save(data.anchors, file = paste0(owncache,"anchors.RData"))
    }
  }else{
    rm(alldata)
    gc()
    load(paste0(owncache,"anchors.RData"))
  }
  
  alldata <- IntegrateData(anchorset = data.anchors, dims = 1:30,
                                    new.assay.name = "CCA")
  
  rm(data.anchors)
  gc()
  
  alldata <- ScaleData(alldata)
  alldata <- RunPCA(alldata, npcs = 30)
  alldata <- RunUMAP(alldata, dims = 1:30, n.neighbors = 50)
  
  if(file.exists(owncache)){
    save(alldata, file = paste0(owncache,"integration.RData"))
  }
}else{
  load(paste0(owncache,"integration.RData"))
}

```

```{r, fig.height = 3, fig.width=6}

plot_grid(ncol = 2,
  
  DimPlot(alldata, reduction = "pca", group.by = "orig.ident", pt.size = 0.01, shuffle = TRUE) +
    NoAxes() + ggtitle("PCA integrated") + NoLegend() + dimtheme,
  DimPlot(alldata, reduction = "umap", group.by = "orig.ident", pt.size = 0.01, shuffle = TRUE) +
    NoAxes() + ggtitle("UMAP integrated") + NoLegend() + dimtheme
)

alldata@active.assay = "RNA"
FeaturePlot(alldata, reduction = "umap", features = c("Ptprc","Epcam"), 
    order = T, slot = "data", combine = T) & dimtheme & NoAxes()
alldata@active.assay = "CCA"

```


```{r, fig.height = 6, fig.width=4}

FeaturePlot(alldata, reduction = "umap", 
            features = c("percent_mito","percent_ribo","S.Score",
                         "G2M.Score", "nFeature_RNA"), 
    order = T, slot = "data", combine = T, ncol = 2) &
  NoAxes() & NoLegend() & dimtheme

```

```{r}

if(!file.exists(paste0(owncache, "clustered.RData"))){
  alldata@active.assay <- "CCA"
  
  alldata@active.assay

  alldata <- FindNeighbors(alldata, dims = 1:30, k.param = 350, prune.SNN = 1/15)
  names(alldata@graphs)
  
  # Clustering with louvain (algorithm 1)
  for (res in c(0.25, 0.5, 1, 1.5)) {
      alldata <- FindClusters(alldata, graph.name = "CCA_snn",
                              resolution = res, algorithm = 1)
  }
  if(file.exists(owncache)){
    save(alldata, file = paste0(owncache, "clustered.RData"))
  }
}else{
  load(paste0(owncache, "clustered.RData"))
}

```


```{r dimplots, fig.height=6, fig.width = 6}

plot_grid(ncol = 2, 
          DimPlot(alldata, reduction = "umap", group.by = "CCA_snn_res.0.25",
                  label = TRUE) + dimtheme + NoLegend() + NoAxes() + 
            ggtitle("louvain_0.25"),
          DimPlot(alldata, reduction = "umap", group.by = "CCA_snn_res.0.5", 
                  label = TRUE) + dimtheme +  NoLegend() + NoAxes() + 
            ggtitle("louvain_0.5"),
          DimPlot(alldata, reduction = "umap", group.by = "CCA_snn_res.1", 
                  label = TRUE) + dimtheme + NoLegend() + NoAxes() + 
            ggtitle("louvain_1"), 
          DimPlot(alldata, reduction = "umap", group.by = "CCA_snn_res.1.5", 
                  label = TRUE) + dimtheme + NoLegend() + NoAxes() + 
            ggtitle("louvain_1.5"))


```

```{r clustreeplot}

clustree(alldata@meta.data, prefix = "CCA_snn_res.")


```

```{r}

clusternames = 
  data.frame(Cluster = c(0:13),
             Clustername = c("FiB Igfbp3","FiB Has1","MyoC",
                             "Trophocytes","Telocytes","FiB Igfbp6",
                             "Endothelial","Pericytes","ICC",
                             "MyoC Id4","Neurons","VSMC",
                             "Immune","ImmunoFiB"))


clusterorder = c("FiB Has1","FiB Igfbp3","FiB Igfbp6","ImmunoFiB",
                 "Trophocytes","Telocytes","MyoC","MyoC Id4","Endothelial",
                 "Pericytes","VSMC","Neurons","ICC","Immune")


alldata = AddMetaData(alldata, 
                      factor(clusternames$Clustername[
                        match(alldata$CCA_snn_res.0.5,clusternames$Cluster)], 
                        levels = clusterorder),
                      "NamedCluster")

alldata = SetIdent(alldata, value = "NamedCluster")

color_list = get_color_list(14)
names(color_list) = clusterorder

```

```{r, fig.height = 4, fig.width = 4}

dp = DimPlot(alldata, label = TRUE, shuffle = TRUE) + 
  NoAxes() + NoLegend() + dimtheme + scale_color_manual(values = color_list)

dp$layers[[2]]$aes_params$size = 3

dp

```

```{r, fig.width = 6, fig.height = 8}

VlnPlot(alldata, features = c("nFeature_RNA", "nCount_RNA", "percent_mito", "percent_ribo", 
                              "S.Score", "G2M.Score"), 
    ncol = 2, pt.size = 0) & scale_fill_manual(values = color_list) & violintheme

```


```{r,  fig.height = 3, fig.width=2}

if(!file.exists(paste0(owncache, "clustermarkers.rds"))){

  markers_genes <- FindAllMarkers(alldata, logfc.threshold = 0.2, test.use = "wilcox", 
  min.pct = 0.1, min.diff.pct = 0.2, only.pos = TRUE, max.cells.per.ident = 1000, 
  assay = "RNA", random.seed = 42)
  
  if(file.exists(owncache)){
    saveRDS(markers_genes, paste0(owncache, "clustermarkers.rds"))
  }
}else{
  
  markers_genes = readRDS(paste0(owncache, "clustermarkers.rds"))
  
}

top25 <- markers_genes %>% group_by(cluster) %>% top_n(-25, p_val)

out = lapply(unique(top25$cluster), function(i){
  markersdata = top25[top25$cluster==i,]
  markersdata$gene = factor(markersdata$gene, levels = rev(markersdata$gene))
  print(ggplot(markersdata, aes(x = avg_log2FC, y = gene)) + 
          labs(title = i) + 
    geom_bar(stat = "identity") + 
    theme(axis.title.y = element_blank(),
          axis.title.x = element_text(size = 8),
          plot.title = element_text(size = 8),
          axis.text = element_text(size = 8)))
  
    # barplot(sort(setNames(top25$avg_log2FC, top25$gene)[top25$cluster == i], F), horiz = T, 
    #     las = 1, main = paste0(i, " vs. rest"), border = "white", yaxs = "i")
    # abline(v = c(0, 0.2), lty = c(1, 2))
})

```

```{r}

markergenes = 
  c("Cd34","Col1a1","Grem1","Has1","Pdpn","Igfbp3","Adamdec1",
    "Col6a5","Wnt4","Igfbp6","Gas1","Clu","Ccl19","Cxcl13",
    "Pdgfra","Ptn","Fibin","Wnt5a","Bmp5","Foxl1","Myh11","Acta2",
    "Myl9","Hhip","Actc1","Id4","Plvap","Cdh5","Cd81","Cxcl12",
    "Rgs5","Pdgfrb","Pln","Ccl21a","Lyve1","Prox1","Nog","Pcp4",
    "Nrtn","Kit","Etv1","Ano1","Ptprc")

```




```{r, fig.height = 6, fig.width = 4}

dot = DotPlot(alldata, features=rev(markergenes), assay = "RNA")
dot = dot + aes(x = id, y = features.plot) + scale_x_discrete(position = "top") + 
  labs(y = "Genes",x = "Clusters") + 
  scale_size(range = c(0.5,2.5), breaks = c(0,50,100), limits = c(0,100)) +
  theme(legend.position = "none", legend.title = element_blank(),
        axis.text.x = element_text(size = 8, angle = -45, hjust = 1),
        axis.text.y = element_text(size = 8, face = "italic"),
        legend.key.width = unit(0.3, "cm"), 
        legend.key.height = unit(0.3,"cm"), 
        legend.text = element_text(size = 8),
        axis.title = element_text(size = 8))
dot$scales$scales[[1]]$limits = c(-1.8,2.7)

plot_grid(dot,get_legend(dot + theme(legend.position = "right")),
          rel_widths = c(5,1))

```


```{r}


freqs = table(alldata$orig.ident, alldata$NamedCluster)

df = as.data.frame(apply(freqs,1,function(x){x/sum(x)}))
df$Cluster = rownames(df)

fulldf = melt(df)
fulldf$variable = as.numeric(fulldf$variable)

fulldf = fulldf[!is.nan(fulldf$value),]


```



```{r}


frequencies = merge(melt(table(data.frame(cluster=alldata$NamedCluster,
                 sample = alldata$orig.ident)), value.name = "Freq"),
                 melt(apply(table(data.frame(cluster=alldata$NamedCluster,
                                             sample = alldata$orig.ident)),2,
                            function(x){x/sum(x)}), value.name = "Proportion"))

frequencies$time = as.numeric(gsub("W","",alldata$Week[match(frequencies$sample, alldata$orig.ident)]))
frequencies$genotype = factor(alldata$Genotype[match(frequencies$sample, alldata$orig.ident)],
                              levels = c("WT","Il10KO"))

frequencies = frequencies[order(frequencies$cluster),]

cumulFreq = frequencies
cumulFreq$cumulProp = NA
for(i in unique(cumulFreq$sample)){
  cumulFreq$cumulProp[cumulFreq$sample %in% i] = rev(cumsum(rev(cumulFreq$Proportion[cumulFreq$sample %in% i])))
}
cumulFreq = rbind(cumulFreq,cumulFreq)
cumulFreq$xadj = rep(c(-1,1), c(nrow(frequencies),nrow(frequencies)))


```

# Ordered by increase over time


```{r, fig.height = 7}

corlist = lapply(unique(frequencies$cluster), function(i){
  clufreq = frequencies[frequencies$cluster==i  & 
                        frequencies$genotype == "Il10KO",]

  clucor =  cor(clufreq$Proportion,as.numeric((clufreq$time)))
  clucor
  })
names(corlist) = unique(frequencies$cluster)

averagecor = unlist(corlist)

frequencies$cluster = 
  factor(frequencies$cluster, 
         levels = names(averagecor)[order(unlist(averagecor), decreasing = TRUE)])
cumulFreq$cluster = 
  factor(cumulFreq$cluster,
         levels = names(averagecor)[order(unlist(averagecor), decreasing = TRUE)])

frequencies$sample = factor(frequencies$sample, 
                          levels = 
                            unique(frequencies$sample[order(frequencies$genotype,
                                                   frequencies$time)]))
frequencies$samplenumber = as.numeric(frequencies$sample)


cumulFreq$sample = factor(cumulFreq$sample, 
                          levels = 
                            levels(frequencies$sample))
cumulFreq$samplenumber = as.numeric(cumulFreq$sample)


```


```{r, fig.height = 3.5, fig.width = 3.5}


datasets = lapply(unique(frequencies$genotype), function(bg){
  bgfreq = frequencies[frequencies$genotype == bg,]
  
  cumbgfr = cumulFreq[paste(cumulFreq$sample,
                                   cumulFreq$oldtime) %in% 
                               paste(bgfreq$sample,
                                     bgfreq$oldtime),]
  list(bgfreq, cumbgfr)
})


barplot = ggplot(data = frequencies, 
       aes(x = samplenumber, y = Proportion, fill = cluster)) + 
  theme(panel.background = element_blank(), axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
            scale_fill_manual(values =
                                color_list[as.character(unique(sort(frequencies$cluster)))]) + 
            scale_x_continuous(breaks = sort(unique(frequencies$samplenumber)),
                               labels = 
                                 gsub("BALBc","WT",
                                      gsub("_IMM","",
                                           gsub("G_","", 
                                                gsub("IL10_","",
                                                     levels(frequencies$sample))))))

for(ds in datasets){
  barplot = barplot + 
    geom_area(data = ds[[2]],
              alpha = 0.7, linewidth = 0.3, stat = 'identity',
              aes(fill = cluster, group = cluster,
                  y = Proportion, x = samplenumber + xadj*0.3))
}

barplot + geom_bar(stat = "identity", width = 0.6, color = "black") + 
  theme(legend.key.size = unit(8,"pt"))


```

# Subset analyses


```{r}

alldata = AddMetaData(alldata, alldata$NamedCluster, "fibroblasts")

alldata$fibroblasts[!(alldata$fibroblasts %in% 
                  c("FiB Has1","FiB Igfbp3","FiB Igfbp6","ImmunoFiB",
                 "Trophocytes","Telocytes"))] = NA


```

```{r, fig.height = 3, fig.width = 3}

plot_grid(plotlist = lapply(c("fibroblasts"),
       function(x){
         dp = DimPlot(alldata, group.by = x, label = TRUE) +
            scale_color_manual(values = color_list, na.value = "#CACACA") +
            dimtheme & NoLegend() & NoAxes()
         dp$layers[[2]]$aes_params$size = 3
         dp
       }))

```

```{r}

fibroblasts = subset(alldata, cells = WhichCells(alldata, expression = fibroblasts %in% alldata$NamedCluster))

rm(alldata)
gc()

```


## Fibroblasts



```{r}

ntop = 40
downsample_n = 1000
txt = element_text(size = 8)

```


```{r}

fibroblasts@active.assay <- "RNA"


fibroblasts <- FindVariableFeatures(fibroblasts)
fibroblasts <- ScaleData(fibroblasts, features = VariableFeatures(fibroblasts))

fibroblasts <- RunNMF(object = fibroblasts, assay = "RNA", nfactors = ntop,
              n.cores = 1)
 

for(i in 1:ntop){
  fibroblasts <- AddMetaData(fibroblasts, fibroblasts@reductions$NMF@cell.embeddings[,paste0("factor_",i)],
                     paste0("nmf_",i))
}

fibroblasts$Week = as.numeric(gsub("W","",fibroblasts$Week))

fibroblasts = AddMetaData(fibroblasts, factor(paste(fibroblasts$Genotype, fibroblasts$Week),
                              levels = c("WT 6", 
                                         paste0("Il10KO ", 
                                               c(6,8,10,12,14)))), 
                  "G_W")

```

```{r}

pvaluelist = getfactorpvalues_Il10(fibroblasts, downsample_n, ntop)
modelpvalues = pvaluelist$modelpvalues
rank_modelpvalues = pvaluelist$rank_modelpvalues
rm(pvaluelist)


pvaluegg = melt(modelpvalues, variable.name = "factor",
                value.name =  "p.value")
pvaluegg$depvar = factor(pvaluegg$depvar, levels = unique(pvaluegg$depvar))


fibroblasts_values = fibroblasts@meta.data[,grep("^nmf_", colnames(fibroblasts@meta.data))]

fibroblasts_scaled = apply(fibroblasts_values, 2, scale)

fibroblasts_scaled_means_id = aggregate(fibroblasts_scaled, by = list(fibroblasts$G_W), mean)
fibroblasts_means_id = aggregate(fibroblasts_values, by = list(fibroblasts$G_W), mean)
fibroblasts_means_scaled_id = as.data.frame(apply(fibroblasts_means_id[,-1], 2, scale))
fibroblasts_means_scaled_id$Group.1 = fibroblasts_means_id$Group.1

fibroblasts_scaled_means_cluster = aggregate(fibroblasts_scaled, by = list(fibroblasts$NamedCluster), mean)
fibroblasts_means_cluster = aggregate(fibroblasts_values, by = list(fibroblasts$NamedCluster), mean)
if(nrow(fibroblasts_means_cluster)==1){
  fibroblasts_means_scaled_cluster = as.data.frame(rbind(apply(fibroblasts_means_cluster[,-1], 2, scale)))
  fibroblasts_means_scaled_cluster$Group.1 = fibroblasts_means_cluster$Group.1
  fibroblasts_means_scaled_cluster[1,is.nan(as.numeric(fibroblasts_means_scaled_cluster[1,]))] = 0
}else{
  fibroblasts_means_scaled_cluster = as.data.frame(apply(fibroblasts_means_cluster[,-1], 2, scale))
  fibroblasts_means_scaled_cluster$Group.1 = fibroblasts_means_cluster$Group.1
}

```

```{r, fig.show="hide"}

ph_sample = pheatmap(fibroblasts_means_scaled_id[,paste0("nmf_", 1:ntop)])

ph_order = ph_sample$tree_col$order

```

```{r, fig.width = 2.5, fig.height = 6}

my_dotp = ggplot(melt(fibroblasts_means_scaled_id), 
                 aes(x = Group.1, 
                     y = factor(variable, 
                                levels = rev(paste0("nmf_",ph_order))))) +
  geom_tile(aes(fill = value)) + 
  geom_vline(color = "white", xintercept = 1.5) +
  scale_fill_viridis_c() + 
  scale_y_discrete(labels = function(x){gsub("nmf_", "f",x)}) + 
  theme_classic() + 
  theme(axis.text = txt,
        axis.title = element_blank(),
        plot.title = element_blank(),
        legend.text = txt,
        legend.title = txt,
        axis.text.x = element_text(size = 8, angle = 90, hjust = 1, vjust = 0.5)) + 
  scale_size_continuous(limits = c(0,100), range= c(0,3))

my_dotp_cl = ggplot(melt(fibroblasts_means_scaled_cluster), 
                 aes(x = Group.1, 
                     y = factor(variable, 
                                levels = rev(paste0("nmf_",ph_order))))) +
  geom_tile(aes(fill = value)) + scale_fill_viridis_c() +
  theme_classic() +
  theme(axis.text = txt,
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.title = element_blank(),
        legend.text = txt,
        axis.text.y = element_blank(),
        legend.key.size = unit(8,"pt"),
        plot.title = element_blank(),
        legend.position = "bottom",
        legend.title = txt) + 
  scale_size_continuous(limits = c(0,100))

lmp = ggplot(pvaluegg,
       aes(y = factor(factor, levels = rev(paste0("nmf_",ph_order))), x = depvar)) + 
  geom_tile(aes(fill = -log10(p.value))) + 
  scale_fill_viridis_c(option = "inferno", na.value = "#FCFFA4FF", limits = c(0,NA)) + 
  scale_x_discrete(labels = function(x){gsub("SCID","",x)}) + 
  theme_classic() + 
  theme(axis.text.x = element_text(size = 8, angle = 90, hjust = 1, vjust = 0.5),
        axis.text.y = element_blank(),
        axis.title = element_blank(),
        legend.title = element_text(size = 8),
        legend.position = "bottom",
        legend.direction = "vertical",
        legend.text = element_text(size = 8),
        legend.key.size = unit(8,"pt"))

scalerange = c(min(my_dotp$data$value, my_dotp_cl$data$value),
               max(my_dotp$data$value, my_dotp_cl$data$value))

plot_grid(my_dotp + 
            scale_fill_viridis_c(limits = scalerange) + 
            theme(axis.title = element_blank(),
                  legend.position = "bottom",
                      legend.key.size = unit(8,"pt")),
          my_dotp_cl + 
            scale_fill_viridis_c(limits = scalerange),
          lmp + 
            labs(fill = ""), 
          align = "h", ncol = 3, rel_widths = c(3.5,2.7,1)
          )

```

```{r, fig.height = 2, fig.width = 4}

showfactors_Il10(fibroblasts)

```


# Session info

```{r}

sessionInfo()

```
