---
title: "Analysis of bulk RNA seq in Il10-/- colitis"
author: "Jennifer Fransson"
editor_options:
  chunk_output_type: inline
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    code_folding: hide
    toc: yes
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
params:
  dataset: 1
---

```{r Setup, echo = FALSE}
knitr::opts_chunk$set(results    = "hold",
                      message    = FALSE,
                      warning    = FALSE)

if(!knitr::is_html_output()){
  my_plot_hook <- function(x, options)
   paste("\n", knitr::hook_plot_tex(x, options), "\n")
  knitr::knit_hooks$set(plot = my_plot_hook)
}
```



# Introduction

This script describes the analyses performed to describe the disease process of murine Il10-/- colitis for the study "The Spatiotemporal Transcriptomic Landscape in Murine Chronic Colitis". The data analyzed are bulk RNA-sequenced biopsies of distal colon. Samples were collected from Il10-/- mice at 5 time points (weeks 6, 8, 10, 12 and 14) and WT mice at week 6. 

# Load libraries and other scripts

In addition to the packages loaded, an extra script (bulk_functions.R) is sourced prior to analysis. This script contains functions and parameters for visualization that are used for this analysis as well as the analysis of T cells transfer bulk RNAseq.

```{r}

library(ggplot2)
library(cowplot)
library(grid)
library(mgcv)
library(rafalib)
library(edgeR)
library(reshape2)
library(enrichR)
library(ggvenn)
library(data.table)
library(ggpubr)
library(progeny)
library(pheatmap)


source("../code/Bulk/bulk_functions.R")
datadir = "../data/Bulk/Il10KO/"

```

# Read data files

The gene expression matrices are available from GEO. In order to generate them, raw sequences were pseudo-aligned and quantified using kallisto. The generated abundance-files were renamed according to the samples and concatenated using the code below:

```{r, eval = FALSE}

rawdatadir = "../data/Bulk/Il10KO/kallisto/")

### Generation of counts and TPM tables -
####### Final files can be downloaded from GEO, but the code below describes
####### how they were generated from the kallisto output files


filenames <- list.files(rawdatadir,"Dist.*\\.tsv", full.names = TRUE)

# Compile count and TPM tables from kallisto counts

raw_counts <- sapply(filenames,function(x){
  read.delim(x)[,"est_counts"] })

colnames(raw_counts) <- sub(".*/","",sub(".tsv","",colnames(raw_counts)))
rownames(raw_counts) <- read.delim(filenames[1])[,"target_id"]


raw_tpm <- sapply(filenames,function(x){
  read.delim(x)[,"tpm"] })
colnames(raw_tpm) <- sub(".*/","",sub(".tsv","",colnames(raw_tpm)))
rownames(raw_tpm) <- read.delim(paste0("../raw_data/",filenames[1]))[,"target_id"]


# Load metadata and match to counts data

metadata <- read.csv2(paste0(rawdatadir, "/raw_metadata.csv"), row.names = 1 )
common <- rownames(metadata)[ rownames(metadata) %in% colnames(raw_counts)

metadata <- metadata[common, ]
raw_counts <- round(raw_counts[,common],0)
raw_tpm <- raw_tpm[,common]

# Translate to gene names
# Next, we can summarize ensembl transcript IDs to gene symbols for easier data interpretation. Here, we will download the GTF file from ensemble used for the annotation in the ammping step from Kallisto and annotate accordingly.

gtf <- getgtf(datadir)


annot <- gtf$gene_name [ match( sub("[.].*","",rownames(raw_counts)),gtf$transcript_id ) ]
annot[is.na(annot)] <- "NA"

counts <- rowsum(raw_counts , group = annot)
counts <- counts[rownames(counts) != "NA",]

tpm <- rowsum(raw_tpm , group = annot)
tpm <- tpm[rownames(tpm) != "NA",]

# Filter counts and TPM tables

sel <- rowSums( counts > 10 ) >= 3

filt_counts <- counts[sel,]
filt_tpm <- tpm[sel,]

# Save filtered tables

write.table(filt_counts, paste0(datadir,"gene_counts.tsv"))
write.table(filt_tpm, paste0(datadir,"gene_tpm.tsv"))

####

```

Once the gene matrices have been created and saved (or downloaded), they can be loaded as follows without repeating the above code.

```{r}

filt_counts = read.table(paste0(datadir,"gene_counts.tsv"))
filt_tpm = read.table(paste0(datadir,"gene_tpm.tsv"))
metadata <- read.csv2(paste0(datadir,"metadata.csv"), row.names = 1 )

filt_counts = filt_counts[,rownames(metadata)]
filt_tpm = filt_tpm[,rownames(metadata)]

colnames(metadata) = gsub("X","",gsub("\\.", "", colnames(metadata)))
metadata$time = factor(metadata$time, levels = paste0(seq(6,14,2), "w"))
metadata$sample = rownames(metadata)
metadata$genotype = factor(metadata$genotype, levels = c("BALB/c WT", "BALB/c Il10KO"),
                           labels = c("WT", "Il10KO"))

```


# Quality control

Homogeneity in sample quality is assessed through number of counts, number of features and % ribosomal RNA. Integrity values (RIN) provided by Novogene UK (provider of cDNA library preparation and sequencing) have also been included. 

```{r}

metadata$nFeatures <- colSums(filt_counts>0)

metadata$nCounts <- colSums(filt_counts)

is_mito <- grepl("mt-",rownames(filt_counts))

metadata$perc_mito <- colSums(filt_counts[is_mito,]) / colSums(filt_counts) * 100

is_ribo <- grepl("Rp[ls]",rownames(filt_counts))

metadata$perc_ribo <- colSums(filt_counts[is_ribo,]) / colSums(filt_counts) * 100

```



```{r, fig.height = 6, fig.width=4}

x <- c("nFeatures","nCounts","perc_ribo","perc_mito","RIN")

glist = list()
for(i in x){
  for(g in c("time", "cage","genotype")){
    imean = aggregate(metadata[,i],by=list(metadata[,g]),mean)
    imd = metadata
    imd = imd[order(imd[,g]),]
    imd$sample = factor(imd$sample, levels = imd$sample)
    imd$mean = imean$x[match(imd[,g], imean$Group.1)]
    imeanx = aggregate(as.numeric(imd$sample), by = list(imd[,g]), mean)
    glist[[g]][[i]] = ggplot(imd, aes_string(y = i, x = "as.numeric(sample)")) + 
      geom_bar(aes_string(fill = g), stat = "identity", position = "dodge", show.legend = FALSE, alpha = 0.8) +
      theme(panel.background = element_blank(),
            axis.text.x = element_text(angle = 30, hjust = 1, size = 10),
            axis.title.x = element_blank()) + 
      geom_line(aes_string(y = "mean", group = g)) + 
      scale_x_continuous(breaks = imeanx[,2], labels = imeanx[,1])
  }
  
}
plot_grid(textGrob("time", hjust = 0.5, gp=gpar(fontsize=16)),
          plot_grid(plotlist = glist[["time"]], ncol = 1, align = "v"),ncol = 1, rel_heights = c(0.05,1))
plot_grid(textGrob("cage", hjust = 0.5, gp=gpar(fontsize=16)),
          plot_grid(plotlist = glist[["cage"]], ncol = 1, align = "v"),ncol = 1, rel_heights = c(0.05,1))
plot_grid(textGrob("genotype", hjust = 0.5, gp=gpar(fontsize=16)),
          plot_grid(plotlist = glist[["genotype"]], ncol = 1, align = "v"),ncol = 1, rel_heights = c(0.05,1))

```


# Principal component analysis

In order to visualize the overall variation between samples, we perform a principal component analysis (PCA) of all samples, using the top 500 most highly variable genes across samples. The relation between mean expression and standard deviation is shown below (each dot is one gene), indicating the 500 genes used for PCA in red.

```{r, fig.height = 3, fig.width = 3}

log_tpm <- log2(filt_tpm+1)
log_sd <- apply(log_tpm,1,sd)
log_means <- apply(log_tpm,1,mean)
HVGs <- names(sort(log_sd,T))[1:500] # highly variable genes
plot(log_means,log_sd,
     col=ifelse(names(log_sd) %in% HVGs,"red","black" ),
     cex=.5,pch=16)
lines(smooth.spline(log_means,log_sd,spar = 1,nknots = 50),col="blue",lwd=3)

```

The PCA shows that >60% of the total variance between samples can be explained by PC1 (shown below).

```{r, fig.height = 2, fig.width = 3}

PC <- prcomp( t(log_tpm[HVGs, ncol(log_tpm):1]) ,center = TRUE, scale. = T )
PC$rotation[,1] = 0-PC$rotation[,1]
PC$x[,1] = 0-PC$x[,1]

PCvar <- PC$sdev^2 / sum(PC$sdev^2) * 100
names(PCvar) <- 1:length(PCvar)
pcdata <- cbind(PC$x, metadata[nrow(metadata):1,])

# Elbow plot
ggplot(data.frame(variance = PCvar, PC = as.numeric(names(PCvar))), aes(x = PC, y = variance)) + 
  geom_point() + geom_line() +
  theme(panel.background = element_blank()) +
  labs(y = "% Variance explained")

```

The first two components of the PCA are visualized below (left). Each point is one sample. The points are colored according to time points, and shape indicates genotype. The correlation between PC1 and histological score is also visualized (right). Points are colored according to lipocalin levels (log10(ng/g of feces)), and shape indicates genotype.

```{r, fig.height = 1.5, fig.width = 6}


cor_histo = cor(pcdata$PC1, pcdata$histologicalscore)^2
cor_lcn = cor(pcdata$PC1, log10(pcdata$lipocalinngg), use = "pairwise.co")^2


g1 = ggplot(pcdata, aes(x = PC1,y = PC2) ) +
  geom_point(size = pointsize, aes(shape = genotype, color = time), 
             stroke = strokewidth) + 
  
  scale_color_viridis_d(option = viridistype, direction = viridisdirection)  +
   guides(fill = guide_legend(override.aes = list(shape = 22, color=NA) ) ) + plottheme

g2<- ggplot(pcdata, aes(x = PC1,y = histologicalscore)) + 
  geom_point(aes(color = log10(lipocalinngg), shape = genotype),
             size =pointsize, stroke = strokewidth) +
  theme_classic() + theme(legend.position = "bottom") + plottheme + 
  theme(axis.ticks.y = element_line(color = "black", linewidth = 0.25))

plot_grid(g1 + theme(legend.position = "right",
                     legend.key.size = unit(8,"pt")), 
          g2 + theme(legend.position = "right",
                     legend.key.size = unit(8,"pt"),
                     axis.text.y = element_text(size = 8)) + 
            labs(y = "Histological score"),
          ncol = 2)

```

The genes that contribute the most (positively and negatively) to PC1 and PC2 are shown below, with loading to the PC indicated on the x-axis.

```{r, fig.width = 4, fig.height = 2.5}

up1 = sort(PC$rotation[,"PC1"], decreasing = TRUE)[1:20]
down1 = sort(PC$rotation[,"PC1"], decreasing = FALSE)[1:20]

up2 = sort(PC$rotation[,"PC2"], decreasing = TRUE)[1:20]
down2 = sort(PC$rotation[,"PC2"], decreasing = FALSE)[1:20]

ngenes = 20

loadings1 = ggplot(data.frame(x = c(ngenes:1,ngenes:1), loading = c(up1[1:ngenes], down1[1:ngenes]),
           gene = c(names(up1[1:ngenes]), names(down1[1:ngenes])),
           dir = rep(c("up","down"),each = ngenes)),
       aes(x = x, y = loading)) +
  geom_bar(stat = "identity", aes(fill = dir)) + coord_flip() +
  scale_fill_manual(values= c(up = "#FFBBBB", down = "#BBBBFF")) +
  geom_text(aes(label = gene), hjust = rep(c(0,1),each = ngenes), size = 2) +
  labs(title = "PC1") + ylim(-0.1,0.1) +
  bartheme


loadings2 = ggplot(data.frame(x = c(ngenes:1,ngenes:1), loading = c(up2[1:ngenes], down2[1:ngenes]),
           gene = c(names(up2[1:ngenes]), names(down2[1:ngenes])),
           dir = rep(c("up","down"),each = ngenes)),
       aes(x = x, y = loading)) +
  geom_bar(stat = "identity", aes(fill = dir)) + coord_flip() +
  scale_fill_manual(values= c(up = "#FFBBBB", down = "#BBBBFF")) +
  geom_text(aes(label = gene), hjust = rep(c(0,1),each = ngenes), size = 2) +
  labs(title = "PC2") + ylim(-0.2,0.2) +
  bartheme

plot_grid(loadings1, loadings2, ncol = 2)

```


# Genes of interest

The expression of certain genes of interest for the study are shown below. The gene names in _genes_ can be replaced with any gene present in the filtered dataset. Values are shown as log2(tpm+1).

```{r, fig.height = 3.75, fig.width = 6}

genes = c("Cd3g","Gzmk","Cd14", "Il1b","S100a9",
          "Mcpt1","Xcl1","Klrc1",
          "Il7r", "Icos","Nrgn","Xcr1","Siglech",
          "Muc2", "Muc3","Chga")

genedf = cbind(metadata, t(log_tpm[genes,]))

genedf$time <- factor(as.numeric(gsub("w","",genedf$time)))

plottheme =   theme(legend.position = "none",
        axis.text = element_text(size = 8),
        axis.title = element_text(size = 8),
        panel.background = element_blank(),
        panel.grid = element_blank())

means = aggregate(genedf[,genes], by = list(genedf$genotype, genedf$time), mean)
colnames(means)[1:2] = c("genotype", "time")

ymaxes = c(1,2,5,10,20)
ybreaks = list()
ybreaks[[1]] = c(0,1)
ybreaks[[2]] = c(0,1,2)
ybreaks[[5]] = c(0,2,4)
ybreaks[[10]] = c(0,3,6,9)
ybreaks[[20]] = c(0,5,10,15)


geneplots = lapply(genes, function(gene){
  ymax = ymaxes[ymaxes>max(genedf[,gene])][1]
  ggplot(genedf, aes_string(x = "time", y = gene, color = "genotype")) + 
          geom_line(data = means, aes(group = genotype)) + 
          geom_point(aes(shape = genotype), alpha = 0.5) + 
    labs(title = gene) + 
          plottheme + 
    scale_y_continuous(breaks = ybreaks[[ymax]]) + 
    theme(axis.ticks.x = element_blank(), 
          axis.title = element_blank(),
          plot.title = element_text(size = 10, hjust = 0.5)) + 
          coord_cartesian(ylim=c(0,NA))
})


plot_grid(plotlist = geneplots, ncol = 6)

```

# Differential expression analysis

Differentially expressed genes are identified using edgeR linear modeling between all groups. By defining the timepoint & genotype as one factor, each experimental group is considered as an independent variable in the model.

```{r}

# Fixing time to factor including wt
trmetadata = metadata
trmetadata$time = factor(trmetadata$time, levels = c("WT",levels(trmetadata$time)))
trmetadata$time[trmetadata$genotype %in% "WT"] = "WT"


# Creating the linear model and dataset
y <- DGEList(counts=filt_counts,group=trmetadata$time)
keep <- filterByExpr(y)
y <- y[keep,,keep.lib.sizes=FALSE]
y <- calcNormFactors(y)

time = trmetadata$time
design <- model.matrix(~time)

y <- estimateDisp(y,design)

# Fitting the data to the model

fit <- glmQLFit(y,design)


timelrtfit = glmLRT(fit,coef = 2:ncol(fit)) # (null hypothesis: expression is equal at all time points and genotypes)


# Summarizing WTlrt results and adding adjusted p 

timelrt <- timelrtfit$table
timelrt <- timelrt[ order(timelrt$PValue) , ]
timelrt$FDR <- p.adjust(timelrt$PValue,"BH")


```


## Modules of differentially expressed genes over time in T cell transfer colitis

The results table from the test defined above is used to define the differentially expressed genes across time. All genes with FDR < 0.01 are included in the analysis. LogFC values are normalized to max = 1 and min = 0 for each gene. Hierarchical clustering is performed between genes using these values. The dendrogram is cut to create clusters of co-regulated genes ("modules"). The height at which to cut is selected in order to create cohesive modules with minimal redundancy.

```{r}

# Add the Week 0 baseline (0) to the table
timelrt$logFC.timeWT = 0 

# Rename the columns in order to adhere to format logFC.WX
colnames(timelrt) = gsub("time","",colnames(timelrt))

# Create a table containing only the logFCs for each time point, and only of the genes with FDR<0.01
logFCs = timelrt[timelrt$FDR<0.01,grepl("logFC",colnames(timelrt))]
logFCs = logFCs[,c(ncol(logFCs),1:(ncol(logFCs)-1))]

```


```{r}

# Normalize gene expression to min = 0 and max = 1
logFCs0 = t(apply(logFCs,1,function(x){x-min(x)}))
logFCs01 = t(apply(logFCs0,1,function(x){x/max(x)}))

# Perform hierarchical clustering of DE genes
geneDistance <- as.dist((1-cor(t(logFCs01)))/2)
geneCluster = hclust(geneDistance,method="ward.D2")

```

```{r}

# Choose a height at which to cut the dendrogram of clustered genes
height_cut <- 4

# Cut the dendrogram
groups <- cutree(geneCluster,h=height_cut)


# Reorder the cluster in order of kinetics
genes = names(groups)
groups = match(groups, c(1,2,3,4,5))
names(groups) = genes

mousecolors = gg_color_hue(max(groups))
names(mousecolors) = 1:length(mousecolors)

timelrt$module = groups[rownames(timelrt)]
write.csv2(timelrt, paste0(datadir, "DE_modules.csv"))

```

```{r}

myplclust(geneCluster, labels=rep("-", length(groups)),
          lab.col = mousecolors[groups])
abline( h=height_cut, col="red", lwd=0.5)

```

The numbers of genes per module are shown here:

```{r}

knitr::kable(table(groups), 
             col.names = c("Module","No of genes"), 
             align = "c")

```

For each module, the logFC over time is shown for each gene in the module (black lines). The average logFC at each time point is shown with the colored line.

```{r}

linelist = list()

for(i in sort(unique(groups))){
  clusters_logFCs01 = as.data.frame(logFCs01[groups==i,])
  clusters_logFCs01$Gene = rownames(clusters_logFCs01)
  clf01long = melt(clusters_logFCs01)
  clf01long$Week = factor(gsub("logFC.","",clf01long$variable),
                          levels = c("WT",paste0(seq(6,14,2),"w")))
  linelist[[i]] = ggplot(clf01long, aes(x = Week, y = value)) + 
    geom_line(aes(group = Gene), alpha = 0.4) +  
    labs(title = paste("Module",i)) + 
    stat_summary(geom="line", color = mousecolors[i], 
                 size = 2, fun = mean, group = 1) + 
    theme(plot.title = element_text(size = 10),
          axis.text = element_text(size = 8),
          axis.title = element_text(size = 8))
}

cowplot::plot_grid(plotlist = linelist)

```


### Over-representation analysis

For each module, the list of included genes is analyzed to identify over-represented GO biological process terms and KEGG pathways using Enrichr. The results are shown in the summaries below.

```{r, results=FALSE}

GOBP = list()
KEGG = list()

for(i in 1:max(groups)){
  enres = enrichr(genes = names(groups[groups==i]),
                  databases =c("GO_Biological_Process_2021",
                               "KEGG_2019_Mouse"))
  GOBP[[i]] = enres[[1]]
  KEGG[[i]] = enres[[2]]
}

```



```{r, fig.height = 5.5, fig.width = 4.5}

axis_size = 8

gobplist = lapply(sort(unique(groups)),function(i){
  x = GOBP[[i]]
  sig = x[x$Adjusted.P.value<0.05 & as.numeric(gsub("/.*","",x$Overlap))>=4,c("Term","Overlap","Adjusted.P.value","Genes")]
  if(nrow(sig) == 0){
    return(NULL)
  }else{
  if(nrow(sig)>5){sig = sig[1:3,]}
  sig$Term = gsub("\\(GO.*","",sig$Term)
  sig$Term = shortenterms(sig$Term)
  sig$Term = factor(sig$Term, levels = rev(sig$Term))
  return(ggplot(sig, aes(x = -log10(Adjusted.P.value), y = Term)) + 
           geom_bar(stat="identity", fill = "#DDDDDD") + 
           labs(title = paste("Module",i)) + 
           geom_text(x = 80, aes(label = Term), vjust = 0.5,
                     hjust = 0, size = 3) +
           theme(axis.text.y = element_blank(),
                 axis.title.y = element_blank(),
                 axis.ticks.y = element_blank()) + 
           theme(panel.background = element_blank(),
                 plot.title = element_blank()) + 
           scale_x_continuous(n.breaks = 3, limits = c(0,75), 
         expand = expansion(mult = c(0,0.1), add = 0)) + 
           scale_y_discrete(expand = expansion(mult = 0, add = 0)) + 
           coord_cartesian(clip = "off"))
  }
  })


```


```{r}

KEGGlist = lapply(sort(unique(groups)),function(i){
  x = KEGG[[i]]
  sig = x[x$Adjusted.P.value<0.05 & as.numeric(gsub("/.*","",x$Overlap))>=4,c("Term","Overlap","Adjusted.P.value","Genes")]
  if(nrow(sig) == 0){
    return(NULL)
  }else{
    if(nrow(sig)>5){sig = sig[1:3,]}
    sig$Term = gsub("\\(GO.*","",sig$Term)
    sig$Term = shortenterms(sig$Term)
    sig$Term = factor(sig$Term, levels = rev(sig$Term))
    return(ggplot(sig, aes(x = -log10(Adjusted.P.value), y = Term)) + 
             geom_bar(stat="identity", fill = "#DDDDDD") + 
             labs(title = paste("Module",i)) + 
             geom_text(x = 80, aes(label = Term), size = 3,
                       vjust = 0.5, hjust = 0) +
             theme(axis.text.y = element_blank(),
                   axis.title.y = element_blank(),
                   axis.ticks.y = element_blank()) + 
             theme(panel.background = element_blank(),
                   plot.title = element_blank()) + 
             scale_x_continuous(n.breaks = 3, limits = c(0,75), 
           expand = expansion(mult = c(0,0.1), add = 0)) + 
             scale_y_discrete(expand = expansion(mult = 0, add = 0)) + 
             coord_cartesian(clip = "off"))
    }
  })

```

```{r}

heatmapdf = melt(logFCs01[order(groups),])
colnames(heatmapdf) = c("Gene","Week","logFC")
heatmapdf$Week = gsub("logFC.","",heatmapdf$Week)
heatmapdf$Week = factor(heatmapdf$Week, 
                        levels = c("WT", paste0(seq(6,14,2),"w")))
heatmapdf$Gene = factor(heatmapdf$Gene, #levels = ylevels) 
                        rev(rownames(logFCs01)[order(groups)]))
heatmapdf$Module = factor(groups[as.character(heatmapdf$Gene)])


ggheatmap = ggplot(heatmapdf, aes(y = Gene, x = Week)) + 
  geom_tile(aes(fill = logFC)) +
            theme(panel.background = element_blank(), axis.text.y = element_blank(),
                  axis.title.y = element_blank(), axis.ticks.y = element_blank(),
                 axis.text.x = element_text(size = axis_size)) + 
            scale_fill_gradient2(low = "#00BFC4", mid ="white",
                                 high = "#F8766D", midpoint = 0.5) + 
  labs(fill = "Expression") +
  scale_y_discrete(expand = expansion(mult = 0, add = 0)) +
  scale_x_discrete(expand = expansion(mult = 0, add = 0)) +
  theme(legend.position = "bottom")

freqs = rev(table(heatmapdf[heatmapdf$Week=="WT","Module"])/nrow(heatmapdf[heatmapdf$Week=="WT",]))

modulehm = ggplot(heatmapdf[heatmapdf$Week=="WT",]) + 
  geom_bar(mapping = aes(fill = Module, x = Week), position ="fill") +
  theme(panel.background = element_blank(), 
                  axis.text.y = element_blank(),
                  axis.ticks.y = element_blank(),
                  axis.title.y = element_blank()) +
            labs(x = "Module") + 
            scale_x_discrete(labels = c(""), 
                             expand = expansion(mult = 0, add = 0)) + 
            scale_y_continuous(expand = expansion(mult = 0, add = 0)) + 
  geom_text(data = as.data.frame(cumsum(freqs) - freqs/2), size = 2.7,
                      aes(y = Freq, label = Var1), x = 1)

```

### Summaries

Below, summaries are shown for the defined modules. The summary includes a heatmap showing the smoothed and normalized logFC values, line plots showing the average expression in each module, and bar plots showing the top over-represented GO terms (first figure) and KEGG pathways (second figure).


```{r, fig.height = 5, fig.width = 8}

col_widths = c(1.5,1,0.5,2.5,0.5,2.5)

heatmapblock = plot_grid(
        plot_grid(
          modulehm + theme(legend.position = "none",
                   axis.text.x = element_blank(),
                   axis.ticks.x = element_blank(),
                   axis.title.x = element_blank()) , 
          
          ggheatmap + theme(legend.position = "none",
                          axis.text.x = element_blank(),
                   axis.title.x = element_blank(),
                   axis.ticks.x = element_blank()) ,
          
          NULL,
          
          ggdraw(get_x_axis(ggheatmap, position = "bottom")) ,
          
          ncol = 2, rel_widths = c(1,4),nrow = 2, 
          rel_heights = c(20,1), align = "v"),
         textGrob("Week",gp = gpar(fontsize = 8), hjust = 0.25), ncol = 1, 
          rel_heights = c(41,1)
        )

lineblock = plot_grid(
        plot_grid(plotlist = c(lapply(linelist, function(x){
          x$layers[[1]]$aes_params$size = 0.1
          x$layers[[2]]$aes_params$size = 0.5
          x + theme(panel.background = element_blank(),
                    panel.grid = element_blank(),
                    axis.title = element_blank(),
                    plot.title = element_blank(),
                    axis.text = element_blank(),
                    axis.ticks = element_blank()) +
            scale_y_continuous(expand = expansion(mult = 0, add = 0) ) 
          }),
          
          list(ggdraw(get_x_axis(linelist[[1]] +
            scale_y_continuous(expand = expansion(mult = 0, add = 0) ) +
              theme(axis.text = element_text(size = axis_size)))))), 
          ncol = 1, 
          rel_heights = c(rep(20/length(linelist),length(linelist)),1), 
          align = "v"),
        textGrob("Week",gp = gpar(fontsize = 8)),
        ncol = 1, 
          rel_heights = c(41,1)
      )

KEGGblock = plot_grid(plotlist = c(lapply(KEGGlist, function(x){
          x + theme(panel.background = element_blank(), 
                    panel.grid = element_blank(),
                    axis.title = element_blank(), axis.text = element_blank(),
                    axis.ticks = element_blank())
          }),
          list(ggdraw(get_x_axis(KEGGlist[[1]] + 
                                   theme(axis.text = 
                                           element_text(size = axis_size)))))),
          ncol = 1, 
          rel_heights = c(rep(20/length(KEGGlist),length(KEGGlist)),1),
          align = "v")

GOBPblock = plot_grid(plotlist = c(lapply(gobplist, function(x){
          x + theme(panel.background = element_blank(), 
                    panel.grid = element_blank(),
                    axis.title = element_blank(), axis.text = element_blank(),
                    axis.ticks = element_blank())
          }),
          list(ggdraw(get_x_axis(gobplist[[1]] + 
                                   theme(axis.text = 
                                           element_text(size = axis_size)))))),
          ncol = 1, 
          rel_heights = c(rep(20/length(gobplist),length(gobplist)),1),
          align = "v")



  plot_grid(
    plot_grid(textGrob("logFCs",gp = gpar(fontsize = 10), hjust = 0.25),
              textGrob("Average logFCs",gp = gpar(fontsize = 10), hjust = 0.5),
              textGrob("Top GO BP terms",gp = gpar(fontsize = 10), hjust = 0),
              NULL,
              textGrob("Top KEGG pathways",gp = gpar(fontsize = 10), hjust = 0),
              NULL,
              ncol = 6, rel_widths = col_widths),
    plot_grid(
      heatmapblock,
      
      lineblock,
      
      plot_grid(GOBPblock,
        textGrob("log10(p-value)",gp = gpar(fontsize = 8), hjust = 0.25), 
        ncol = 1, 
        rel_heights = c(41,1)
      ),
      NULL,
      plot_grid(KEGGblock,
        textGrob("log10(p-value)",gp = gpar(fontsize = 8), hjust = 0.25), 
        ncol = 1, 
        rel_heights = c(41,1)
      ),
      NULL,
      ncol = 6, rel_widths = col_widths), 
    
    plot_grid(get_legend(ggheatmap + 
                           theme(legend.key.height = unit(10,"points"),
                                 legend.key.width = unit(15,"points"),
                                 legend.title = element_text(size = 8),
                                 legend.text = element_text(size = 8))),
              ncol = 2, rel_width = c(2,1)), 
    nrow = 3, rel_heights = c(0.5,10,1)
    )


```

# Comparison with human datasets

To evaluate the translational relevance of the transcriptional changes in mouse, we utilize data from two studies:

1. A list of DEGs between treatment-naĂŻve UC patient and non-IBD control colon biopsies from [Taman et al, 2018](https://doi.org/10.1093/ecco-jcc/jjx139).

2. Counts from rectal and colon biopsies from UC and CD patients and non-IBD controls from [Lloyd-Price et al, 2019](https://doi.org/10.1038/s41586-019-1237-9). The counts are used to generate DEGs between UC vs controls and CD vs controls.

The mouse homologs of these genes are extracted from the [MGI homolog database](https://www.informatics.jax.org/). The genes are then compared to the mouse DEGs defined above. The number of overlapping genes are shown in the Venn diagram below. Genes that are DE in mouse and both UC comparisons are considered "UC conserved" whereas genes that are DE in all four comparisons are considered "IBD conserved".

```{r}

mousemodules = as.data.frame(logFCs01)
mousemodules$cluster = groups[rownames(mousemodules)]

mouseorder = sort(unique(mousemodules$cluster))

```

```{r}

if(!file.exists(paste0(datadir,"/human_DE.csv"))){
  
  homologs <- fread('http://www.informatics.jax.org/downloads/reports/HMD_HumanPhenotype.rpt')
  homologs <- data.frame("mouse"=homologs$V3, "human"=homologs$V1)
  
  mouse_human = homologs[match(rownames(mousemodules), homologs$mouse),]
  mouse_human = mouse_human[!is.na(mouse_human$mouse),]
  
  mouse_human$cluster = mousemodules$cluster[match(mouse_human$mouse, rownames(mousemodules))]
  
  DE_DF = getHumanDEs(mouse_human, homologs)
  
  write.csv2(DE_DF, paste0(datadir,"Human_DE.csv"))
}else{
  DE_DF = read.csv2(paste0(datadir,"Human_DE.csv"))
}

```


```{r}

gv = ggvenn(list(Taman_UC = unique(DE_DF$mouse[!is.na(DE_DF$Taman_UC_reg)]),
            Lloyd_UC= unique(DE_DF$mouse[!is.na(DE_DF$Lloyd_UC_reg)]),
            CD= unique(DE_DF$mouse[!is.na(DE_DF$Lloyd_CD_reg)]),
            SCID = unique(DE_DF$mouse[!is.na(DE_DF$mousemodule)])),
       show_percentage = FALSE)

gv$coordinates$clip = "off"

gv

```


```{r}

# Label genes regulated in both UC datasets and in UC+CD

DE_DF$UCup = DE_DF$Taman_UC_reg %in% "up" & DE_DF$Lloyd_UC_reg %in% "up"
DE_DF$UCdown = DE_DF$Taman_UC_reg %in% "down" & DE_DF$Lloyd_UC_reg %in% "down"
DE_DF$IBDup = DE_DF$Taman_UC_reg %in% "up" & DE_DF$Lloyd_UC_reg %in% "up" & 
                DE_DF$Lloyd_CD_reg %in% "up"
DE_DF$IBDdown = DE_DF$Taman_UC_reg %in% "down" & DE_DF$Lloyd_UC_reg %in% "down" & 
                  DE_DF$Lloyd_CD_reg %in% "down"
DE_DF$UC = factor(c("down",NA,"up")[DE_DF$UCdown*-1 + DE_DF$UCup + 2], levels = c(NA,"up","down"))
DE_DF$IBD = factor(c("down",NA,"up")[DE_DF$IBDdown*-1 + DE_DF$IBDup + 2], levels = c(NA,"up","down"))

```

```{r}

message("loading KEGG")

# Load kegg and gene ontology information for over-representation analysis

kegg_pathways = loadKEGGGeneSets(
  filepath_kegg = paste0(datadir,"../mouseKEGG.rds"))

universe = rownames(log_tpm)

```




## Conserved UC genes

For each group of conserved genes, the number and regulation of the genes in human are shown per mouse module.

```{r, fig.height = 2.5, fig.width = 7}

plot_grid(ggplot(DE_DF[apply(DE_DF[,c("UC","mousemodule")], 1, function(x){all(!is.na(x))}),], 
                 aes(x = mousemodule)) + geom_bar(aes(fill = as.factor(UC)))  + 
  labs(title = "UC conserved genes", fill = "UC regulation", x = "SCID module") +
    bartheme_human + scale_x_continuous(breaks = mouseorder) + 
    theme(axis.text.y = element_text(size = 8)),
  ggplot(DE_DF[apply(DE_DF[,c("UC","mousemodule")], 1, function(x){all(!is.na(x))}),], 
         aes(x = mousemodule)) + geom_bar(aes(fill = as.factor(UC)), position = "fill")  + 
  labs(title = "UC conserved genes", fill = "UC regulation", x = "SCID module", y = "Proportion") +
    bartheme_human + scale_x_continuous(breaks = mouseorder) + 
    theme(axis.text.y = element_text(size = 8)),
ncol = 2)

```


The distributions of p-values in mouse DEGs for conserved and non-conserved genes are shown below.

```{r, fig.height=2, fig.width=3}


DE_DF$mouse_FDR = timelrt$FDR[match(DE_DF$mouse, rownames(timelrt))]

plottext = element_text(size = 8)
plottheme = theme(axis.text = plottext,
                  axis.title = plottext)


plot_grid(
ggplot(DE_DF[!is.na(DE_DF$mousemodule),], aes(x = !is.na(UC), y = -log10(mouse_FDR))) +
  stat_compare_means(method = "wilcox", label = "p.signif", 
                     comparisons = list(c("TRUE","FALSE"))) + 
  plottheme + scale_y_continuous(expand = expansion(mult = c(0,0.1))) + 
  geom_point(position = "jitter", size = 0.1, alpha =0.3, shape = 16)  +
  geom_boxplot(outlier.size = -1, alpha = 0.5, color = "#7F79B7")
,
ggplot(DE_DF[!is.na(DE_DF$mousemodule),], aes(x = !is.na(IBD), y = -log10(mouse_FDR))) + 
  stat_compare_means(method = "wilcox", label = "p.signif", 
                     comparisons = list(c("TRUE","FALSE"))) + 
  plottheme + scale_y_continuous(expand = expansion(mult = c(0,0.1))) + 
  geom_point(position = "jitter", size = 0.1, alpha =0.3, shape = 16)  +
  geom_boxplot(outlier.size = -1, alpha = 0.5, color = "#7F79B7")

)

```

```{r}

sessionInfo()

```

