---
title: "T cell transfer single cell analysis - Immune fraction"
editor_options:
  chunk_output_type: console
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    code_folding: hide
    toc: yes
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
params:
  dataset: 1
---

```{r Setup, echo = FALSE}
knitr::opts_chunk$set(fig.width  = 10,
                      results    = "hold",
                      message    = FALSE,
                      warning    = FALSE)


```

```{r setup}

library(Seurat)
library(ggplot2)
library(cowplot)
library(clustree)
library(dplyr)
library(reshape2)

source("../code/SingleCell/color_list.R")

```

```{r}
 
datadir = "../data/SingleCell/TcellTransfer/R_objects/"

owncache = "../data/SingleCell/TcellTransfer/cache_immune/"

violintheme = theme(axis.text = element_text(size = 8),
        plot.title = element_text(size = 8),
        legend.key.size = unit(8, "pt"),
        axis.title = element_blank())
dimtheme = theme(legend.text = element_text(size = 8),
        plot.title = element_text(size = 8),
        legend.key.size = unit(8, "pt"))

qc_feats <- c("nFeature_RNA", "nCount_RNA", "percent_mito", "percent_ribo")

```

```{r, include = FALSE}

if(file.exists(owncache)){
  message(paste0("Intermediate data files will be loaded from and saved to ",owncache))
}else{
  message("Intermediate data files will not be saved or loaded")
}

```

# Initial data 

This report describes the analysis performed on scRNA-seq data from immune cells isolated from T cell transfer colitis mouse colons. The data has first been processed together with stroma and epithelial cells, see "TcellTransfer_SingleCell_DataPreparation.html". The data object produced from that analysis is saved as the object data/SingleCell/TcellTransfer/R_objects/immune_data.rds.

```{r loaddata}

if(!file.exists(paste0(owncache,"first_dr.RData"))){
  alldata = readRDS(paste0(datadir,"immune_data.rds"))

DefaultAssay(alldata) = "RNA"

suppressWarnings(
  suppressMessages(
    alldata <-FindVariableFeatures(alldata, selection.method = "vst", 
                                   nfeatures = 2000, verbose = FALSE,
                                   assay = "RNA")))

  alldata <- ScaleData(alldata, vars.to.regress = c("percent_mito", "nFeature_RNA"), 
      assay = "RNA")
  alldata <- RunPCA(alldata, npcs = 50, verbose = F)
  alldata <- RunUMAP(alldata, reduction = "pca", dims = 1:30, 
                     n.components = 2, n.neighbors = 30, 
      n.epochs = 200, min.dist = 0.3, learning.rate = 1, 
      spread = 1, reduction.name = "umap")
  
  if(file.exists(owncache)){
    save(alldata, file = paste0(owncache,"first_dr.RData"))
  }
}else{
  load(paste0(owncache,"first_dr.RData"))
}


```


# Quality control

Number of features and counts and percentage of mitochondrial and ribosomal RNA are plotted below.

```{r qc, fig.height=4, fig.width=6}

VlnPlot(alldata, group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) &
  violintheme

```

# Dimensional reduction - pre-integration

The data is first processed in the same way as the full dataset, with PCA on the 2000 most highly variable genes and subsequent UMAP for visualization. As can be seen below, some differences can be seen between the Rag2-/- and SCID backgrounds at this level of reduction.

## Highly variable features

```{r topgenes, fig.height = 3, fig.width = 5}

top20 <- head(VariableFeatures(alldata), 20)
lp = LabelPoints(plot = VariableFeaturePlot(alldata), points = top20, repel = TRUE) +
  theme(axis.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        axis.text = element_text(size = 8))

lp$layers[[2]]$aes_params$size = 3

lp

```

## PCA

```{r, fig.height = 3, fig.width = 6}

plot_grid(plot_grid(ncol = 3, 
          DimPlot(alldata, reduction = "pca", group.by = "orig.ident", 
                  dims = 1:2, shuffle = TRUE) +
            NoLegend() + dimtheme + 
            theme(axis.text = element_text(size = 8),
                  axis.title = element_text(size = 8)),
          DimPlot(alldata, reduction = "pca", group.by = "orig.ident", 
                  dims = 3:4, shuffle = TRUE) + 
            NoLegend() + dimtheme + 
            theme(axis.text = element_text(size = 8),
                  axis.title = element_text(size = 8)),
          DimPlot(alldata, reduction = "pca", group.by = "orig.ident",
                  dims = 5:6, shuffle = TRUE) + 
            NoLegend() + dimtheme + 
            theme(axis.text = element_text(size = 8),
                  axis.title = element_text(size = 8))),
         get_legend(
           DimPlot(alldata, reduction = "pca", group.by = "orig.ident") +
             theme(legend.position = "top") + dimtheme),
         ncol = 1, rel_heights = c(3,1))

```


```{r, fig.height = 3, fig.width = 6}

VizDimLoadings(alldata, dims = 1:4, reduction = "pca", 
               ncol = 4, balanced = T, nfeatures = 20) &
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 8))

```

```{r, fig.height = 2, fig.width = 4}

ElbowPlot(alldata, reduction = "pca", ndims = 50) + 
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 8))

```

## UMAP

```{r, fig.height = 2.5, fig.width=4}

plot_grid(ncol = 1, DimPlot(alldata, reduction = "umap", raster = TRUE,
                            group.by = "orig.ident", shuffle = TRUE) +
            theme(legend.position = "right") + NoAxes() + dimtheme)

```

```{r firstumap, fig.height = 2.5, fig.width = 5}

plot_grid(ncol = 2, 
      DimPlot(alldata, reduction = "umap", raster = TRUE,
                  group.by = "Model", shuffle = TRUE) +
            theme(legend.position = "bottom") + NoAxes() + dimtheme,
      DimPlot(alldata, reduction = "umap", raster = TRUE,
                  group.by = "Treatment", shuffle = TRUE) +
            theme(legend.position = "bottom") + NoAxes() + dimtheme)


plot_grid(ncol = 2,
          DimPlot(alldata, reduction = "umap", raster = TRUE,
                  group.by = "RPCA_cluster", label = TRUE, shuffle = TRUE) +
            NoLegend() + NoAxes() + dimtheme)

FeaturePlot(alldata, reduction = "umap", features = c("Ptprc","Epcam"), raster = TRUE,
    order = T, slot = "data", combine = T) & dimtheme & NoAxes() & NoLegend()

```

# Integration

To enable equivalent clustering across samples, CCA integration is run with standard Seurat functions. 

```{r}

plan("multisession", workers = 6)

if(!file.exists(paste0(owncache,"integration.RData"))){
  if(!file.exists(paste0(owncache,"anchors.RData"))){

    # Works with 256GB on 6 workers
    alldata <- SplitObject(alldata, split.by = "orig.ident")
    
    alldata = lapply(alldata, function(x){
        x <- NormalizeData(x, verbose = FALSE)
        x <- FindVariableFeatures(x, selection.method = "vst",
                nfeatures = 2000, verbose = FALSE)
      })
        
    data.anchors <- FindIntegrationAnchors(object.list = alldata, 
                                                dims = 1:30, reduction = "cca")
    rm(alldata)
    gc()

    if(file.exists(owncache)){
          save(data.anchors, file = paste0(owncache,"anchors.RData"))
    }
  }else{
    rm(alldata)
    gc()
    load(paste0(owncache,"anchors.RData"))
  }
  
  plan("sequential")
  
  # works with 256GB in sequential 
  alldata <- IntegrateData(anchorset = data.anchors, dims = 1:30, new.assay.name = "CCA",
     sample.tree = cbind(c(-4,-1,2,3,-8,5,-10,7,-15,9,4,8,11,12),
                         c(-5,-2,-3,1,-7,-6,-11,-12,-14,-13,6,10,-9,13)))
  
  rm(data.anchors)
  gc()
  
  alldata <- ScaleData(alldata)
  alldata <- RunPCA(alldata, npcs = 30, verbose = FALSE)
  alldata <- RunUMAP(alldata, dims = 1:30)
  
  if(file.exists(owncache)){
    save(alldata, file = paste0(owncache,"integration.RData"))
  }
}else{
  load(paste0(owncache,"integration.RData"))
}

plan("sequential")

```

## Dimensional reduction

```{r, fig.height = 2.5, fig.width=7}

plot_grid(ncol = 3,
          DimPlot(alldata, reduction = "pca", group.by = "orig.ident", 
                  pt.size = 0.01, shuffle = TRUE) +
            NoAxes() + ggtitle("PCA integrated") + NoLegend() + dimtheme,
          DimPlot(alldata, reduction = "umap", group.by = "orig.ident", 
                  pt.size = 0.01, shuffle = TRUE) +
            NoAxes() + ggtitle("UMAP integrated") + NoLegend() + dimtheme,
          get_legend(DimPlot(alldata, reduction = "pca", group.by = "orig.ident") +
                       theme(legend.text = element_text(size = 8)) + dimtheme)
)
plot_grid(ncol = 3,
          DimPlot(alldata, reduction = "umap", group.by = "Model", 
                  pt.size = 0.01, shuffle = TRUE) +
            NoAxes() + dimtheme + theme(legend.position = "bottom"),
          DimPlot(alldata, reduction = "umap", group.by = "Treatment", 
                  pt.size = 0.01, shuffle = TRUE) +
            NoAxes() + dimtheme + theme(legend.position = "bottom"),
          DimPlot(alldata, reduction = "umap", group.by = "RPCA_cluster", 
                  pt.size = 0.01, shuffle = TRUE, label = TRUE) +
            NoAxes() + NoLegend() + dimtheme
)

alldata@active.assay = "RNA"
plot_grid(FeaturePlot(alldata, reduction = "umap", features = c("Ptprc","Epcam"), 
    order = T, slot = "data", combine = T) & dimtheme & NoAxes(), 
    NULL, rel_widths = c(4,1))
alldata@active.assay = "CCA"

```

```{r, fig.height = 5, fig.width=7}

FeaturePlot(alldata, reduction = "umap", 
            features = c("percent_mito","percent_ribo","S.Score",
                         "G2M.Score", "nFeature_RNA"), 
    order = T, slot = "data", combine = T, ncol = 3) &
  NoAxes() & NoLegend() & dimtheme

```

# Clustering

The dataset is clustered using SNN and the Louvain algorithm at 4 resolutions. Resolution 0.5 is chosen for further analysis.

```{r}

if(!file.exists(paste0(owncache, "clustered.RData"))){
  alldata@active.assay <- "CCA"
  
  alldata@active.assay

  alldata <- FindNeighbors(alldata, dims = 1:30, k.param = 350, prune.SNN = 1/15)

  
  # Clustering with louvain (algorithm 1)
  for (res in c(0.25, 0.5, 1, 1.5)) {
      alldata <- FindClusters(alldata, graph.name = "CCA_snn",
                              resolution = res, algorithm = 1)
  }
  if(file.exists(owncache)){
    save(alldata, file = paste0(owncache, "clustered.RData"))
  }
}else{
  load(paste0(owncache, "clustered.RData"))
}



```

```{r dimplots, fig.height=5, fig.width = 5}

plot_grid(ncol = 2, 
          DimPlot(alldata, reduction = "umap", group.by = "CCA_snn_res.0.25",
                  label = TRUE) + dimtheme + NoLegend() + NoAxes() + 
            ggtitle("louvain_0.25"),
          DimPlot(alldata, reduction = "umap", group.by = "CCA_snn_res.0.5", 
                  label = TRUE) + dimtheme +  NoLegend() + NoAxes() + 
            ggtitle("louvain_0.5"),
          DimPlot(alldata, reduction = "umap", group.by = "CCA_snn_res.1", 
                  label = TRUE) + dimtheme + NoLegend() + NoAxes() + 
            ggtitle("louvain_1"), 
          DimPlot(alldata, reduction = "umap", group.by = "CCA_snn_res.1.5", 
                  label = TRUE) + dimtheme + NoLegend() + NoAxes() + 
            ggtitle("louvain_1.5"))


```

```{r clustreeplot, fig.height = 3, fig.width = 6}

ct = clustree(alldata@meta.data, prefix = "CCA_snn_res.")

ct$layers[[1]]$aes_params$edge_width = 1

plot_grid(ggplot(unique(ct$data[,c("y","CCA_snn_res.")]),
       aes(y = y)) +
  geom_text(x = 1, aes(label = CCA_snn_res.), size = 3) + 
  scale_x_continuous(limits = c(0,2), position = "top") + 
  theme(panel.background = element_blank(),
        panel.grid = element_blank(),
        axis.text = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_text(size = 8),
        axis.ticks = element_blank()) + 
    labs(x= "Resolution"),
  ct + 
  scale_size_continuous(range = c(1,5)) + 
  scale_colour_discrete(guide = "none") + 
  theme(legend.key.size = unit(8,"pt"),
        legend.title = element_text(size = 8),
        legend.margin = margin(1,1,1,1,unit = "pt")),
  align = "h", rel_widths = c(1,8))

```

# Differential expression and annotation of clusters

The different clusters are annotated using differential expression and known marker genes. For legibility, the annotations are here given prior to differential expression and gene marker information.

```{r}

clusternames = 
  data.frame(Cluster = c(0:16),
             Clustername = c("CD4+ T cells Emb","Monocytes",
                             "Cytotoxic CD4+","Macrophages Apoe",
                             "NK","ILC2","ILC3","Proliferating",
                             "Low counts","DC1","Neutrophils",
                             "Mature DC","Treg","Mast cells",
                             "Macrophages Retnla","pDC",
                             "IFN response CD4+"))

clusterorder = 
  c("CD4+ T cells Emb","Cytotoxic CD4+","Treg","IFN response CD4+",
    "Proliferating","ILC2","ILC3","NK","Monocytes","Macrophages Apoe",
    "Macrophages Retnla","DC1","Mature DC","pDC","Neutrophils",
    "Mast cells","Low counts")

alldata = AddMetaData(alldata, 
                      factor(clusternames$Clustername[
                        match(alldata$CCA_snn_res.0.5,clusternames$Cluster)], 
                        levels = clusterorder),
                      "NamedCluster")

alldata = SetIdent(alldata, value = "NamedCluster")

color_list = get_color_list(length(clusternames$Clustername))
names(color_list) = clusterorder

```

```{r, fig.height = 4, fig.width = 4}

dp = DimPlot(alldata, label = TRUE, shuffle = TRUE) + 
  NoAxes() + NoLegend() + dimtheme + scale_color_manual(values = color_list)

dp$layers[[2]]$aes_params$size = 3

dp

```

## Differential expression

The top 25 DEGs (based on p-value) for each cluster are shown below, ordered by log2(FC).

```{r, fig.width = 2, fig.height = 3}

alldata@active.assay = "RNA"

# Becomes slower when parallel

plan("sequential")

if(!file.exists(paste0(owncache, "clustermarkers.rds"))){
  markers_genes <- FindAllMarkers(alldata, logfc.threshold = 0.2, test.use = "wilcox", 
    min.pct = 0.1, min.diff.pct = 0.2, only.pos = TRUE, max.cells.per.ident = 1000, 
    assay = "RNA", random.seed = 42)
            
  if(file.exists(owncache)){
    saveRDS(markers_genes, paste0(owncache, "clustermarkers.rds"))
  }
}else{
   markers_genes = readRDS(paste0(owncache, "clustermarkers.rds"))
}

top25 <- markers_genes %>% group_by(cluster) %>% top_n(-25, p_val)

out = lapply(unique(top25$cluster), function(i){
  markersdata = top25[top25$cluster==i,]
  markersdata$gene = 
    factor(markersdata$gene, 
           levels = markersdata$gene[order(markersdata$avg_log2FC)])
  print(ggplot(markersdata, aes(x = avg_log2FC, y = gene)) + 
          labs(title = i) + 
    geom_bar(stat = "identity") + 
    theme(axis.title.y = element_blank(),
          axis.title.x = element_text(size = 8),
          plot.title = element_text(size = 8),
          axis.text = element_text(size = 8)))
})

```

```{r}

markergenes = c('Icos', 'Il7r', 'Cd3d', 'Cd3e', 'Cd3g', 'Cd4',
                'Emb', 'Lef1', 'Cxcr6', 'Ccl5', 'Eomes', 'Gzmk',
                'Cd8a', 'Tnfrsf4', 'Ctla4', 'Foxp3', 'Tnfsf8',
                'Ifit1', 'Zbp1', 'Stmn1', 'Mki67', 'Gata3', 'Areg', 
                'Il22', 'Nrgn', 'Gpr183', 'Ncr1', 'Xcl1', 'Klrc1',
                'Cd19', 'Ms4a1', 'Bank1', 'Jchain', 'Mzb1', 'Cd14',
                'Csf1r', 'Lyz2', 'Ms4a6d', 'Apoe', 'C1qa', 'Mafb',
                'Retnla', 'Fn1', 'Cd74', 'Ly6c2', 'Xcr1', 'Cd86',
                'Ccr7', 'Irf8', 'Bst2', 'Siglech', 'S100a9', 'S100a8',
                'Cpa3', 'Mcpt1', 'Retnlg'
)

```

```{r dotplot, fig.height = 7, fig.width = 5}

alldata@active.ident = factor(alldata$NamedCluster,
                              levels = clusterorder)
dot = DotPlot(alldata, features=rev(markergenes))
dot = dot + aes(x = id, y = features.plot) +
  scale_x_discrete(position = "top") + 
  labs(y = "Genes",x = "Clusters") + 
  scale_size(range = c(0.5,2.5), breaks = c(0,50,100), limits = c(0,100)) +
  theme(legend.position = "right", legend.title = element_blank(),
        axis.text.x = element_text(size = 8, angle = 90, hjust = 0),
        axis.text.y = element_text(size = 8, face = "italic"),
        legend.key.width = unit(0.3, "cm"), 
        legend.key.height = unit(0.3,"cm"), 
        legend.text = element_text(size = 8),
        axis.title = element_text(size = 8)) 

dot$scales$scales[[1]]$limits = c(-1.8,2.7)

print(dot)

```


# CITE-seq

```{r}

DefaultAssay(alldata) = "FB"

alldata <- NormalizeData(alldata)
alldata <- ScaleData(alldata)
  
```

```{r, fig.height = 3.5, fig.width = 5}

alldata@active.assay <- "FB"

dot = DotPlot(alldata, assay = "FB", features = rev(rownames(alldata)), 
              group.by = "NamedCluster") 
dot = dot + aes(x = id, y = features.plot) + scale_x_discrete(position = "top") +
  labs(y = "Antibodies",x = "Clusters") +
  scale_size(range = c(0.1,2.5), breaks = c(60,80,100)) +
  theme(legend.position = "bottom", legend.title = element_blank(),
        axis.text.x = element_text(size = 8, hjust = 0, angle = 90),
        axis.text.y = element_text(size = 8),
        legend.key.width = unit(0.3, "cm"),
        legend.key.height = unit(0.3,"cm"),
        legend.text = element_text(size = 8),
        axis.title.y = element_text(size = 8),
        axis.title.x = element_blank()) +
  scale_y_discrete(labels = function(x){gsub("-TotalSeqB","",x)})

dot

```

# Population changes over time

The proportion of each cluster among the total stromal population is shown across time points below. The order of clusters is based on the average correlation between proportion and time point for each background (untreated samples only), with positive correlation at the top and negative correlation on the bottom.

```{r}

clean_sample_names = function(sam){
  for(s in c("_Imm","I_","_Epi","_E","_Str","S_","_ut","G_")){
    sam = gsub(s, "",sam)
  }
  gsub("il","aIL12",gsub("_","\n",sam))
}

```

```{r}

alldata$Treatment = factor(alldata$Treatment,
                           levels = c("ut","il"))

frequencies = as.data.frame(table(data.frame(cluster=alldata$NamedCluster,
                 sample = alldata$orig.ident, model = alldata$Model)))
frequencies$Proportion=NA
for(i in 1:nrow(frequencies)){
  frequencies$Proportion[i] =
    frequencies$Freq[i]/
    sum(frequencies$Freq[frequencies$sample==frequencies$sample[i]])
}

frequencies$time = alldata$Time[match(frequencies$sample,
                                      alldata$orig.ident)]
frequencies$treatment = 
  alldata$Treatment[match(frequencies$sample, 
                          alldata$orig.ident)]

frequencies = frequencies[
  (substr(frequencies$time,1,1) == "W" & frequencies$model == "SCID") |
    (substr(frequencies$time,1,1) == "D" & frequencies$model == "Rag2"),]


clustercors = lapply(c("SCID","Rag2"), function(bg){
  corlist = lapply(unique(frequencies$cluster), function(i){
    clufreq = frequencies[frequencies$cluster==i  & 
                          frequencies$treatment == "ut" & 
                          frequencies$model == bg,]

    clucor = cor(clufreq$Proportion,as.numeric(
      as.factor(as.character(clufreq$time))))
    clucor
    })
  names(corlist) = unique(frequencies$cluster)
  corlist
  })
names(clustercors) = c("SCID","Rag2")

averagecor = lapply(1:length(clustercors[[1]]), function(i){
  mean(c(clustercors[[1]][[i]], clustercors[[2]][[i]]))
})
names(averagecor) = names(clustercors[[1]])

frequencies$cluster = 
  factor(frequencies$cluster, 
         levels = names(averagecor)[order(unlist(averagecor),
                                          decreasing = TRUE)])

frequencies$sample = 
  factor(frequencies$sample, 
         levels = 
           unique(frequencies$sample[
             order(factor(frequencies$model, levels = c("SCID","Rag2")),
                   frequencies$treatment,
                   frequencies$time)]))
frequencies$samplenumber = as.numeric(frequencies$sample)


cumulFreq = frequencies
cumulFreq$cumulProp = NA
for(i in unique(cumulFreq$sample)){
  cumulFreq$cumulProp[cumulFreq$sample %in% i] = rev(cumsum(rev(cumulFreq$Proportion[cumulFreq$sample %in% i])))
}
cumulFreq = rbind(cumulFreq,cumulFreq)
cumulFreq$xadj = rep(c(-1,1), c(nrow(frequencies),nrow(frequencies)))
cumulFreq$oldtime = cumulFreq$time
cumulFreq$time = as.numeric(
  gsub("D","",gsub("W","",as.character(cumulFreq$time))))

datasets = lapply(names(clustercors), function(bg){
  bgfreq = frequencies[frequencies$model == bg,]
  bglist = lapply(unique(bgfreq$treatment), function(i){
    trfr = bgfreq[bgfreq$treatment==i,]
    cumtrfr = cumulFreq[paste(cumulFreq$sample,
                                     cumulFreq$model,
                                     cumulFreq$oldtime) %in% 
                                 paste(trfr$sample,
                                       trfr$model,
                                       trfr$time),]
    list(trfr, cumtrfr)
  })
  names(bglist) = unique(bgfreq$treatment)
  bglist
})

barplot = ggplot(data = frequencies, 
       aes(x = samplenumber, y = Proportion, fill = cluster)) + 
  theme(panel.background = element_blank(), axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
            scale_fill_manual(
              values =
                color_list[as.character(unique(sort(frequencies$cluster)))]) + 
            scale_x_continuous(
              breaks = sort(unique(frequencies$samplenumber)),
              labels = clean_sample_names(levels(frequencies$sample)))

for(ds in datasets){
  for(dst in ds){
    barplot = barplot + 
      geom_area(data = dst[[2]],
                alpha = 0.7, size = 0.3, stat = 'identity',
                aes(fill = cluster, group = cluster,
                    y = Proportion, x = samplenumber + xadj*0.3))}
}

```

```{r, fig.height = 3.5, fig.width = 7}

barplot + geom_bar(stat = "identity", width = 0.6, color = "black") + 
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 8),
        legend.key.size = unit(11,"pt")) 

```

# Subset analyses

To evaluate the changes of gene expression specfically in CD4+ T cells as well as neutrophils, these clusters are isolated. The UMAP shows the included cells for each analysis.

```{r}

DefaultAssay(alldata) = "RNA"

alldata$ModelTreatmentTime = paste(alldata$Model, 
                                     alldata$Treatment, 
                                     alldata$Time)
alldata$ModelTreatmentTime = factor(alldata$ModelTreatmentTime,
      levels = unique(alldata$ModelTreatmentTime[order(
        factor(alldata$Model, levels = c("SCID", "Rag2")),
        factor(alldata$Treatment, levels = c("ut","il")),
        alldata$Time)]))
alldata$ModelTreatment = paste(alldata$Model, 
                                     alldata$Treatment)
alldata$ModelTreatment = factor(alldata$ModelTreatment,
      levels = unique(alldata$ModelTreatment[order(
        factor(alldata$Model, levels = c("SCID", "Rag2")),
        factor(alldata$Treatment, levels = c("ut","il")))]))

alldata = AddMetaData(alldata, alldata$NamedCluster, "CD4Tcells")
alldata = AddMetaData(alldata, alldata$NamedCluster, "neutrophils")

alldata$CD4Tcells[!(alldata$CD4Tcells %in% 
                  c("CD4+ T cells Emb","Cytotoxic CD4+",
                    "Treg","IFN response CD4+"))] = NA
alldata$CD4Tcells[alldata$RelTime < 0.1] = NA

alldata$neutrophils[!(alldata$neutrophils %in% 
                  c("Neutrophils"))] = NA

```

```{r, fig.height = 3, fig.width = 6}

plot_grid(plotlist = lapply(c("CD4Tcells","neutrophils"),
       function(x){
         dp = DimPlot(alldata, group.by = x, label = TRUE, shuffle = TRUE) +
            scale_color_manual(values = color_list, na.value = "#CACACA") +
            dimtheme & NoLegend() & NoAxes()
         dp$layers[[2]]$aes_params$size = 3
         dp
       }), ncol = 2)

```

```{r}

alldata@active.assay <- "RNA"
alldata[["CCA"]] = NULL

CD4Tcells = 
  subset(alldata, 
         cells = WhichCells(
           alldata, expression = 
             CD4Tcells %in% alldata$NamedCluster))
neutrophils = 
  subset(alldata, 
         cells = WhichCells(
           alldata, 
           expression = neutrophils %in% alldata$NamedCluster))

rm(alldata)
trash = gc()

saveRDS(CD4Tcells,paste0(datadir,"CD4Tcells.rds"))
saveRDS(neutrophils,paste0(datadir,"neutrophils.rds"))

```

# Session info

```{r}

sessionInfo()

```
