---
title: "Preparation of data for T cell transfer single cell analysis"
editor_options:
  chunk_output_type: console
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    code_folding: hide
    toc: yes
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
params:
  dataset: 1
---

```{r Setup, echo = FALSE}
knitr::opts_chunk$set(fig.width  = 10,
                      results    = "hold",
                      message    = FALSE,
                      warning    = FALSE)


```

```{r setup}

library(Seurat)
library(ggplot2)
library(cowplot)
library(clustree)

library(future)
plan("sequential")

source("../code/SingleCell/color_list.R")
source("../code/SingleCell/NMFsetup.R", chdir = TRUE)


```

```{r}
 
datadir = "../data/SingleCell/TcellTransfer/R_objects/"

owncache = "../data/SingleCell/TcellTransfer/cache_immune/"

violintheme = theme(axis.text = element_text(size = 8),
        plot.title = element_text(size = 8),
        legend.key.size = unit(8, "pt"),
        axis.title = element_blank())
dimtheme = theme(legend.text = element_text(size = 8),
        plot.title = element_text(size = 8),
        legend.key.size = unit(8, "pt"))

qc_feats <- c("nFeature_RNA", "nCount_RNA", "percent_mito", "percent_ribo")

```

```{r, include = FALSE}

if(file.exists(owncache)){
  message(paste0("Intermediate data files will be loaded from and saved to ",owncache))
}else{
  message("Intermediate data files will not be saved or loaded")
}

```

# Initial data 


```{r loaddata}

if(!file.exists(paste0(owncache,"first_dr.RData"))){
  alldata = readRDS(paste0(datadir,"immune_data.rds"))

DefaultAssay(alldata) = "RNA"

suppressWarnings(
  suppressMessages(
    alldata <-FindVariableFeatures(alldata, selection.method = "vst", 
                                   nfeatures = 2000, verbose = FALSE,
                                   assay = "RNA")))

  alldata <- ScaleData(alldata, vars.to.regress = c("percent_mito", "nFeature_RNA"), 
      assay = "RNA", reduction.name = "pca")
  alldata <- RunPCA(alldata, npcs = 50, verbose = F)
  alldata <- RunUMAP(alldata, reduction = "pca", dims = 1:30, 
                     n.components = 2, n.neighbors = 30, 
      n.epochs = 200, min.dist = 0.3, learning.rate = 1, 
      spread = 1, reduction.name = "umap")
  
  if(file.exists(owncache)){
    save(alldata, file = paste0(owncache,"first_dr.RData"))
  }
}else{
  load(paste0(owncache,"first_dr.RData"))
}


```


# Plot QC vars

Number of features and counts and percentage of mitochondrial and ribosomal RNA are plotted below.

```{r qc, fig.height=4, fig.width=6}

VlnPlot(alldata, group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) &
  violintheme

```

# Highly variable features

```{r topgenes, fig.height = 3, fig.width = 5}

top20 <- head(VariableFeatures(alldata), 20)
lp = LabelPoints(plot = VariableFeaturePlot(alldata), points = top20, repel = TRUE) +
  theme(axis.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        axis.text = element_text(size = 8))

lp$layers[[2]]$aes_params$size = 3

lp

```

# PCA

```{r, fig.height = 3, fig.width = 6}

plot_grid(ncol = 3, 
          DimPlot(alldata, reduction = "pca", group.by = "orig.ident", 
                  dims = 1:2, shuffle = TRUE) +
            NoLegend() + dimtheme + 
            theme(axis.text = element_text(size = 8),
                  axis.title = element_text(size = 8)),
          DimPlot(alldata, reduction = "pca", group.by = "orig.ident", 
                  dims = 3:4, shuffle = TRUE) + 
            NoLegend() + dimtheme + 
            theme(axis.text = element_text(size = 8),
                  axis.title = element_text(size = 8)),
          DimPlot(alldata, reduction = "pca", group.by = "orig.ident",
                  dims = 5:6, shuffle = TRUE) + 
            NoLegend() + dimtheme + 
            theme(axis.text = element_text(size = 8),
                  axis.title = element_text(size = 8)))

```


```{r, fig.height = 3, fig.width = 6}

VizDimLoadings(alldata, dims = 1:5, reduction = "pca", ncol = 5, balanced = T) &
  theme(axis.text = element_text(size = 8), axis.title = element_text(size = 8))

```


```{r, fig.height = 2, fig.width = 4}

ElbowPlot(alldata, reduction = "pca", ndims = 50) + 
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 8))

```

# UMAP

```{r firstumap, fig.height = 3, fig.width = 6}

plot_grid(ncol = 2, DimPlot(alldata, reduction = "pca", raster = TRUE,
                            group.by = "orig.ident", shuffle = TRUE) + 
            NoLegend() + NoAxes() + dimtheme, 
    DimPlot(alldata, reduction = "umap", group.by = "orig.ident", raster = TRUE,
            shuffle = TRUE) +
      NoLegend() + NoAxes() + dimtheme)

plot_grid(ncol = 2, DimPlot(alldata, reduction = "umap", raster = TRUE,
                            group.by = "orig.ident", shuffle = TRUE) +
            theme(legend.position = "bottom") + NoAxes() + dimtheme, 
          DimPlot(alldata, reduction = "umap", raster = TRUE,
                  group.by = "Model", shuffle = TRUE) +
            theme(legend.position = "bottom") + NoAxes() + dimtheme)

plot_grid(ncol = 2, DimPlot(alldata, reduction = "umap", raster = TRUE,
                  group.by = "Treatment", shuffle = TRUE) +
            theme(legend.position = "bottom") + NoAxes() + dimtheme,
          DimPlot(alldata, reduction = "umap", raster = TRUE,
                  group.by = "RPCA_cluster", label = TRUE, shuffle = TRUE) + 
            NoLegend() + NoAxes() + dimtheme)

FeaturePlot(alldata, reduction = "umap", features = c("Ptprc","Epcam"), raster = TRUE,
    order = T, slot = "data", combine = T) & dimtheme & NoAxes() & NoLegend()

```


# Integration


```{r}


if(!file.exists(paste0(owncache,"integration.RData"))){
  if(!file.exists(paste0(owncache,"anchors.RData"))){
    alldata <- SplitObject(alldata, split.by = "orig.ident")
    
    alldata = lapply(alldata, function(x){
        x <- NormalizeData(x, verbose = FALSE)
        x <- FindVariableFeatures(x, selection.method = "vst",
                nfeatures = 2000, verbose = FALSE)
      })
        
    data.anchors <- FindIntegrationAnchors(object.list = alldata, 
                                                dims = 1:30, reduction = "cca")
    rm(alldata)
    gc()

    if(file.exists(owncache)){
          save(data.anchors, file = paste0(owncache,"anchors.RData"))
    }
  }else{
    rm(alldata)
    gc()
    load(paste0(owncache,"anchors.RData"))
  }
 alldata <- IntegrateData(anchorset = data.anchors, dims = 1:30, new.assay.name = "CCA",
     sample.tree = cbind(c(-4,-1,2,3,-8,5,-10,7,-15,9,4,8,11,12),
                         c(-5,-2,-3,1,-7,-6,-11,-12,-14,-13,6,10,-9,13)))
  
  rm(data.anchors)
  gc()
  
  alldata <- ScaleData(alldata)
  alldata <- RunPCA(alldata, npcs = 30, verbose = FALSE)
  alldata <- RunUMAP(alldata, dims = 1:30)
  
  if(file.exists(owncache)){
    save(alldata, file = paste0(owncache,"integration.RData"))
  }
}else{
  load(paste0(owncache,"integration.RData"))
}

```

```{r, fig.height = 2, fig.width=5}

plot_grid(ncol = 2,
  
  DimPlot(alldata, reduction = "pca", group.by = "orig.ident", pt.size = 0.01, shuffle = TRUE) +
    NoAxes() + ggtitle("PCA integrated") + NoLegend() + dimtheme,
  DimPlot(alldata, reduction = "umap", group.by = "orig.ident", pt.size = 0.01, shuffle = TRUE) +
    NoAxes() + ggtitle("UMAP integrated") + NoLegend() + dimtheme
)

alldata@active.assay = "RNA"
FeaturePlot(alldata, reduction = "umap", features = c("Ptprc","Epcam"), 
    order = T, slot = "data", combine = T) & dimtheme & NoAxes()
alldata@active.assay = "CCA"

```



```{r, fig.height = 6, fig.width=4}

FeaturePlot(alldata, reduction = "umap", 
            features = c("percent_mito","percent_ribo","S.Score",
                         "G2M.Score", "nFeature_RNA"), 
    order = T, slot = "data", combine = T, ncol = 2) &
  NoAxes() & NoLegend() & dimtheme

```



# Clustering

```{r}

if(!file.exists(paste0(owncache, "clustered.RData"))){
  alldata@active.assay <- "CCA"
  
  alldata@active.assay

  alldata <- FindNeighbors(alldata, dims = 1:30, k.param = 350, prune.SNN = 1/15)

  
  # Clustering with louvain (algorithm 1)
  for (res in c(0.25, 0.5, 1, 1.5)) {
      alldata <- FindClusters(alldata, graph.name = "CCA_snn",
                              resolution = res, algorithm = 1)
  }
  if(file.exists(owncache)){
    save(alldata, file = paste0(owncache, "clustered.RData"))
  }
}else{
  load(paste0(owncache, "clustered.RData"))
}



```



```{r dimplots, fig.height=6, fig.width = 6}

plot_grid(ncol = 2, 
          DimPlot(alldata, reduction = "umap", group.by = "CCA_snn_res.0.25",
                  label = TRUE) + dimtheme + NoLegend() + NoAxes() + 
            ggtitle("louvain_0.25"),
          DimPlot(alldata, reduction = "umap", group.by = "CCA_snn_res.0.5", 
                  label = TRUE) + dimtheme +  NoLegend() + NoAxes() + 
            ggtitle("louvain_0.5"),
          DimPlot(alldata, reduction = "umap", group.by = "CCA_snn_res.1", 
                  label = TRUE) + dimtheme + NoLegend() + NoAxes() + 
            ggtitle("louvain_1"), 
          DimPlot(alldata, reduction = "umap", group.by = "CCA_snn_res.1.5", 
                  label = TRUE) + dimtheme + NoLegend() + NoAxes() + 
            ggtitle("louvain_1.5"))


```


```{r clustreeplot}

clustree(alldata@meta.data, prefix = "CCA_snn_res.")

```


```{r}

clusternames = 
  data.frame(Cluster = c(0:16),
             Clustername = c("CD4+ T cells Emb","Monocytes",
                             "Cytotoxic CD4+","Macrophages Apoe",
                             "NK","ILC2","ILC3","Proliferating",
                             "Low counts","DC1","Neutrophils",
                             "Mature DC","Treg","Mast cells",
                             "Macrophages Retnla","pDC",
                             "IFN response CD4+"))

clusterorder = 
  c("CD4+ T cells Emb","Cytotoxic CD4+","Treg","IFN response CD4+",
    "Proliferating","ILC2","ILC3","NK","Monocytes","Macrophages Apoe",
    "Macrophages Retnla","DC1","Mature DC","pDC","Neutrophils",
    "Mast cells","Low counts")

alldata = AddMetaData(alldata, 
                      factor(clusternames$Clustername[
                        match(alldata$CCA_snn_res.0.5,clusternames$Cluster)], 
                        levels = clusterorder),
                      "NamedCluster")

alldata = SetIdent(alldata, value = "NamedCluster")

color_list = get_color_list(17)
names(color_list) = clusterorder

```

```{r, fig.height = 4, fig.width = 4}

dp = DimPlot(alldata, label = TRUE, shuffle = TRUE) + 
  NoAxes() + NoLegend() + dimtheme + scale_color_manual(values = color_list)

dp$layers[[2]]$aes_params$size = 3

dp

```

# Differential expression

```{r}

sel.lev = 0.5
sel.clust = paste0("CCA_snn_res.",sel.lev)

alldata <- SetIdent(alldata, value = sel.clust)

```

```{r}

markers_genes <- FindAllMarkers(alldata, logfc.threshold = 0.2, test.use = "wilcox", 
    min.pct = 0.1, min.diff.pct = 0.2, only.pos = TRUE, max.cells.per.ident = 1000, 
    assay = "RNA", random.seed = 42)

top25 <- markers_genes %>% group_by(cluster) %>% top_n(-25, p_val)

mypar(2, 5, mar = c(4, 6, 3, 1))
for (i in unique(top25$cluster)) {
    barplot(sort(setNames(top25$avg_log2FC, top25$gene)[top25$cluster == i], F),
            horiz = T, 
        las = 1, main = paste0(i, " vs. rest"), border = "white", yaxs = "i")
    abline(v = c(0, 0.25), lty = c(1, 2))
}

```


```{r}

markergenes = c('Icos', 'Il7r', 'Cd3d', 'Cd3e', 'Cd3g', 'Cd4',
                'Emb', 'Lef1', 'Cxcr6', 'Ccl5', 'Eomes', 'Gzmk',
                'Cd8a', 'Tnfrsf4', 'Ctla4', 'Foxp3', 'Tnfsf8',
                'Ifit1', 'Zbp1', 'Stmn1', 'Mki67', 'Gata3', 'Areg', 
                'Il22', 'Nrgn', 'Gpr183', 'Ncr1', 'Xcl1', 'Klrc1',
                'Cd19', 'Ms4a1', 'Bank1', 'Jchain', 'Mzb1', 'Cd14',
                'Csf1r', 'Lyz2', 'Ms4a6d', 'Apoe', 'C1qa', 'Mafb',
                'Retnla', 'Fn1', 'Cd74', 'Ly6c2', 'Xcr1', 'Cd86',
                'Ccr7', 'Irf8', 'Bst2', 'Siglech', 'S100a9', 'S100a8',
                'Cpa3', 'Mcpt1', 'Retnlg'
)

clusternames = 
  data.frame(Cluster = 0:16, 
             imm_clusters = c('CD4+ T cells Emb', 'Monocytes', 'Cytotoxic CD4+',
                              'Macrophages Apoe', 'NK', 'ILC2', 'ILC3',
                              'Proliferating', 'Low counts', 'DC1', 'Neutrophils',
                              'Mature DC', 'Treg', 'Mast cells', 'Macrophages Retnla',
                              'pDC', 'IFN response CD4+')
             )
clusterorder = c('CD4+ T cells Emb', 'Cytotoxic CD4+', 'Treg', 'IFN response CD4+',
                           'Proliferating', 'ILC2', 'ILC3', 'NK', 'Monocytes',
                           'Macrophages Apoe', 'Macrophages Retnla', 'DC1', 'Mature DC', 
                           'pDC', 'Neutrophils', 'Mast cells', 'Low counts')

alldata = 
  AddMetaData(alldata, 
              factor(clusternames[match(alldata@active.ident, clusternames$Cluster),
                                  paste0(celltype,"_clusters")], levels = clusterorder),
              col.name = "namedcluster")

color_list <- get_color_list(n=length(unique(alldata@active.ident)))
names(color_list) = clusterorder

```

```{r dotplot, fig.height = 2.2, fig.width = 6}

alldata@active.ident = factor(alldata$namedcluster, levels = rev(clusterorder))
dot = DotPlot(alldata, features=markergenes)
dot = dot + aes(x = id, y = features.plot) +
  scale_y_discrete(position = "right") + 
  coord_flip() + 
  labs(y = "Genes",x = "Clusters") + 
  scale_size(range = c(0.5,2.5), breaks = c(0,50,100), limits = c(0,100)) +
  theme(legend.position = "right", legend.title = element_blank(),
        axis.text.x.top = element_text(size = 6, angle = 90, vjust =0.5,
                                   hjust = 0, face = "italic"),
        axis.text.y = element_text(size = 6),
        legend.key.width = unit(0.3, "cm"), 
        legend.key.height = unit(0.3,"cm"), 
        legend.text = element_text(size = 6),
        axis.title = element_text(size = 8))
dot$scales$scales[[1]]$limits = c(-1.8,2.7)

print(dot)

```

```{r annotateddimplot}

plot_grid(ncol = 2, DimPlot(alldata, label = T, group.by = "namedcluster") +
            NoAxes() + NoLegend() + scale_color_manual(values = color_list))

```

# Population changes over time

```{r}

clean_sample_names = function(sam){
  for(s in c("_Imm","I_","_Epi","_E","_Str","S_","_ut","G_", "_aIL12","_il")){
    sam = gsub(s, "",sam)
  }
  sam
}

```



```{r}

frequencies = as.data.frame(table(data.frame(cluster=metadata_all[,sel.clust],
                 sample = metadata_all$orig.ident, model = metadata_all$Model)))
frequencies$Proportion=NA
for(i in 1:nrow(frequencies)){
  frequencies$Proportion[i] =
    frequencies$Freq[i]/
    sum(frequencies$Freq[frequencies$sample==frequencies$sample[i]])
}

frequencies$time = metadata_all$Time[match(frequencies$sample, metadata_all$orig.ident)]
frequencies$treatment = metadata_all$Treatment[match(frequencies$sample, 
                                                     metadata_all$orig.ident)]

frequencies$namedcluster = clusternames[match(frequencies$cluster,
                                              clusternames$Cluster),
                                   paste0(celltype,"_clusters")]
frequencies$namedcluster = factor(frequencies$namedcluster, 
                                  levels = clusternames[
                                    clusternames[,paste0(celltype,"_order")]!="",
                                    paste0(celltype,"_order")])

cumulFreq = frequencies
cumulFreq$cumulProp = NA
for(i in unique(cumulFreq$sample)){
  cumulFreq$cumulProp[cumulFreq$sample %in% i] = rev(cumsum(rev(cumulFreq$Proportion[cumulFreq$sample %in% i])))
}
cumulFreq = rbind(cumulFreq,cumulFreq)
cumulFreq$xadj = rep(c(-1,1), c(nrow(frequencies),nrow(frequencies)))
cumulFreq$oldtime = cumulFreq$time
cumulFreq$time = as.numeric(gsub("D","",gsub("W","",as.character(cumulFreq$time))))


cleanfrequencies = frequencies[
  (substr(frequencies$time,1,1) == "W" & frequencies$model == "SCID") |
    (substr(frequencies$time,1,1) == "D" & frequencies$model == "Rag2"),]

clustercors = lapply(c("SCID","Rag2"), function(bg){
  corlist = lapply(unique(cleanfrequencies$namedcluster), function(i){
    clufreq = cleanfrequencies[cleanfrequencies$namedcluster==i  & 
                          cleanfrequencies$treatment == "untreated" & 
                          cleanfrequencies$model == bg,]

    clucor = cor(clufreq$Proportion,as.numeric(as.factor(as.character(clufreq$time))))
    clucor
    })
  names(corlist) = unique(cleanfrequencies$namedcluster)
  corlist
  })
names(clustercors) = c("SCID","Rag2")

averagecor = lapply(1:length(clustercors[[1]]), function(i){
  mean(c(clustercors[[1]][[i]], clustercors[[2]][[i]]))
})
names(averagecor) = names(clustercors[[1]])

cleanfrequencies$namedcluster = 
  factor(cleanfrequencies$namedcluster, 
         levels = names(averagecor)[order(unlist(averagecor), decreasing = TRUE)])
cumulFreq$namedcluster = 
  factor(cumulFreq$namedcluster,
         levels = names(averagecor)[order(unlist(averagecor), decreasing = TRUE)])

cleanfrequencies$sample = 
  factor(cleanfrequencies$sample, 
         levels = 
           unique(cleanfrequencies$sample[
             order(factor(cleanfrequencies$model, levels = c("SCID","Rag2")),
                   cleanfrequencies$treatment,
                   cleanfrequencies$time)]))
cleanfrequencies$samplenumber = as.numeric(cleanfrequencies$sample)

cumulFreq = cumulFreq[paste(cumulFreq$sample,
                cumulFreq$model,
                cumulFreq$oldtime) %in% 
            paste(cleanfrequencies$sample,
                  cleanfrequencies$model,
                  cleanfrequencies$time),]

cumulFreq$sample =
  factor(cumulFreq$sample, 
         levels = 
           unique(cumulFreq$sample[order(factor(cumulFreq$model, levels = c("SCID","Rag2")),
                                         cumulFreq$treatment,
                                         cumulFreq$time)]))
cumulFreq$samplenumber = as.numeric(cumulFreq$sample)


datasets = lapply(names(clustercors), function(bg){
  bgfreq = cleanfrequencies[cleanfrequencies$model == bg,]
  bglist = lapply(unique(bgfreq$treatment), function(i){
    trfr = bgfreq[bgfreq$treatment==i,]
    cumtrfr = cumulFreq[paste(cumulFreq$sample,
                                     cumulFreq$model,
                                     cumulFreq$oldtime) %in% 
                                 paste(trfr$sample,
                                       trfr$model,
                                       trfr$time),]
    list(trfr, cumtrfr)
  })
  names(bglist) = unique(bgfreq$treatment)
  bglist
})

barplot = ggplot(data = cleanfrequencies, 
       aes(x = samplenumber, y = Proportion, fill = namedcluster)) + 
  theme(panel.background = element_blank(), axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
            scale_fill_manual(values =
                                color_list[as.character(unique(sort(bgfreq$namedcluster)))]) + 
            scale_x_continuous(breaks = sort(unique(cleanfrequencies$samplenumber)),
                               labels = clean_sample_names(levels(cleanfrequencies$sample)))

for(ds in datasets){
  for(dst in ds){
    barplot = barplot + 
      geom_area(data = dst[[2]],
                alpha = 0.7, size = 0.3, stat = 'identity',
                aes(fill = namedcluster, group = namedcluster,
                    y = Proportion, x = samplenumber + xadj*0.3))}
}

barplot + geom_bar(stat = "identity", width = 0.6, color = "black")

```

# CITE-seq

```{r}

DefaultAssay(alldata) = "FB"

alldata <- NormalizeData(alldata)
alldata <- ScaleData(alldata)
  
```



```{r, fig.height = 2.2, fig.width = 2.3}

alldata@active.assay <- "FB"

dot = DotPlot(alldata, assay = "FB", features = featureorder, group.by = "Named_clusters") 
dot = dot + aes(x = id, y = features.plot) + scale_x_discrete(position = "top") +
  labs(y = "Antibodies",x = "Clusters") +
  scale_size(range = c(0.1,2.5), breaks = c(60,80,100)) +
  theme(legend.position = "bottom", legend.title = element_blank(),
        axis.text.x = element_text(size = 6, hjust = 0, angle = 90),
        axis.text.y = element_text(size = 6),
        legend.key.width = unit(0.3, "cm"),
        legend.key.height = unit(0.3,"cm"),
        legend.text = element_text(size = 6),
        axis.title.y = element_text(size = 8),
        axis.title.x = element_blank())

dot + theme(legend.position = "none")
plot_grid(get_legend(dot))

```


# NMF

```{r subsetting}


# Making subsets

## CD4

CD4_color_list = color_list
CD4_color_list[!(names(color_list) %in% clusterorder[1:4])] = NA
DimPlot(alldata, group.by = "Named_cluster") + 
  NoLegend() + NoAxes() +
  scale_color_manual(values = CD4_color_list)

CD4data = subset(alldata, idents = c(0,2,12,16))

CD4data = CD4data[,CD4data$RelTime > 0.1]

DefaultAssay(CD4data) = "RNA"
CD4data[["CCA"]] = NULL

CD4data <- ScaleData(CD4data)
CD4data <- RunPCA(CD4data, npcs = 30, verbose = FALSE)
CD4data <- RunUMAP(CD4data, dims = 1:30)

CD4data <- FindVariableFeatures(CD4data)
CD4data <- ScaleData(CD4data, features = VariableFeatures(CD4data))

## Neutrophils

neutrophil_color_list = color_list
neutrophil_color_list[!(names(color_list) %in% "Neutrophils")] = NA
DimPlot(alldata, group.by = "Named_cluster") + 
  NoLegend() + NoAxes() +
  scale_color_manual(values = neutrophil_color_list)

neutrophildata = subset(alldata, idents = 10)

DefaultAssay(neutrophildata) = "RNA"
neutrophildata[["CCA"]] = NULL

neutrophildata <- ScaleData(neutrophildata)
neutrophildata <- RunPCA(neutrophildata, npcs = 30, verbose = FALSE)
neutrophildata <- RunUMAP(neutrophildata, dims = 1:30)

neutrophildata <- FindVariableFeatures(neutrophildata)
neutrophildata <- ScaleData(neutrophildata, features = VariableFeatures(neutrophildata))


rm(alldata)
gc()

```


## CD4 T cells

```{r CD4NMF}

ntop = 40
downsample_n = 500

nmf <- RunNMF(object = CD4data, assay = "RNA", nfactors = ntop,
                n.cores = 1)

```

```{r}

for(i in 1:ntop){
  nmf <- AddMetaData(nmf, nmf@reductions$NMF@cell.embeddings[,paste0("factor_",i)],
                     paste0("nmf_",i))
}

nmf$ModelTreatmentTime = factor(nmf$ModelTreatmentTime,
      levels = unique(nmf$ModelTreatmentTime[order(factor(nmf$Model, levels = c("SCID", "Rag2")),
                                    factor(nmf$Treatment, levels = c("ut","il")),
                                    nmf$Time)]))

```

```{r}


breaksList = seq(-1, 1, by = 0.01)

ph = pheatmap(cor(nmf@meta.data[,grep("^nmf", colnames(nmf@meta.data))]), 
         color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(breaksList)),
         breaks = breaksList)


```


```{r}

Groups = SplitObject(nmf, split.by = "ModelTreatment")

```


```{r}


modelpvalues = getfactorpvalues(nmf, downsample = downsample_n, ntop)


pvaluegg = melt(modelpvalues, variable.name = "factor", value.name =  "p.value")
pvaluegg$factor = factor(pvaluegg$factor, levels = rev(paste0("nmf_",ph$tree_row$order)))
pvaluegg$depvar = factor(pvaluegg$depvar, levels = unique(pvaluegg$depvar))


lmp = ggplot(pvaluegg,
       aes(y = factor, x = depvar)) + 
  geom_tile(aes(fill = -log10(p.value))) + 
  scale_fill_viridis_c(option = "inferno", limits = c(0,NA)) + 
  scale_x_discrete(labels = function(x){gsub("SCID","",x)}) + 
  theme_classic() + 
  scale_y_discrete(labels = function(x){gsub("nmf_", paste0(factorprefix,"f"),x)}) + 
  theme(axis.text.x = element_text(size = 8, angle = 90, hjust = 1, vjust = 0.5),
        axis.text.y = element_text(size = 8),
        axis.title = element_blank(),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        legend.key.size = unit(8,"pt"))

```

```{r}

nmf_values = nmf@meta.data[,grep("^nmf_", colnames(nmf@meta.data))]

nmf_scaled = apply(nmf_values, 2, scale)

nmf_scaled_means_id = aggregate(nmf_scaled, by = list(nmf$ModelTreatmentTime), mean)
nmf_means_id = aggregate(nmf_values, by = list(nmf$ModelTreatmentTime), mean)
nmf_means_scaled_id = as.data.frame(apply(nmf_means_id[,-1], 2, scale))
nmf_means_scaled_id$Group.1 = nmf_means_id$Group.1

nmf_scaled_means_cluster = aggregate(nmf_scaled, by = list(nmf$named_cluster), mean)
nmf_means_cluster = aggregate(nmf_values, by = list(nmf$named_cluster), mean)
if(nrow(nmf_means_cluster)==1){
  nmf_means_scaled_cluster = as.data.frame(rbind(apply(nmf_means_cluster[,-1], 2, scale)))
  nmf_means_scaled_cluster$Group.1 = nmf_means_cluster$Group.1
  nmf_means_scaled_cluster[1,is.nan(as.numeric(nmf_means_scaled_cluster[1,]))] = 0
}else{
  nmf_means_scaled_cluster = as.data.frame(apply(nmf_means_cluster[,-1], 2, scale))
  nmf_means_scaled_cluster$Group.1 = nmf_means_cluster$Group.1
}


```



```{r}

ph_sample = pheatmap(nmf_means_scaled_id[,paste0("nmf_", 1:ntop)])

```

```{r, fig.width=3.4, fig.height=6}

ph_order = ph_sample$tree_col$order

txt = element_text(size = 8)

my_dotp_cl = ggplot(melt(nmf_means_scaled_cluster), 
                 aes(x = Group.1, 
                     y = factor(variable, 
                                levels = rev(paste0("nmf_",ph_order))))) +
  geom_tile(aes(fill = value)) + scale_fill_viridis_c() +
  scale_x_discrete(labels = function(x){gsub("nmf_", paste0(factorprefix,"f"),x)}) + 
  theme_classic() + 
  theme(axis.text = txt,
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.title = element_blank(),
        legend.text = txt,
        legend.title = txt) + 
  scale_size_continuous(limits = c(0,100))

my_dotp = ggplot(melt(nmf_means_scaled_id), 
                 aes(x = Group.1, 
                     y = factor(variable, 
                                levels = rev(paste0("nmf_",ph_order))))) +
  geom_tile(aes(fill = value)) + scale_fill_viridis_c() + 
  theme_classic() + 
  theme(axis.text = txt,
        axis.title = element_blank(),
        legend.text = txt,
        legend.title = txt) + 
  scale_x_discrete(labels = function(x){gsub("nmf_", paste0(factorprefix,"f"),x)}) + 
  scale_size_continuous(limits = c(0,100), range= c(0,3)) +
    labs(title = "Ordered by similarity") + 
  theme(plot.title = element_text(size = 10),
        axis.text.x = element_text(size = 8, angle = 90, hjust = 1, vjust = 0.5))



lmp = ggplot(pvaluegg,
       aes(y = factor(factor, levels = rev(paste0("nmf_",ph_order))), x = depvar)) + 
  geom_tile(aes(fill = -log10(p.value))) + 
  scale_fill_viridis_c(option = "inferno", limits = c(0,NA)) + 
  scale_x_discrete(labels = function(x){gsub("SCID","",x)}) + 
  theme_classic() + 
  scale_y_discrete(labels = function(x){gsub("nmf_", paste0(factorprefix,"f"),x)}) + 
  theme(axis.text.x = element_text(size = 8, angle = 90, hjust = 1, vjust = 0.5),
        axis.text.y = element_text(size = 8),
        axis.title = element_blank(),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        legend.key.size = unit(8,"pt"))


plot_grid(my_dotp_cl + 
            scale_fill_viridis_c(limits = scalerange) + 
            theme(axis.title = element_blank(),
                  legend.position = "bottom",
                      legend.key.size = unit(8,"pt")),
          my_dotp + 
            scale_fill_viridis_c(limits = scalerange) +
            theme(axis.title = element_blank(),
                  legend.position = "bottom",
                      legend.key.size = unit(8,"pt"),
                  plot.title = element_blank(),
                  axis.text.y = element_blank()
                  ),
          lmp + 
            theme(legend.position = "bottom",
                  axis.text.y = element_blank()
                  ) +
            labs(fill = ""), 
          align = "h", ncol = 3, rel_widths = c(2,2.7,0.8)
          )



```


```{r bytime, fig.height = 2.5, fig.width=3.5}

vlns = makeViolins(Groups,paste0("nmf_",1:ntop), relvln = c(4,1))

bars = lapply(1:ntop,function(i){
  barbreaks = seq(0,1,0.2)
  if(max(nmf@reductions$NMF@feature.loadings[,i])<0.2){
    barbreaks = seq(0,1,0.1)
  }
  x = FactorGeneLoadingPlot(nmf, factor = i, topn = 10) +
    scale_y_continuous(expand = expansion(mult = 0),
                       breaks = barbreaks) + 
    theme(axis.text.y = element_text(face = "italic",
                                     size = 6, color = "black"),
          axis.text.x = element_text(size = 6, color = "black"),
          axis.title.y = element_blank(),
          axis.title.x = element_text(size = 8,hjust = 0),
          plot.margin = margin(t = 1, b = 1, l = 3, unit = "pt")) +
    labs(y = "Weight")
  x$layers[[1]]$aes_params$colour = NA
  x 
  })

plots <- lapply(1:length(vlns), function(i){
  print(plot_grid(
    ggdraw() + 
      draw_label(
        "Factor score",
        hjust = 0.5, 
        size =  8, angle = 90
      ),
    plot_grid(ggdraw() + 
  draw_label(
    paste0(factorprefix, "f",i),
    fontface = 'bold',
    hjust = 0.5, 
    size =  10
  ),
  plot_grid(vlns[[i]] + theme(plot.title = element_blank()),
            bars[[i]], ncol = 2, rel_widths = c(4,1)), 
  rel_heights = c(1,6), ncol = 1), nrow = 1, rel_widths = c(1,10)))
})

```


# Distributions between clusters

```{r byident, fig.height = 10}

vpall <- VlnPlot(nmf, features = paste0("nmf_",1:ntop), 
        pt.size = 0.1, combine = FALSE)

plot_grid(plotlist = lapply(vpall, function(vp){
  vp$layers[[2]]$aes_params$shape = 16
  vp$layers[[1]]$aes_params$alpha = 0.4
  vpts = vp$layers[[2]]
  vp$layers[[2]] = vp$layers[[1]]
  vp$layers[[1]] <- ggrastr::rasterise(vpts, dpi = 300)
  vp +
    NoLegend()}))


```

# Enriched pathways

## GO BP

```{r}


top.genes <- lapply(1:ntop, function(i) {
  vals <- nmf@reductions$NMF@feature.loadings[, i]
  gg <- data.frame(x = 1:length(vals), vals = sort(vals, decreasing = T))
  gg$gene <- rownames(gg)
  gg <- subset(gg, vals > 0)
  gg$logvals <- log(gg$vals)
  gg$smooth_logvals <- smth.gaussian(x = gg$logvals, windowLength = 10)
  ggs <- gg[1:min(500, nrow(gg)), ]
  knee <- uik(x = ggs$x, y = ggs$smooth_logvals)
  ggs[1:knee, ]$gene
})

pathways <- lapply(seq_along(top.genes), function(i) {
  gset <- top.genes[[i]]
  df <- 
    gost(query = gset, organism = "mmusculus", 
         sources = "GO:BP", evcodes = TRUE)$result
  if (is.null(df)) return(NULL)
  df$factor <- paste0("factor_", i)
  return(df)
})
pathways <- do.call(rbind, pathways)
pathways$GeneRatio <- pathways$intersection_size/pathways$query_size

pathways.summarized <- pathways %>% 
  group_by(factor) %>%
  top_n(n = 20, wt = -log10(p_value))

pathways.summarized$term_name = shortenterms(pathways.summarized$term_name)

```

`r datatable(pathways.summarized, class = 'cell-border stripe compact')`


## KEGG

```{r}


  pathwaysKEGG <- lapply(seq_along(top.genes), function(i) {
    message(paste("KEGG analysis", i))
    gset <- top.genes[[i]]
    df <- gost(query = gset, organism = "mmusculus", sources = "KEGG")$result
    if (is.null(df)) return(NULL)
    df$factor <- paste0("factor_", i)
    return(df)
  })
  pathwaysKEGG <- do.call(rbind, pathwaysKEGG)


pathwaysKEGG$GeneRatio <- pathwaysKEGG$intersection_size/pathwaysKEGG$query_size
pathwaysKEGG.summarized <- pathwaysKEGG %>% 
  group_by(factor) %>%
  top_n(n = 20, wt = -log10(p_value))

pathwaysKEGG.summarized$term_name = shortenterms(pathwaysKEGG.summarized$term_name)

```

`r datatable(pathwaysKEGG.summarized, class = 'cell-border stripe compact')`


# Session info

```{r}

sessionInfo()

```
