---
title: "Preparation of data for T cell transfer single cell analysis"
editor_options:
  chunk_output_type: console
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    code_folding: hide
    toc: yes
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
params:
  dataset: 1
---

```{r Setup, echo = FALSE}
knitr::opts_chunk$set(fig.width  = 7,
                      results    = "hold",
                      message    = FALSE,
                      cache.lazy = FALSE,
                      warning    = FALSE)


```

```{r setup}

library(Seurat)
library(Matrix)
library(ggplot2)
library(cowplot)
library(DoubletFinder)
library(clustree)

library(future)
n_workers = 20
plan("multisession", workers = n_workers)
options(future.globals.maxSize = 1000 * 1024^3)
#options(future.globals.maxSize = 10 * 1024^3)
library(future.apply)

datadir = "../data/SingleCell/TcellTransfer/expression_matrices"

owncache = "../data/SingleCell/TcellTransfer/cache/"

violintheme = theme(axis.text = element_text(size = 8),
        plot.title = element_text(size = 10),
        axis.title = element_blank())

```


# Load data

```{r loading_matrices}

if(!file.exists(paste0(owncache,"/loading_matrices.RData"))){

loadObject = function(d){
  sample = sub("\\.h5","",sub(".*/","",d))
  expression_mat = Read10X_h5(filename = d)
  if(class(expression_mat)=="dgCMatrix"){
    data_obj = CreateSeuratObject(expression_mat, sample)
  }else{
      data_obj = CreateSeuratObject(expression_mat$`Gene Expression`, sample)
      data_obj[["FB"]] = CreateAssayObject(expression_mat$`Antibody Capture`)
      data_obj@active.assay = "RNA"
  }
  data_obj
}

system.time(
  {
    datafiles = list.files(datadir, full.names = TRUE)
    all_matrices = future_lapply(datafiles, function(d){
      message(d)
      sample = sub("\\.h5","",sub(".*/","",d))
      message(sample)
  loadObject(d)
})
names(all_matrices) = sub("\\.h5","",sub(".*/","",datafiles))
gc()})

alldata = merge(x = all_matrices[[1]],
                y = all_matrices[-1],
                add.cell.ids = names(all_matrices))

rm(all_matrices)
rm(loadObject)
gc()

if(file.exists(owncache)){
  save(alldata, paste0(owncache,"/loading_matrices.RData"))
}

}else{
  l = load(paste0(owncache,"/loading_matrices.RData"))
}


```

```{r addmetadata}

metadata = read.csv2(paste0(datadir,"/../metadata.csv"))
colnames(metadata) = gsub("X","",gsub("\\.","",colnames(metadata)))

alldata = AddMetaData(alldata, 
                      factor(metadata[match(alldata$orig.ident, metadata$libraryname),"genotype"],
                             levels = c("BALB/c SCID", "Rag2 KO"),
                             labels = c("SCID","Rag2")),
                      "Model")
alldata = AddMetaData(alldata, metadata[match(alldata$orig.ident, metadata$libraryname),"time"],
                      "Time")
alldata = AddMetaData(alldata, 
                      metadata[match(alldata$orig.ident, metadata$libraryname),"celltype"],
                      "Type")
alldata = AddMetaData(alldata, 
                      factor(metadata[match(alldata$orig.ident, metadata$libraryname),"antiIL12p40"],
                             levels = c("no", "yes"), labels = c("ut","il")),
                      "Treatment")

alldata <- AddMetaData(alldata,
                       sapply(alldata$Time,
                              function(x){if(substr(x,1,1) == "D"){
                                return(as.numeric(gsub("D","",x))/60)
                                }else{
                                  return(as.numeric(gsub("W","",x))/7)}}), "RelTime")


alldata <- PercentageFeatureSet(alldata, "^mt-", col.name = "percent_mito")

alldata <- PercentageFeatureSet(alldata, "^Rp[sl]", col.name = "percent_ribo")

```

# Plot QC vars

Number of features and counts and percentage of mitochondrial and ribosomal RNA are plotted below.

```{r qc, fig.height=10}

qc_feats <- c("nFeature_RNA", "nCount_RNA", "percent_mito", "percent_ribo")
VlnPlot(alldata, group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) &
  violintheme
  

```


# Filter cells and genes

The same QC as above are plotted below, only including cells and genes that pass filtering limits (gene limit: > 3 copies expressed, cell filter: > 200 unique genes). The top expressed genes are also shown. Mitochondrial and ribosomal genes are filtered out of the data, as are the highly expressed genes Malat1 and Gm42418. 

```{r filtering}


alldata <- subset(alldata, features = rownames(alldata)[Matrix::rowSums(alldata) > 3], 
                  cells = intersect(WhichCells(alldata, expression = nFeature_RNA > 200),
                                    intersect(WhichCells(alldata, expression = percent_mito < 10),
                                    WhichCells(alldata, expression = percent_ribo > 5))))

gc()

```

```{r type_qc_violins, fig.height = 10}

VlnPlot(alldata, group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) &
    NoLegend() & violintheme


VlnPlot(subset(alldata,
               idents = unique(alldata@meta.data$orig.ident[alldata@meta.data$orig.ident == "Immune"])),
        group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) + 
    NoLegend() & violintheme
VlnPlot(subset(alldata,
               idents = unique(alldata@meta.data$orig.ident[alldata@meta.data$orig.ident == "Stroma"])),
        group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) + 
    NoLegend() & violintheme
VlnPlot(subset(alldata,
               idents = unique(alldata@meta.data$orig.ident[alldata@meta.data$orig.ident == "Epithelial"])),
        group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) + 
    NoLegend() & violintheme

```

```{r topexpressed, fig.height = 10, cache = TRUE}

par(mar = c(4, 8, 2, 1))
C <- subset(alldata, downsample = 1000, seed = 42)@assays$RNA@counts
C <- Matrix::t(Matrix::t(C)/Matrix::colSums(C)) * 100
#nC = ncol(C)

most_expressed <- order(apply(C, 1, median), decreasing = T)[20:1]
 
boxplot(as.matrix(t(C[most_expressed, ])), cex = 0.1, las = 1, xlab = "% total count per cell",
     col = (scales::hue_pal())(20)[20:1], horizontal = TRUE)


rm(C)
rm(most_expressed)



# Filter MALAT1, Gm42418, mitochondrial genes and ribosomal genes
alldata <- alldata[!grepl("Malat1", rownames(alldata)) & 
                     !grepl("Gm42418", rownames(alldata)) & 
                     !grepl("^mt-", rownames(alldata)) & ! 
                     grepl('^Rp[sl]', rownames(alldata)), ]

```



knitr::kable(table(alldata$orig.ident))


```{r}

message(ls())

```


# Cell cycle scores

The S-phase and G2M-phase scores are calculated based on the cell cycle gene lists defined in the Seurat package (translated to mouse genes originally using biomaRt).

```{r cellcyclescores}

if(!file.exists(paste0(owncache,"/cellcyclescores.RData"))){

  g2mFeat = c("Smc4","Dlgap5","Ctcf","Cdc25c","Gtse1","Kif20b",
              "Ttk","Tacc3","Top2a","Tpx2","Cdca3","Ncapd2",
              "Cks2","Anp32e","G2e3","Lbr","Cdca2","Cks1brt",
              "Ckap2","Hmgb2","Kif11","Nek2","Cenpe","Hjurp",
              "Ect2","Aurkb","Cks1b","Kif23","Nuf2","Hmmr",
              "Cdca8","Psrc1","Anln","Cdc20","Birc5","Ndc80",
              "Rangap1","Ckap5","Kif2c","Cenpf","Nusap1","Cenpa",
              "Aurka","Ube2c","Ckap2l","Mki67","Tubb4b","Bub1",
              "Ccnb2")
  sFeat = c("Msh2","Exo1","Mcm4","Rrm2","Mcm2","Chaf1b","Gmnn",
            "Cdc45","Slbp","Ubr7","Cdc6",
            "Rad51ap1","Rpa2","Hells","Fen1","Gins2","Uhrf1",
            "Mcm6","Ung","Dscc1","Usp1","Clspn",
            "Cdca7","Pola1","Nasp","Dtl","Mcm5","Wdr76","Prim1",
            "Casp8ap2","Tipin","Blm","Rrm1",
            "Brip1","Rad51","Tyms","E2f8","Rfc2","Ccne2","Pcna")




# Before running CellCycleScoring the data need to be normalized and
# logtransformed.

plan("sequential")
alldata = NormalizeData(alldata)

alldata <- CellCycleScoring(object = alldata, g2m.features = g2mFeat, 
    s.features = sFeat)

rm(g2mFeat)
rm(sFeat)
gc()


if(file.exists(owncache)){
  save(alldata, paste0(owncache,"/cellcyclescores.RData"))
}

}else{
  l = load(paste0(owncache,"/cellcyclescores.RData"))
}


```


```{r}

VlnPlot(alldata, features = c("S.Score", "G2M.Score"), group.by = "orig.ident", 
    ncol = 3, pt.size = 0.1)

VlnPlot(alldata, features = c("S.Score", "G2M.Score"), group.by = "orig.ident", 
    ncol = 3, pt.size = 0)


```


# Find Doublets

Doublets are identified and excluded using the DoubletFinder package.


```{r df_prep, cache = TRUE}

plan("multisession", workers = n_workers)

  alldata = FindVariableFeatures(alldata, verbose = F)
  alldata = ScaleData(alldata, vars.to.regress = c("nFeature_RNA", "percent_mito"), 
      verbose = F)
  alldata = RunPCA(alldata, verbose = F, npcs = 20)
  alldata = RunUMAP(alldata, dims = 1:10, verbose = F)
  

```

```{r doublets, cache = TRUE}
  
plan("multisession", workers = n_workers)

sweeplist = list()

singletList = future_lapply(unique(alldata$orig.ident), function(i){
  message(i)
  subdata=subset(alldata, idents = i)
  sweep.res <- paramSweep_v3(subdata)#, num.cores = 3)
  sweep.stats <- summarizeSweep(sweep.res, GT = FALSE)
  bcmvn = find.pK(sweep.stats)

  nExp <- round(ncol(subdata) * ifelse(ncol(subdata)<1200,0.08,ifelse(ncol(subdata)<2200,0.16,0.24)))  # the expected number of doublets was overestimated in the published analysis, but correcting this does not lead to a significant difference in the prediction
  mypK = as.numeric(as.character(bcmvn$pK[which.max(bcmvn$BCmetric)]))
  subdata <- doubletFinder_v3(subdata, pN = 0.25, pK = mypK, nExp = nExp, PCs = 1:10)
  DF.name = colnames(subdata@meta.data)[grepl("DF.classification", colnames(subdata@meta.data))]
  sl = subdata@meta.data[, DF.name]
  names(sl) = colnames(subdata) 
})
names(singletList) = unique(alldata$orig.ident)

alldata = AddMetaData(alldata, unlist(singletList)[
  match(colnames(alldata), 
        gsub(".*\\.","",names(unlist(singletList))))], "DF")

cowplot::plot_grid(ncol = 2, DimPlot(alldata, group.by = "orig.ident") + NoAxes(), 
    DimPlot(alldata, group.by = "DF") + NoAxes())


```


```{r df_dimplot, fig.height = 10}

DimPlot(alldata, group.by = "DF", split.by = "orig.ident", ncol = 6) + NoAxes()

```


```{r df_violins}

VlnPlot(alldata, features = "nFeature_RNA", group.by = "DF",pt.size = 0.1)
VlnPlot(alldata, features = "nFeature_RNA", split.by = "DF", group.by = "orig.ident", pt.size = 0.1)

alldata = alldata[, alldata@meta.data[, "DF"] == "Singlet"]

```



```{r cellcounts}

table(alldata$orig.ident)

cellfreq = as.data.frame(table(alldata$orig.ident), stringsAsFactors = FALSE)
cellfreq$Time = alldata$Time[match(cellfreq$Var1, alldata$orig.ident)]
cellfreq$Type = alldata$Type[match(cellfreq$Var1, alldata$orig.ident)]
cellfreq$Treatment = alldata$Treatment[match(cellfreq$Var1, alldata$orig.ident)]
ggplot(cellfreq, aes(x = paste(Week, Treatment), y = Freq, color = Type)) + geom_point() +
  theme(panel.background = element_blank(), legend.key = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Week and treatment", y = "Number of cells")

```


```{r violins}

VlnPlot(alldata, group.by = "orig.ident", features = qc_feats, pt.size = 0.1, ncol = 2) +
    NoLegend()
VlnPlot(alldata, group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) +
    NoLegend()

```


```{r typeviolins}

VlnPlot(alldata[,alldata$Type == "Stroma"],
        group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) +
    NoLegend()
VlnPlot(alldata[,alldata$Type == "Epithelial"],
        group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) +
    NoLegend()
VlnPlot(alldata[,alldata$Type == "Immune"],
        group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) +
    NoLegend()



```




```{r firstnormalize, cache = TRUE}

alldata = NormalizeData(alldata)

suppressWarnings(suppressMessages(alldata <- FindVariableFeatures(alldata, selection.method = "vst", 
    nfeatures = 2000, verbose = FALSE, assay = "RNA")))
alldata <- ScaleData(alldata, vars.to.regress = c("percent_mito", "nFeature_RNA"),     assay = "RNA")
alldata <- RunPCA(alldata, npcs = 50, verbose = F, reduction.name = "pca_raw")
alldata <- RunTSNE(alldata, reduction = "pca_raw", dims = 1:30, perplexity = 30, max_iter = 1000, 
    theta = 0.5, eta = 200, num_threads = 0, reduction.name = "tsne_raw")
alldata <- RunUMAP(alldata, reduction = "pca_raw", dims = 1:30, n.components = 2, n.neighbors = 30, 
    n.epochs = 200, min.dist = 0.3, learning.rate = 1, spread = 1, reduction.name = "umap_raw")

```


```{r topgenes}

top20 <- head(VariableFeatures(alldata), 20)

LabelPoints(plot = VariableFeaturePlot(alldata), points = top20, repel = TRUE)

```


# PCA

```{r firstpca}

plot_grid(ncol = 3, 
          DimPlot(alldata, reduction = "pca_raw", group.by = "orig.ident", 
                  dims = 1:2) + NoLegend(),
          DimPlot(alldata, reduction = "pca_raw", group.by = "orig.ident", 
                  dims = 3:4) + NoLegend(),
          DimPlot(alldata, reduction = "pca_raw", group.by = "orig.ident", 
                  dims = 5:6) + NoLegend())


VizDimLoadings(alldata, dims = 1:5, reduction = "pca_raw", ncol = 5, balanced = T)

ElbowPlot(alldata, reduction = "pca_raw", ndims = 50)

```

# t-SNE

```{r firsttsne}

plot_grid(ncol = 2, DimPlot(alldata, reduction = "tsne_raw", group.by = "orig.ident") + 
            NoAxes() + NoLegend(), 
          DimPlot(alldata, reduction = "tsne_raw", group.by = "RelTime") + 
            NoAxes() + NoLegend(), 
          DimPlot(alldata, reduction = "tsne_raw", group.by = "Type") + 
            NoAxes() + NoLegend())

```


# UMAP

```{r firstumap}

plot_grid(ncol = 3, DimPlot(alldata, reduction = "umap_raw", group.by = "orig.ident") +
            NoAxes() + NoLegend(),
    FeaturePlot(alldata, "RelTime", reduction = "umap_raw") + NoAxes() + NoLegend(),
    DimPlot(alldata, group.by = "Type") + NoAxes() + NoLegend()) 

```


# Integrate data

```{r getanchors, cache = TRUE}

alldata <- AddMetaData(alldata, gsub("_I_","_",gsub("_S_","_",
                            gsub("_E_","_",gsub("_Str","",gsub("_Imm","",
                            gsub("_Epi","",alldata$orig.ident)))))), 
                     "source")
data.list <- SplitObject(alldata, split.by = "source")
    
data.list = lapply(data.list, function(x){
    x <- NormalizeData(x, verbose = FALSE)
    x <- FindVariableFeatures(x, selection.method = "vst",
        nfeatures = 2000, verbose = FALSE)
    x
})
    
features <- SelectIntegrationFeatures(object.list = data.list)
data.list <- lapply(X = data.list, FUN = function(x) {
  x <- ScaleData(x, features = features, verbose = FALSE)
  x <- RunPCA(x, features = features, verbose = FALSE)
})

data.anchors <- FindIntegrationAnchors(object.list = data.list, dims = 1:30, 
    reduction = "rpca", anchor.features = features, 
    reference = match(c("G_W0","G_W7","G_D0_ut","G_D60_ut",
                        "G_W7_aIL12","G_D30_il","G_D40_il",
                        "G_D60_il","G_W3_aIL12","G_W5_aIL12"),
                      names(data.list)))

rm(data.list)
gc()

```

```{r fullintegration, cache = TRUE}

alldata <- IntegrateData(anchorset = data.anchors, dims = 1:30, new.assay.name = "RPCA")

rm(data.anchors)
gc()
  
```


```{r normalizeintegrated, cache = TRUE}

alldata <- ScaleData(alldata)
alldata <- RunPCA(alldata, npcs = 30)
alldata <- RunUMAP(alldata, dims = 1:30)

```


```{r}

plot_grid(ncol = 2,
  DimPlot(alldata, reduction = "pca_raw", group.by = "orig.ident", pt.size = 0.01) +
    NoAxes() + ggtitle("PCA raw_data") + NoLegend(),
  DimPlot(alldata, reduction = "umap_raw", group.by = "orig.ident", pt.size = 0.01) +
    NoAxes() + ggtitle("UMAP raw_data") + NoLegend(),
  
  DimPlot(alldata, reduction = "pca_raw", group.by = "orig.ident") +
    NoAxes() + ggtitle("PCA integrated") + NoLegend(),
  DimPlot(alldata, reduction = "umap_raw", group.by = "orig.ident") +
    NoAxes() + ggtitle("UMAP integrated") + NoLegend()
)

DimPlot(alldata, reduction = "umap", split.by = "orig.ident")+NoAxes() + NoLegend()

DimPlot(alldata, reduction = "umap", group.by = "Model", shuffle = TRUE) + NoAxes()


```

# Plot QC in UMAP

```{r, fig.height = 10}

FeaturePlot(alldata, reduction = "umap", 
            features = qc_feats, 
    order = T, slot = "data", combine = T)

```



# Plot genes of interest

```{r}

DefaultAssay(alldata) = "RNA"

FeaturePlot(alldata, reduction = "umap", features = c("Mki67","Pcna"), 
    order = T, slot = "data", combine = T)

FeaturePlot(alldata, reduction = "umap", features = c("Ptprc","Epcam"), 
    order = T, slot = "data", combine = T)

DefaultAssay(alldata) = "RPCA"

```


# Clustering to identify sorting contaminations


```{r}

getseedcells = function(seeds, meta){
  big = sum(sapply(table(meta$orig.ident), function(x){return(min(x,5000))}))

fullnumbers = sort(table(meta$orig.ident))
proportions = fullnumbers/nrow(meta)
numbers = round(proportions*big)
numbers[numbers<1000] = fullnumbers[numbers<1000]
names(numbers) = names(proportions)


seeddf = data.frame(matrix(nrow=nrow(meta), ncol = length(seeds), 0))
colnames(seeddf) = paste0("seed",seeds)
rownames(seeddf) = rownames(meta)

for(thisseed in seeds){
  for(i in unique(meta$orig.ident)){
    set.seed(thisseed)
    take = numbers[i]
    cells = sample(rownames(meta)[meta$orig.ident == i], size = take)
    seeddf[cells,paste0("seed",thisseed)] = 1
  }
}

seeddf$included = apply(seeddf, 1, sum)
allcols = colnames(seeddf)[grep("seed",colnames(seeddf))]

set.seed(42)

currcol = 1
while(min(seeddf$included)<5){
  if(currcol>length(allcols)){
    currcol = 1
  }
  toofewall = rownames(seeddf)[seeddf$included<5 & seeddf[,currcol] == 0]
  toomanyall = rownames(seeddf)[seeddf$included>7 & seeddf$included<length(allcols) & seeddf[,currcol] == 1]
  least = min(length(toofewall), length(toomanyall))
  toofew = sample(toofewall, size = min(least,1000))
  toomany = sample(toomanyall, size = min(least,1000))
  seeddf[toofew,currcol] = 1
  seeddf[toomany,currcol] = 0
  seeddf[c(toofew, toomany),"included"] = apply(seeddf[c(toofew, toomany),allcols], 1, sum)
  currcol = currcol + 1
  if(sum(seeddf[,currcol]) != 241406){print(sum(seeddf[,currcol]))}
}
}

combineclusters = function(default =, clusters = , meta = ){
  
clusterdf = clusters[[1]]
  
for(i in 2:length(clusters)){
  clusterdf = merge(clusterdf, clusters[[i]], by = "Cell", all = TRUE)
}
  
meta = merge(meta, clusterdf, by.x = "row.names", by.y = "Cell")
rownames(meta) = meta[,1]
meta = meta[,-1]
  
clustercols = grep("seed",colnames(meta))


for(i in colnames(meta)[grep("cluster_seed",colnames(meta))]){
  meta[,i] = factor(meta[,i],levels = c(sort(unique(meta[,i])),9999))
  meta[is.na(meta[,i]),i] = 9999
  
}

translators = list()
for(j in grep("cluster_seed", colnames(meta))){
  for(i in grep("cluster_seed", colnames(meta))){
    if(i!=j){
      ft = table(old = meta[,j], new = meta[,i])
      translators[[colnames(meta)[j]]][[colnames(meta)[i]]] = 
        data.frame(new = c(colnames(ft)[-ncol(ft)],9999), 
                   newnew = c(rownames(ft)[
                     apply(ft[-nrow(ft),-ncol(ft)],2,which.max)],9999))
      translators[[colnames(meta)[j]]][[colnames(meta)[i]]]$prop = 
        c(apply(ft[-nrow(ft),-ncol(ft)],2,function(x){max(x)/sum(x)}),NA)
    }
    
  }
}

tps = matrix(ncol = length(grep("cluster_seed", colnames(meta))), nrow = nrow(meta))
rownames(tps) = rownames(meta)
colnames(tps) = colnames(meta)[grep("cluster_seed", colnames(meta))]

for(i in grep("cluster_seed", colnames(meta))){
  iname = colnames(meta)[i]
  if(iname!=default){
    tps[,iname] = as.numeric(as.character(translators[[default]][[iname]]$newnew[match(
      as.character(meta[,i]),as.character(translators[[default]][[iname]]$new)
    )]))
  }else{
    tps[,iname] = as.numeric(as.character(meta[,i]))
  }
}

tps[tps == 9999] = NA

toptranslation = apply(tps, 1, function(x){
  if(sum(!is.na(x)>0)){names(sort(table(x), decreasing = TRUE))[1]}else{NA}})

meta$RPCA_cluster = as.numeric(unlist(toptranslation))

if(length(stillNA = which(is.na(meta[,"RPCA_cluster"])))>0){stop("there are still non-assigned cells")}

meta

}

```

```{r, cache = TRUE}


meta = alldata@meta.data

seeddf = getseedcells(seeds = c(123,42,666,789,345,636,234,542628,57362,
          059482,7665,2342,82349,953020,03248,1284,
          4135,659,2340,6543,602), meta = meta)


res = 0.5

clusters = lapply(gsub("seed","",colnames(seeddf)), function(seed){
  seedcells = rownames(seeddf)[seeddf[,match(paste0("seed",seed), colnames(seeddf))]==1]
  seeddata = subset(alldata, cells = seedcells)
  
  seeddata <- FindNeighbors(seeddata, dims = 1:30, k.param = 350, prune.SNN = 1/15)
  seeddata <- FindClusters(seeddata, graph.name = paste0("RPCA_snn"),
                            resolution = res, algorithm = 1)
  
  x = seddata@meta.data[,c("X","seurat_clusters")]
  colnames(x) = c("Cell",paste0("cluster_",seed))
  x
})
names(clusters) = gsub("seed","",colnames(seeddf))

rm(seeddf)


meta = combineclusters(default = "cluster_seed57362",
                       clusters = clusters,
                       meta = meta)

if(rownames(meta) != colnames(alldata)){
  meta = meta[colnames(alldata),]
}

alldata = AddMetaData(alldata, meta$RPCA_cluster,"RPCA_cluster")

alldata = SetIdent(alldata, value = "RPCA_cluster")

DefaultAssay(alldata) <- "RNA"

rm(meta, clusters, getseedcells, combineclusters)

```


```{r}

DimPlot(alldata, label = TRUE) + NoAxes() + NoLegend()

```

```{r, fig.height = 3}

DotPlot(alldata, features = c("Epcam","Ptprc")) + coord_flip()

```


```{r}

ggplot(alldata@meta.data, aes(fill = Type, x = RPCA_cluster)) + geom_bar(position = "fill")

ggplot(alldata@meta.data, aes(fill = Model, x = RPCA_cluster)) + geom_bar(position = "fill")

ggplot(alldata@meta.data, aes(fill = as.factor(RelTime), x = RPCA_cluster)) + 
  geom_bar(position = "fill") + 
  scale_fill_manual(values = colorRampPalette(
    c("grey","blue"))(length(unique(alldata@meta.data$RelTime))))

```


```{r}

VlnPlot(alldata, group.by = "RPCA_cluster", features = qc_feats, pt.size = 0.1, ncol = 2)
VlnPlot(alldata, group.by = "RPCA_cluster", features = qc_feats, pt.size = 0, ncol = 2)

```



# Data cleaning

```{r cleandata, cache = TRUE}

immuneclusters = c(0,1,3,8,10,12,14,16,21)
stromaclusters = c(4,7,9,11,13,15,17,18,20)
epiclusters = c(2,5,6,19)

sel.clust = "RPCA_cluster"

alldata <- SetIdent(alldata, value = "Type")

immunetype = WhichCells(alldata, ident = "Immune")
stromatype = WhichCells(alldata, ident = "Stroma")
epitype = WhichCells(alldata, ident = "Epithelial")

alldata <- SetIdent(alldata, value = sel.clust)
alldata$ident_clust = paste0(alldata@active.ident,"_",alldata$orig.ident)

immunecells = WhichCells(alldata, ident = immuneclusters, cells = immunetype)
stromacells = WhichCells(alldata, ident = stromaclusters, cells = stromatype)
epicells = WhichCells(alldata, ident = epiclusters, cells = epitype)

alldata = subset(alldata, cells = c(immunecells,stromacells, epicells))

cleandata = alldata

rm(alldata)

immunedata = subset(cleandata, cells = immunecells)
immunedata@active.assay = "RNA"
immunedata[["RPCA"]] = NULL
saveRDS(immunedata, "../data/SingleCell/TcellTransfer/R_objects/immune_data.rds")
rm(immunedata)

stromadata = subset(cleandata, cells = stromacells)
stromadata@active.assay = "RNA"
stromadata[["RPCA"]] = NULL
saveRDS(stromadata, "../data/SingleCell/TcellTransfer/R_objects/stroma_data.rds")
rm(stromadata)

epidata = subset(cleandata, cells = epicells)
epidata@active.assay = "RNA"
epidata[["RPCA"]] = NULL
saveRDS(epidata, "../data/SingleCell/TcellTransfer/R_objects/epithelial_data.rds")
rm(epidata)

cleandata <- ScaleData(cleandata)
cleandata <- RunPCA(cleandata, npcs = 30)
cleandata <- RunUMAP(cleandata, dims = 1:30)

```


```{r dimplots}

plot_grid(ncol = 2, 
          DimPlot(cleandata, reduction = "umap", group.by = "RPCA_cluster", 
                  label = TRUE) + 
            ggtitle("louvain_0.5") + NoLegend())

```


```{r}

TypeClusters = list(Immune = immuneclusters, Stroma = stromaclusters, Epithelial = epiclusters)

ggplotColours <- function(n = 6, h = c(0, 360) + 15){
  if ((diff(h) %% 360) < 1) h[2] <- h[2] - 360/n
  hcl(h = (seq(h[1], h[2], length = n)), c = 100, l = 65)
}

color_list <- ggplotColours(n=length(unique(cleandata@active.ident)))
names(color_list) = sort(as.numeric(as.character(unique(cleandata@active.ident))))

DimPlot(cleandata, reduction = "umap", split.by = "Type", label = TRUE)+NoAxes()+NoLegend() +
  scale_color_manual(values = color_list)

DimPlot(cleandata, reduction = "umap", split.by = "Model", label = TRUE)+NoAxes()+NoLegend() +
  scale_color_manual(values = color_list)

```


```{r}

sampleorder = unique(cleandata$orig.ident[
  order(cleandata$Model,cleandata$Treatment,cleandata$Time, 
        decreasing = c(FALSE, TRUE, FALSE))])

plot_grid(plotlist = lapply(unique(cleandata$Type), function(type){
  typemeta = cleandata@meta.data[cleandata@meta.data$Type %in% type,]
  typemeta$Treatment = factor(typemeta$Treatment, levels = c("untreated","aIL12"))
  typemeta = typemeta[order(typemeta$Model,typemeta$Treatment,typemeta$Time),]
  typemeta$orig.ident = factor(typemeta$orig.ident, levels = unique(typemeta$orig.ident))
  typemeta$RPCA_cluster = factor(typemeta$RPCA_cluster, levels = 
                                   sort(as.numeric(unique(typemeta$RPCA_cluster))))
  ggplot(typemeta, aes(x = orig.ident, fill = RPCA_cluster)) +
    geom_bar(position = "fill") + 
    labs(fill = "Cluster") + 
  scale_fill_manual(values = color_list[as.character(TypeClusters[[type]])]) + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1), 
          legend.key = element_blank(), 
          legend.position = "bottom", panel.background = element_blank(),
          panel.grid = element_blank())
}), ncol = 3)

```


```{r umaps, fig.height = 9}

source("stylesource.R")

allplots = list(
          DimPlot(cleandata, reduction = "umap", group.by = "Model", shuffle = TRUE) + NoAxes() + 
             scale_color_manual(values = modelcolors),
          FeaturePlot(cleandata, cells = sample(Cells(cleandata)),
                      reduction = "umap", features = "RelTime") + 
             NoAxes() + labs(title = "Time point") + timecolors,
          DimPlot(cleandata, reduction = "umap", group.by = "Treatment", shuffle = TRUE) + NoAxes() +
             scale_color_manual(values = treatmentcolors),
          DimPlot(cleandata, reduction = "umap", group.by = "Type", shuffle = TRUE) +
            NoAxes() + labs(title = "Sort fraction") +
             scale_color_manual(values = typecolors))

plot_grid(ncol = 2, plotlist = allplots)

plot_grid(ncol = 2, plotlist = lapply(allplots, function(x){x + NoLegend()}))

```


# Session info

```{r}

sessionInfo()

```
