---
title: "Preparation of data for T cell transfer single cell analysis"
editor_options:
  chunk_output_type: console
date: "27 Februrary, 2024"
output:
  bookdown::html_document2:
    code_folding: hide
    toc: yes
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
params:
  dataset: 1
---

```{r Setup, echo = FALSE}
knitr::opts_chunk$set(fig.width  = 7,
                      results    = "hold",
                      message    = FALSE,
                      cache.lazy = FALSE,
                      warning    = FALSE)


```

```{r setup}

library(Seurat)
library(Matrix)
library(ggplot2)
library(cowplot)
library(DoubletFinder)
library(clustree)

library(future)
n_workers = 10
plan("multisession", workers = n_workers, gc = TRUE)
options(future.globals.maxSize = 100 * 1024^3)
#options(future.globals.maxSize = 10 * 1024^3)
library(future.apply)

datadir = "../data/SingleCell/TcellTransfer/expression_matrices"

owncache = "../data/SingleCell/TcellTransfer/cache/"

violintheme = theme(axis.text = element_text(size = 8),
        plot.title = element_text(size = 8),
        legend.key.size = unit(8, "pt"),
        axis.title = element_blank())
dimtheme = theme(legend.text = element_text(size = 8),
        plot.title = element_text(size = 8),
        legend.key.size = unit(8, "pt"))

metadata = read.csv2(paste0(datadir,"/../metadata.csv"))
sampleorder = metadata[substr(metadata[,1],1,1)=="G",1]

qc_feats <- c("nFeature_RNA", "nCount_RNA", "percent_mito", "percent_ribo")

```

```{r, include = FALSE}

if(file.exists(owncache)){
  message(paste0("Intermediate data files will be loaded from and saved to ",owncache))
}else{
  message("Intermediate data files will not be saved or loaded")
}

```

# Load data

```{r loading_matrices}

if(!file.exists(paste0(owncache,"/loading_matrices.RData"))){

plan("multisession", workers = 16, gc = TRUE)

loadObject = function(d){
  sample = sub("\\.h5","",sub(".*/","",d))
  expression_mat = Read10X_h5(filename = d)
  if(class(expression_mat)=="dgCMatrix"){
    data_obj = CreateSeuratObject(expression_mat, sample)
  }else{
      data_obj = CreateSeuratObject(expression_mat$`Gene Expression`, sample)
      data_obj[["FB"]] = CreateAssayObject(expression_mat$`Antibody Capture`)
      data_obj@active.assay = "RNA"
  }
  data_obj
}

system.time(
  {
    datafiles = list.files(datadir, full.names = TRUE)
    all_matrices = future_lapply(datafiles, function(d){
      message(d)
      sample = sub("\\.h5","",sub(".*/","",d))
      message(sample)
  loadObject(d)
})
names(all_matrices) = sub("\\.h5","",sub(".*/","",datafiles))
gc()})

alldata = merge(x = all_matrices[[sampleorder[1]]],
                y = all_matrices[sampleorder[-1]],
                add.cell.ids = sampleorder)

  immunesamples = sampleorder[grep("_I",sampleorder)]
  immune_matrices = lapply(all_matrices[immunesamples], 
                         function(x){
                           x@active.assay = "FB"
                           x[["RNA"]] = NULL
                           x
                         })

  alldata_FB = merge(immune_matrices[[1]],
                   immune_matrices[-1],
                   add.cell.ids = immunesamples)
  
  if(file.exists(owncache)){
    save(alldata,alldata_FB, file = paste0(owncache,"/loading_matrices.RData"))
  }
  
  rm(all_matrices)
  rm(loadObject)
  gc()

}else{
  load(paste0(owncache,"/loading_matrices.RData"))
}


```

```{r addmetadata}


colnames(metadata) = gsub("X","",gsub("\\.","",colnames(metadata)))

alldata = AddMetaData(alldata, 
                      factor(metadata[match(alldata$orig.ident, metadata$libraryname),"genotype"],
                             levels = c("BALB/c SCID", "Rag2 KO"),
                             labels = c("SCID","Rag2")),
                      "Model")
alldata = AddMetaData(alldata, metadata[match(alldata$orig.ident, metadata$libraryname),"time"],
                      "Time")
alldata = AddMetaData(alldata, 
                      metadata[match(alldata$orig.ident, metadata$libraryname),"celltype"],
                      "Type")
alldata = AddMetaData(alldata, 
                      factor(metadata[match(alldata$orig.ident, metadata$libraryname),"antiIL12p40"],
                             levels = c("no", "yes"), labels = c("ut","il")),
                      "Treatment")

alldata <- AddMetaData(alldata,
                       sapply(alldata$Time,
                              function(x){if(substr(x,1,1) == "D"){
                                return(as.numeric(gsub("D","",x))/60)
                                }else{
                                  return(as.numeric(gsub("W","",x))/7)}}), "RelTime")


alldata <- PercentageFeatureSet(alldata, "^mt-", col.name = "percent_mito")

alldata <- PercentageFeatureSet(alldata, "^Rp[sl]", col.name = "percent_ribo")

```

# Plot QC vars

Number of features and counts and percentage of mitochondrial and ribosomal RNA are plotted below.


```{r qc, fig.height=8, fig.width=8}

VlnPlot(alldata, group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 1) &
  violintheme
  

```


# Filter cells and genes

The same QC as above are plotted below, only including cells and genes that pass filtering limits (gene limit: > 3 copies expressed, cell filter: > 200 unique genes, mitochondrial RNA < 10%, ribosomal RNA > 5%). The top expressed genes are also shown. Mitochondrial and ribosomal genes are then filtered out of the data, as are the highly expressed genes Malat1 and Gm42418. 

```{r filtering}

if(!file.exists(paste0(owncache,"/filtering.RData"))){


alldata <- subset(alldata, features = rownames(alldata)[Matrix::rowSums(alldata) > 3], 
                  cells = intersect(WhichCells(alldata, expression = nFeature_RNA > 200),
                                    intersect(WhichCells(alldata, expression = percent_mito < 10),
                                    WhichCells(alldata, expression = percent_ribo > 5))))

gc()
save(alldata, file = paste0(owncache,"/filtering.RData"))

}else{
  rm(alldata)
  gc()
  load(paste0(owncache,"/filtering.RData"))
}

```


```{r, fig.height=8, fig.width=8}

VlnPlot(alldata, group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 1) &
    NoLegend() & violintheme

```

```{r, fig.height=4, fig.width=6}

VlnPlot(alldata,
        idents = unique(alldata$orig.ident[
          alldata$Type == "Immune"]),
        group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) + 
    NoLegend() & violintheme
VlnPlot(alldata,  
        idents = unique(alldata$orig.ident[
          alldata$Type == "Stroma"]),
        group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) + 
    NoLegend() & violintheme
VlnPlot(alldata,
        idents = unique(alldata$orig.ident[
          alldata$Type == "Epithelial"]),
        group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) + 
    NoLegend() & violintheme

```

```{r topexpressed, fig.height = 6, fig.width = 6, cache = TRUE}

par(mar = c(4, 8, 2, 1))
C <- subset(alldata, downsample = 1000, seed = 42)@assays$RNA@counts
C <- Matrix::t(Matrix::t(C)/Matrix::colSums(C)) * 100
#nC = ncol(C)

most_expressed <- order(apply(C, 1, median), decreasing = T)[20:1]
 
boxplot(as.matrix(t(C[most_expressed, ])), cex = 0.1, las = 1, xlab = "% total count per cell",
     col = (scales::hue_pal())(20)[20:1], horizontal = TRUE)


rm(C)
rm(most_expressed)



# Filter MALAT1, Gm42418, mitochondrial genes and ribosomal genes
alldata <- alldata[!grepl("Malat1", rownames(alldata)) & 
                     !grepl("Gm42418", rownames(alldata)) & 
                     !grepl("^mt-", rownames(alldata)) & ! 
                     grepl('^Rp[sl]', rownames(alldata)), ]

```



`r knitr::kable(table(alldata$orig.ident))`


```{r}

message(ls())

```


# Cell cycle scores

The S-phase and G2M-phase scores are calculated based on the cell cycle gene lists defined in the Seurat package (originally translated to mouse genes using biomaRt).

```{r cellcyclescores}

if(!file.exists(paste0(owncache,"/cellcyclescores.RData"))){

  g2mFeat = c("Smc4","Dlgap5","Ctcf","Cdc25c","Gtse1","Kif20b",
              "Ttk","Tacc3","Top2a","Tpx2","Cdca3","Ncapd2",
              "Cks2","Anp32e","G2e3","Lbr","Cdca2","Cks1brt",
              "Ckap2","Hmgb2","Kif11","Nek2","Cenpe","Hjurp",
              "Ect2","Aurkb","Cks1b","Kif23","Nuf2","Hmmr",
              "Cdca8","Psrc1","Anln","Cdc20","Birc5","Ndc80",
              "Rangap1","Ckap5","Kif2c","Cenpf","Nusap1","Cenpa",
              "Aurka","Ube2c","Ckap2l","Mki67","Tubb4b","Bub1",
              "Ccnb2")
  sFeat = c("Msh2","Exo1","Mcm4","Rrm2","Mcm2","Chaf1b","Gmnn",
            "Cdc45","Slbp","Ubr7","Cdc6",
            "Rad51ap1","Rpa2","Hells","Fen1","Gins2","Uhrf1",
            "Mcm6","Ung","Dscc1","Usp1","Clspn",
            "Cdca7","Pola1","Nasp","Dtl","Mcm5","Wdr76","Prim1",
            "Casp8ap2","Tipin","Blm","Rrm1",
            "Brip1","Rad51","Tyms","E2f8","Rfc2","Ccne2","Pcna")




# Before running CellCycleScoring the data need to be normalized and
# logtransformed.

plan("sequential")
alldata = NormalizeData(alldata)

alldata <- CellCycleScoring(object = alldata, g2m.features = g2mFeat, 
    s.features = sFeat)

rm(g2mFeat)
rm(sFeat)
gc()


if(file.exists(owncache)){
  save(alldata, file = paste0(owncache,"/cellcyclescores.RData"))
}

}else{
  rm(alldata)
  gc()
  load(paste0(owncache,"/cellcyclescores.RData"))
}


```


```{r, fig.width = 9, fig.height = 6}

VlnPlot(alldata, features = c("S.Score", "G2M.Score"), group.by = "orig.ident",
        pt.size = 0, ncol = 1) & violintheme

```


# Find Doublets

Doublets are identified and excluded using the DoubletFinder package.


```{r doublet_prep}

if(!file.exists(paste0(owncache,"/doublet_prep.RData"))){
plan("sequential")

  alldata = FindVariableFeatures(alldata, verbose = F)
  alldata = ScaleData(alldata, vars.to.regress = c("nFeature_RNA", "percent_mito"), 
      verbose = F)
  alldata = RunPCA(alldata, verbose = F, npcs = 20)
  alldata = RunUMAP(alldata, dims = 1:10, verbose = F)
  if(file.exists(owncache)){
  save(alldata, file = paste0(owncache,"/doublet_prep.RData"))
}

}else{
  rm(alldata)
  gc()
  load(paste0(owncache,"/doublet_prep.RData"))
}


```



```{r doublets}

if(!file.exists(paste0(owncache,"/doublets.RData"))){

  plan("sequential")

  sweeplist = list()

  for(i in unique(alldata$orig.ident)){
    message(i)
    data.filt=subset(alldata, idents = i)
    sweep.res <- paramSweep_v3(data.filt)#, num.cores = 3)
    sweep.stats <- summarizeSweep(sweep.res, GT = FALSE)
    bcmvn <- find.pK(sweep.stats)

    sweeplist[[i]] = bcmvn }


    singletList = list()
    for(i in unique(alldata$orig.ident)){
      message(i)
        data.filt=subset(alldata, idents = i)
        nExp <- round(ncol(data.filt)*
                        ifelse(ncol(data.filt)<12000,0.08,
                               ifelse(ncol(data.filt)<17000,0.12,
                                      ifelse(ncol(data.filt)<22000,0.16,
                                             ifelse(ncol(data.filt)<27000,0.2,
                                                    0.24)))))
        mypK = as.numeric(as.character(sweeplist[[i]]$pK[
          which.max(sweeplist[[i]]$BCmetric)]))
        data.filt <- doubletFinder_v3(data.filt, pN = 0.25, pK = mypK, nExp = nExp, PCs = 1:10)
        DF.name = colnames(data.filt@meta.data)[grepl("DF.classification", colnames(data.filt@meta.data))]
        singletList[[i]] = data.filt@meta.data[, DF.name]
        names(singletList[[i]]) = colnames(data.filt)

    rm(data.filt)
    gc()
  }

alldata = AddMetaData(alldata, unlist(singletList)[match(colnames(alldata), gsub(".*\\.","",names(unlist(singletList))))], "DF")


if(file.exists(owncache)){
  save(alldata, file = paste0(owncache,"/doublets.RData"))
}


}else{
  rm(alldata)
  gc()
  load(paste0(owncache,"/doublets.RData"))
}


```


```{r}

plan("sequential")

```



```{r, fig.height = 3, fig.width = 6}

cowplot::plot_grid(ncol = 2, DimPlot(alldata, group.by = "orig.ident") +
                     NoAxes() + NoLegend() + dimtheme, 
    DimPlot(alldata, group.by = "DF") + NoAxes() + dimtheme)


```



```{r df_dimplot, fig.height = 12, fig.width = 6}

DimPlot(alldata, group.by = "DF", split.by = "orig.ident", ncol = 6) + 
  NoAxes() & theme(plot.title = element_text(size = 8),
     strip.text = element_text(size = 8)) & dimtheme

```


```{r df_violins, fig.height = 3, fig.width = 8}

VlnPlot(alldata, features = "nFeature_RNA", group.by = "DF", 
        pt.size = 0) + violintheme
VlnPlot(alldata, features = "nFeature_RNA", split.by = "DF", 
        group.by = "orig.ident", pt.size = 0)  &
  theme(plot.title = element_text(size = 8),
        axis.text = element_text(size = 8)) + violintheme


```

`r knitr::kable(table( alldata$orig.ident, alldata$DF))`

```{r}

alldata = alldata[, alldata@meta.data[, "DF"] == "Singlet"]

```



```{r cellcounts, fig.height = 3, fig.width = 8}

cellfreq = as.data.frame(table(alldata$orig.ident), stringsAsFactors = FALSE)
cellfreq$Type = alldata$Type[match(cellfreq$Var1, alldata$orig.ident)]
ggplot(cellfreq, aes(x = Var1, y = Freq, color = Type)) + geom_point() +
  theme_classic() +
  theme(panel.background = element_blank(),  
        legend.key = element_blank(),
        panel.grid.major.y = element_line(color = "#DDDDDD"),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Sample", y = "Number of cells")

```

```{r, fig.height = 8, fig.width = 8}

VlnPlot(alldata, group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 1) &
    NoLegend() & 
  violintheme

```


```{r typeviolins, fig.height = 4, fig.width = 8}

VlnPlot(alldata[,alldata$Type == "Stroma"],
        group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) +
    NoLegend() &
  violintheme
VlnPlot(alldata[,alldata$Type == "Epithelial"],
        group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) +
    NoLegend() &
  violintheme
VlnPlot(alldata[,alldata$Type == "Immune"],
        group.by = "orig.ident", features = qc_feats, pt.size = 0, ncol = 2) +
    NoLegend() & 
  violintheme



```




```{r firstnormalize}

if(!file.exists(paste0(owncache,"/firstnormalize.RData"))){

  plan("sequential")
alldata = NormalizeData(alldata)

suppressWarnings(suppressMessages(alldata <- FindVariableFeatures(alldata, selection.method = "vst", 
    nfeatures = 2000, verbose = FALSE, assay = "RNA")))
alldata <- ScaleData(alldata, vars.to.regress = c("percent_mito", "nFeature_RNA"),     assay = "RNA")

alldata <- RunPCA(alldata, npcs = 50, verbose = F, reduction.name = "pca_raw")
alldata <- RunTSNE(alldata, reduction = "pca_raw", dims = 1:30, perplexity = 30, max_iter = 1000, 
    theta = 0.5, eta = 200, num_threads = 0, reduction.name = "tsne_raw")
alldata <- RunUMAP(alldata, reduction = "pca_raw", dims = 1:30, n.components = 2, n.neighbors = 30, 
    n.epochs = 200, min.dist = 0.3, learning.rate = 1, spread = 1, reduction.name = "umap_raw")


if(file.exists(owncache)){
  save(alldata, file = paste0(owncache,"/firstnormalize.RData"))
}

}else{
  rm(alldata)
  gc()
  load(paste0(owncache,"/firstnormalize.RData"))
}

```


```{r topgenes, fig.height = 3, fig.width = 5}

top20 <- head(VariableFeatures(alldata), 20)

lp = LabelPoints(plot = VariableFeaturePlot(alldata), points = top20, repel = TRUE) +
  theme(axis.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        axis.text = element_text(size = 8))

lp$layers[[2]]$aes_params$size = 3

lp

```


# PCA

```{r firstpca, fig.height = 4, fig.width = 7}

plot_grid(plot_grid(ncol = 3,
          DimPlot(alldata, reduction = "pca_raw", group.by = "orig.ident",
                  dims = 1:2, shuffle = TRUE) + NoLegend() +  dimtheme + 
               theme(axis.text = element_text(size = 8), axis.title = element_text(size = 8)),
          DimPlot(alldata, reduction = "pca_raw", group.by = "orig.ident",
                  dims = 3:4, shuffle = TRUE) + NoLegend() +  dimtheme + 
               theme(axis.text = element_text(size = 8), axis.title = element_text(size = 8)),
          DimPlot(alldata, reduction = "pca_raw", group.by = "orig.ident",
                  dims = 5:6, shuffle = TRUE) + NoLegend() +  dimtheme + 
               theme(axis.text = element_text(size = 8), axis.title = element_text(size = 8))),
          get_legend(DimPlot(alldata, reduction = "pca_raw", group.by = "orig.ident",
                  dims = 1:2, shuffle = TRUE) + theme(legend.position = "bottom") + dimtheme),
          ncol = 1, rel_heights = c(2,1)
)

plot_grid(plot_grid(ncol = 3,
          DimPlot(alldata, reduction = "pca_raw", group.by = "Type",
                  dims = 1:2, shuffle = TRUE) + NoLegend() +  dimtheme+
               theme(axis.text = element_text(size = 8), axis.title = element_text(size = 8)),
          DimPlot(alldata, reduction = "pca_raw", group.by = "Type",
                  dims = 3:4, shuffle = TRUE) + NoLegend() +  dimtheme+
               theme(axis.text = element_text(size = 8), axis.title = element_text(size = 8)),
          DimPlot(alldata, reduction = "pca_raw", group.by = "Type",
                  dims = 5:6, shuffle = TRUE) + NoLegend() +  dimtheme+
               theme(axis.text = element_text(size = 8), axis.title = element_text(size = 8))),
          get_legend(DimPlot(alldata, reduction = "pca_raw", group.by = "Type",
                  dims = 1:2, shuffle = TRUE) + theme(legend.position = "bottom") + dimtheme),
          ncol = 1, rel_heights = c(3,1))

```

```{r, fig.height = 4, fig.width = 7}

VizDimLoadings(alldata, dims = 1:5, reduction = "pca_raw", ncol = 5, balanced = T) &
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 8))

```

```{r, fig.height = 2, fig.width = 4}

ElbowPlot(alldata, reduction = "pca_raw", ndims = 50) + 
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 8))

```



# UMAP

```{r firstumap, fig.height = 3, fig.width = 5}

plot_grid(ncol = 2, DimPlot(alldata, reduction = "umap_raw", group.by = "orig.ident",
                            shuffle = TRUE) + NoAxes() + NoLegend() + dimtheme,
    DimPlot(alldata, reduction = "umap_raw", group.by = "Type", label = TRUE,
                            shuffle = TRUE) + NoAxes() + NoLegend() + dimtheme)

```



# Integrate data



```{r normalizeintegrated}

plan("sequential")

if(!file.exists(paste0(owncache,"/normalizeintegrated.RData"))){

if(!file.exists(paste0(owncache,"/fullintegration.RData"))){

if(!file.exists(paste0(owncache,"/getanchors.RData"))){

plan("multisession", workers = 5, gc = TRUE)

alldata <- AddMetaData(alldata, gsub("_I_","_",gsub("_S_","_",
                            gsub("_E_","_",gsub("_Str","",gsub("_Imm","",
                            gsub("_Epi","",alldata$orig.ident)))))),
                     "source")
data.list <- SplitObject(alldata, split.by = "source")
rm(alldata)
gc()

message("object split")

data.list = future_lapply(data.list, function(x){
  message("sample")
    x <- NormalizeData(x, verbose = FALSE)
    x <- FindVariableFeatures(x, selection.method = "vst",
        nfeatures = 2000, verbose = FALSE)
    x
})


message("data normalized")

features <- SelectIntegrationFeatures(object.list = data.list)

message("features selected")

#plan("multiprocess", workers = 3, gc = TRUE)
plan("sequential")

data.list <- future_lapply(X = data.list, FUN = function(x) {
  x <- ScaleData(x, features = features, verbose = FALSE)
  x <- RunPCA(x, features = features, verbose = FALSE)
})

message("pcas run")

data.anchors <- FindIntegrationAnchors(object.list = data.list, dims = 1:30,
    reduction = "rpca", anchor.features = features,
    reference = match(c("G_W0","G_W7","G_D0_ut","G_D60_ut",
                        "G_W7_aIL12","G_D30_il","G_D40_il",
                        "G_D60_il","G_W3_aIL12","G_W5_aIL12"),
                      names(data.list)))

message("anchors found")

rm(data.list)
gc()


if(file.exists(owncache)){
  save(data.anchors, file = paste0(owncache,"/getanchors.RData"))
}

}else{
  rm(alldata)
  gc()
  load(paste0(owncache,"/getanchors.RData"))
}


plan("sequential")

alldata <- IntegrateData(anchorset = data.anchors, dims = 1:30, new.assay.name = "RPCA")

rm(data.anchors)
gc()
  if(file.exists(owncache)){
  save(alldata, file = paste0(owncache,"/fullintegration.RData"))
}

}else{
  rm(alldata)
  gc()
  load(paste0(owncache,"/fullintegration.RData"))
}



alldata <- ScaleData(alldata)
alldata <- RunPCA(alldata, npcs = 30)
alldata <- RunUMAP(alldata, dims = 1:30)
  if(file.exists(owncache)){
  save(alldata, file = paste0(owncache,"/normalizeintegrated.RData"))
}

}else{
  rm(alldata)
  gc()
  load(paste0(owncache,"/normalizeintegrated.RData"))
}

```




```{r, fig.height = 3, fig.width = 6}

plot_grid(ncol = 3,

  DimPlot(alldata, reduction = "pca", group.by = "orig.ident", shuffle = TRUE) +
    NoAxes() + ggtitle("PCA integrated") + NoLegend() + dimtheme,
  DimPlot(alldata, reduction = "umap", group.by = "orig.ident", shuffle = TRUE) +
    NoAxes() + ggtitle("UMAP integrated") + NoLegend() + dimtheme,
  get_legend(DimPlot(alldata, reduction = "pca", group.by = "orig.ident", shuffle = TRUE) +
    NoAxes() + ggtitle("PCA integrated") + dimtheme)
)

  

```


```{r, fig.height = 12, fig.width = 8}

DimPlot(alldata, reduction = "umap", split.by = "orig.ident", 
        ncol = 6)+NoAxes() + NoLegend() & theme(strip.text= element_text(size = 8)) + 
  dimtheme

```


```{r, fig.height = 3, fig.width = 3.5}

DimPlot(alldata, reduction = "umap", group.by = "Model", 
        shuffle = TRUE) + NoAxes() + dimtheme


```


# Plot QC in UMAP

```{r, fig.height = 4, fig.width = 6}

FeaturePlot(alldata, reduction = "umap", 
            features = qc_feats, 
    order = T, slot = "data", combine = T) &
      NoAxes()  & dimtheme

```



# Plot genes of interest


```{r, fig.height = 4, fig.width = 6}

DefaultAssay(alldata) = "RNA"

FeaturePlot(alldata, reduction = "umap",
            features = c("Mki67","Pcna", "Ptprc","Epcam"),
    order = T, slot = "data", combine = T) & NoAxes() &
  theme(plot.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        legend.key.size = unit(8,"pt"))

DefaultAssay(alldata) = "RPCA"

```


# Clustering to identify sorting contaminations


```{r}

getseedcells = function(seeds, meta){
  big = sum(sapply(table(meta$orig.ident), function(x){return(min(x,5000))}))

fullnumbers = sort(table(meta$orig.ident))
proportions = fullnumbers/nrow(meta)
numbers = round(proportions*big)
numbers[numbers<1000] = fullnumbers[numbers<1000]
names(numbers) = names(proportions)


seeddf = data.frame(matrix(nrow=nrow(meta), ncol = length(seeds), 0))
colnames(seeddf) = paste0("seed",seeds)
rownames(seeddf) = rownames(meta)

for(thisseed in seeds){
  for(i in unique(meta$orig.ident)){
    set.seed(thisseed)
    take = numbers[i]
    cells = sample(rownames(meta)[meta$orig.ident == i], size = take)
    seeddf[cells,paste0("seed",thisseed)] = 1
  }
}

seeddf$included = apply(seeddf, 1, sum)
allcols = colnames(seeddf)[grep("seed",colnames(seeddf))]

set.seed(42)

currcol = 1
while(min(seeddf$included)<5){
  if(currcol>length(allcols)){
    currcol = 1
  }
  toofewall = rownames(seeddf)[seeddf$included<5 & seeddf[,currcol] == 0]
  toomanyall = rownames(seeddf)[seeddf$included>7 & seeddf$included<length(allcols) & seeddf[,currcol] == 1]
  least = min(length(toofewall), length(toomanyall))
  toofew = sample(toofewall, size = min(least,1000))
  toomany = sample(toomanyall, size = min(least,1000))
  seeddf[toofew,currcol] = 1
  seeddf[toomany,currcol] = 0
  seeddf[c(toofew, toomany),"included"] = apply(seeddf[c(toofew, toomany),allcols], 1, sum)
  currcol = currcol + 1
}
seeddf
}

combineclusters = function(default, clusters, meta){
  
  clusters = lapply(clusters, function(x){
    x$Cell = as.character(x$Cell)
    x[,2] = as.character(x[,2])
    x})
  
clusterdf = clusters[[1]]
  
for(i in 2:length(clusters)){
  clusterdf = merge(clusterdf, clusters[[i]], by = "Cell", all = TRUE)
}
  
meta = merge(meta, clusterdf, by.x = "row.names", by.y = "Cell")
rownames(meta) = meta[,1]
meta = meta[,-1]
  
clustercols = colnames(meta)[grep("cluster_seed",colnames(meta))]


for(i in clustercols){
  meta[,i] = factor(meta[,i],levels = c(sort(unique(meta[,i])),9999))
  meta[is.na(meta[,i]),i] = 9999
  
}

translators = list()
for(j in grep("cluster_seed", colnames(meta))){
  for(i in grep("cluster_seed", colnames(meta))){
    if(i!=j){
      ft = table(old = meta[,j], new = meta[,i])
      translators[[colnames(meta)[j]]][[colnames(meta)[i]]] = 
        data.frame(new = c(colnames(ft)[-ncol(ft)],9999), 
                   newnew = c(rownames(ft)[
                     apply(ft[-nrow(ft),-ncol(ft)],2,which.max)],9999))
      translators[[colnames(meta)[j]]][[colnames(meta)[i]]]$prop = 
        c(apply(ft[-nrow(ft),-ncol(ft)],2,function(x){max(x)/sum(x)}),NA)
    }
    
  }
}

tps = matrix(ncol = length(grep("cluster_seed", colnames(meta))), nrow = nrow(meta))
rownames(tps) = rownames(meta)
colnames(tps) = colnames(meta)[grep("cluster_seed", colnames(meta))]

for(i in grep("cluster_seed", colnames(meta))){
  iname = colnames(meta)[i]
  if(iname!=default){
    tps[,iname] = as.numeric(as.character(translators[[default]][[iname]]$newnew[match(
      as.character(meta[,i]),as.character(translators[[default]][[iname]]$new)
    )]))
  }else{
    tps[,iname] = as.numeric(as.character(meta[,i]))
  }
}

tps[tps == 9999] = NA

toptranslation = apply(tps, 1, function(x){
  if(sum(!is.na(x)>0)){names(sort(table(x), decreasing = TRUE))[1]}else{NA}})

meta$RPCA_cluster = as.numeric(unlist(toptranslation))

if(length(which(is.na(meta[,"RPCA_cluster"])))>0){stop("there are still non-assigned cells")}

meta

}

```

```{r fullclustering}

if(!file.exists(paste0(owncache,"/fullclustering.RData"))){

  if(!file.exists(paste0(owncache, "clusters.rds"))){
plan("multisession", workers = 1, gc = TRUE)

meta = alldata@meta.data
meta = meta[order(factor(meta$orig.ident,levels = sampleorder)),]

seeds = c(123,42,666,789,345,636,234,542628,57362,
          059482,7665,2342,82349,953020,03248,1284,
          4135,659,2340,6543,602)

seeddf = getseedcells(seeds = seeds, meta = meta)


res = 0.5

if(file.exists(owncache)){
  
  for(seed in seeds){
    
    if(!file.exists(paste0(owncache, "clusters/cluster_seed", seed, ".rds"))){
      message(seed)
      seedcells = rownames(seeddf)[seeddf[,match(paste0("seed",seed), colnames(seeddf))]==1]
      seeddata = subset(alldata, cells = seedcells)
      message("subsetted")
      seeddata[["RNA"]] = NULL
      
      seeddata <- FindNeighbors(seeddata, dims = 1:30, k.param = 350, prune.SNN = 1/15)
                   seeddata <- FindClusters(seeddata, graph.name = paste0("RPCA_snn"),
                                            resolution = res, algorithm = 1)
      x = seeddata@meta.data[,"seurat_clusters", drop = FALSE]
      x$Cell = rownames(x)
      colnames(x)[1] = paste0("cluster_seed",seed)
      rm(seeddata) 
      gc()
    
      saveRDS(x[c("Cell",paste0("cluster_seed",seed))], 
              paste0(owncache, "clusters/cluster_seed", seed, ".rds"))
      
      rm(x)
      gc()
    
    }
  }
  
  clusters = lapply(paste0(owncache, "clusters/cluster_seed", seeds, ".rds"), readRDS) 
  
}else{

  #clusters = lapply(gsub("seed","",colnames(seeddf)), function(seed){
  for(seed in seeds){
  #for(seed in c(57362, 123, 42)){
    message(seed)
    seedcells = rownames(seeddf)[seeddf[,match(paste0("seed",seed), colnames(seeddf))]==1]
    seeddata = subset(alldata, cells = seedcells)
    message("subsetted")
    seeddata[["RNA"]] = NULL
    
    
    futureAssign(paste0("cluster_seed",seed),
                 {seeddata <- FindNeighbors(seeddata, dims = 1:30, k.param = 350, prune.SNN = 1/15)
                 seeddata <- FindClusters(seeddata, graph.name = paste0("RPCA_snn"),
                                          resolution = res, algorithm = 1)
                 x = seeddata@meta.data[,"seurat_clusters", drop = FALSE]
                 x$Cell = rownames(x)
                 colnames(x)[1] = paste0("cluster_seed",seed)
                 x[c("Cell",paste0("cluster_seed",seed))]})              
    
    rm(seeddata)
    gc()
  
  }
  #)
  
  clusters = lapply(paste0("cluster_",seeds), get)
  #clusters = lapply(paste0("cluster_seed",c(57362, 123, 42)), get)

}

names(clusters) = paste0("cluster_", seeds)
#names(clusters) = paste0("cluster_seed",c(57362, 123, 42))

  if(file.exists(owncache)){
    saveRDS(clusters, paste0(owncache, "clusters.rds"))
  }
  rm(seeddf)
}else{
  clusters = readRDS(paste0(owncache, "clusters.rds"))
}



meta = combineclusters(default = "cluster_seed57362",
                       clusters = clusters,
                       meta = alldata@meta.data)

if(!identical(rownames(meta), colnames(alldata))){
  meta = meta[colnames(alldata),]
}

alldata = AddMetaData(alldata, meta$RPCA_cluster,"RPCA_cluster")

alldata = SetIdent(alldata, value = "RPCA_cluster")

DefaultAssay(alldata) <- "RNA"

rm(meta, clusters)




  if(file.exists(owncache)){
  save(alldata, file = paste0(owncache,"/fullclustering.RData"))
}

}else{
  rm(alldata)
  gc()
  load(paste0(owncache,"/fullclustering.RData"))
}

```

```{r}

plan("sequential")

```

```{r, fig.height = 3, fig.width = 6}

plot_grid(DimPlot(alldata, label = TRUE) + NoAxes() + NoLegend() + dimtheme,
  DimPlot(alldata, group.by = "Type", label = TRUE) + NoAxes() + NoLegend() + 
    labs(title = "Sort fraction") + dimtheme, ncol = 2)

```

```{r, fig.height = 3, fig.width = 6}

DotPlot(alldata, features = c("Epcam","Ptprc")) + coord_flip() +
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 8),
	legend.text = element_text(size = 8),
	legend.title = element_text(size = 8))

```


```{r, fig.height = 3, fig.width = 5}

ggplot(alldata@meta.data, aes(fill = Type, x = RPCA_cluster)) + geom_bar(position = "fill")

ggplot(alldata@meta.data, aes(fill = Model, x = RPCA_cluster)) + geom_bar(position = "fill")


```




```{r, fig.height = 3, fig.width = 6}

VlnPlot(alldata, group.by = "RPCA_cluster", features = qc_feats, pt.size = 0, ncol = 2) &
  theme(plot.title = element_text(size = 8),
        axis.text = element_text(size = 8),
        axis.title = element_blank()) + violintheme

```


# Contaminant removal

```{r cleandata}


if(!file.exists(paste0(owncache, "cleaned.RData"))){

  immuneclusters = c(0,1,3,8,10,12,14,16,21)
  stromaclusters = c(4,7,9,11,13,15,17,18,20)
  epiclusters = c(2,5,6,19)
  
  sel.clust = "RPCA_cluster"
  
  alldata <- SetIdent(alldata, value = "Type")
  
  immunetype = WhichCells(alldata, ident = "Immune")
  stromatype = WhichCells(alldata, ident = "Stroma")
  epitype = WhichCells(alldata, ident = "Epithelial")
  
  alldata <- SetIdent(alldata, value = sel.clust)
  alldata$ident_clust = paste0(alldata@active.ident,"_",alldata$orig.ident)
  
  immunecells = WhichCells(alldata, ident = immuneclusters, cells = immunetype)
  stromacells = WhichCells(alldata, ident = stromaclusters, cells = stromatype)
  epicells = WhichCells(alldata, ident = epiclusters, cells = epitype)
  
  alldata = subset(alldata, cells = c(immunecells,stromacells, epicells))
  
  cleandata = alldata
  
  rm(alldata)
  
  gc()
  
  immunedata = subset(cleandata, cells = immunecells)
  immunedata@active.assay = "RNA"
  immunedata[["RPCA"]] = NULL
  immunedata[["FB"]] = subset(alldata_FB, cells = Cells(immunedata))[["FB"]]
  saveRDS(immunedata, "../data/SingleCell/TcellTransfer/R_objects/immune_data.rds")
  rm(immunedata)
  
  stromadata = subset(cleandata, cells = stromacells)
  stromadata@active.assay = "RNA"
  stromadata[["RPCA"]] = NULL
  saveRDS(stromadata, "../data/SingleCell/TcellTransfer/R_objects/stroma_data.rds")
  rm(stromadata)
  
  epidata = subset(cleandata, cells = epicells)
  epidata@active.assay = "RNA"
  epidata[["RPCA"]] = NULL
  saveRDS(epidata, "../data/SingleCell/TcellTransfer/R_objects/epithelial_data.rds")
  rm(epidata)

  gc()

  plan("sequential")

  cleandata@active.assay = "RPCA"
  
  cleandata <- ScaleData(cleandata)
  cleandata <- RunPCA(cleandata, npcs = 30)
  cleandata <- RunUMAP(cleandata, dims = 1:30)
  
  save(cleandata, file = paste0(owncache, "cleaned.RData"))  

}else{
  rm(alldata)
  gc()
  load(paste0(owncache, "cleaned.RData"))
}

```



```{r dimplots,fig.height = 3, fig.width = 3}

DimPlot(cleandata, reduction = "umap", label = TRUE,
                        group.by = "RPCA_cluster", shuffle = TRUE) +
            NoAxes() + labs(title = "Clustering") + NoLegend() + 
  theme(plot.title = element_text(size = 10)) + dimtheme

```


```{r, fig.height = 3, fig.width=7}

DimPlot(cleandata, reduction = "umap", split.by = "Type", label = TRUE) &
  NoAxes() & NoLegend() & dimtheme

```


```{r umaps, fig.height = 6, fig.width = 6}

typecolors = c(Stroma = "#F4DF9E",Immune = "#D6A7A0", Epithelial = "#A6D8DA")
modelcolors = c(Rag2 = "#9E68B1", SCID = "#FCA549")
treatmentcolors = c(il = "#00C9CF", ut = "#FB8C80", aIL12 = "#00C9CF", untreated = "#FB8C80")
timecolors = scale_color_viridis_c(option = "viridis", direction = 1)


allplots = list(
          DimPlot(cleandata, reduction = "umap", group.by = "Model", shuffle = TRUE) + 
            NoAxes() + 
             scale_color_manual(values = modelcolors) + dimtheme,
          FeaturePlot(cleandata, cells = sample(Cells(cleandata)),
                      reduction = "umap", features = "RelTime") + 
             NoAxes() + labs(title = "Time point") + timecolors + dimtheme,
          DimPlot(cleandata, reduction = "umap", group.by = "Treatment", shuffle = TRUE) +
            NoAxes() + dimtheme +
             scale_color_manual(values = treatmentcolors),
          DimPlot(cleandata, reduction = "umap", group.by = "Type", shuffle = TRUE) +
            NoAxes() + labs(title = "Sort fraction") + dimtheme +
             scale_color_manual(values = typecolors))

plot_grid(ncol = 2, plotlist = allplots)


```


# Session info

```{r}

sessionInfo()

```
